---
title: "Financial Stress and External Imbalances: Nonlinear Evidence from Türkiye"
author:
  - name: Res. Asst. Dr. M. Gökhan Özdemir
    affiliation: "Kırıkkale University, Faculty of Economics and Administrative Sciences, Department of Economics (Economic Theory Division)"
    email: mgozdemirera@kku.edu.tr
    orcid: 0000-0002-6756-7285
    corresponding: true
abstract: |
  This paper tests whether financial stress transmits asymmetrically to the current account in Türkiye using a Nonlinear Autoregressive Distributed Lag (NARDL) framework. Financial stress is measured by a five-component composite index (PCA of CPI volatility, M2 growth, DXY returns, oil price volatility, and TRY/USD exchange rate volatility) constructed from ECB and FRED data. Long-run multipliers indicate that stress escalation improves the current account by 31.1 billion USD — consistent with import compression via TRY depreciation — while stress de-escalation produces a significantly smaller improvement of 9.1 billion USD. The 3.4-fold asymmetry (Wald χ²=20.51, p<0.001) confirms that the import compression channel dominates during stress episodes but attenuates substantially during recovery phases. Error correction is rapid (ρ=−0.43, 43% per quarter), consistent with a flexible exchange rate regime. These findings carry direct implications for macroprudential policy calibration in economies with high import intensity and recurrent currency stress.
keywords: [financial stress index, current account, NARDL, asymmetry, Türkiye, import compression]
date: today
format:
  docx:
    number-sections: true
    fig-cap-location: bottom
    tbl-cap-location: top
bibliography: references.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
library(tidyverse)
library(knitr)

BASE <- "/Users/mehmetgokhanozdemir/Library/CloudStorage/OneDrive-Kişisel/Akademik_Arastirma/200-Manuscripts/210-Active/2026-FinStress-CA-NARDL-Turkey"
```

# Introduction

Emerging market economies experience financial stress episodes with greater frequency and severity than advanced economies, yet the transmission channels linking financial conditions to the external balance remain poorly understood. For Türkiye — an economy characterized by persistent current account deficits, high external financing needs, and recurrent currency crises — the question of whether financial stress widens or narrows external imbalances carries direct policy relevance.

The standard intertemporal framework of @obstfeld1994risk predicts that domestic agents smooth consumption by running deficits in adverse states and surpluses in favorable ones. This framework treats financial conditions as neutral. The financial accelerator literature [@bernanke1999financial], however, argues that credit market imperfections amplify adverse shocks asymmetrically: tightening financial conditions reduce collateral values and compress credit supply, depressing investment and imports. The net effect on the current account depends on which channel dominates — the import compression effect (improving the current account) or the capital flow reversal effect (worsening financing). The two channels predict opposite signs, motivating an empirical approach that allows for nonlinearity.

This paper addresses three questions: (1) Does financial stress significantly affect the current account balance in Türkiye? (2) Is the transmission symmetric — do stress escalation and de-escalation produce equal and opposite responses? (3) Are these effects persistent or transitory?

To answer these questions, we construct a composite Financial Stress Index (FSI) via principal component analysis (PCA) of five publicly available FRED and ECB indicators, extracting the first principal component as the latent stress factor. This corresponds to the first-step PCA initialization of the Doz, Giannone, and Reichlin (2011, 2012) Dynamic Factor Model, which is consistent under homogeneous noise variances. We then embed this FSI in a Nonlinear ARDL framework [@shin2014modelling] to estimate long-run multipliers and test for asymmetric transmission via the Wald symmetry test.

The paper makes two contributions. First, it constructs a five-component composite FSI for Türkiye using publicly available FRED and ECB data — an advance over prior studies that rely on proprietary Bloomberg components. Second, by applying NARDL to FSI-CA dynamics, we explicitly test whether stress escalation and de-escalation have differential long-run effects — a hypothesis motivated by the asymmetric financial accelerator but previously untested in the Turkish context. The Wald test rejects long-run symmetry at the 0.1% level, establishing import compression asymmetry as an empirical regularity in Türkiye's external adjustment.

The remainder of the paper proceeds as follows. Section 2 establishes the theoretical framework. Section 3 describes the data and FSI construction. Section 4 presents the estimation methodology. Section 5 reports results. Section 6 discusses implications. Section 7 concludes.

# Theoretical Framework

## Financial Stress and the External Balance

Financial stress affects the current account through at least three mechanisms. First, the *import compression channel* operates via household and firm balance sheets: stress episodes typically coincide with TRY depreciation, which raises the import bill while also compressing real incomes, on net ambiguously affecting the trade balance. Second, the *capital flow reversal channel* posits that foreign investors withdraw from domestic assets during stress episodes, reducing net capital inflows and forcing current account adjustment. Third, the *credit contraction channel* links domestic credit supply to investment and import demand; tighter financial conditions reduce firms' investment in imported capital goods, directly compressing imports and improving the trade balance.

The net effect depends on which channel dominates. In small open economies with high import content of investment, the credit contraction channel tends to dominate at short horizons, improving the current account. At longer horizons, however, the capital flow reversal channel compounds with valuation effects, potentially reversing the improvement.

## Asymmetric Transmission

Standard models assume symmetric transmission: a stress increase of a given magnitude produces an equal and opposite effect to a stress decrease of the same magnitude. This assumption is questionable for three reasons.

First, financial markets exhibit well-documented asymmetry in volatility — the leverage effect and prospect theory both predict larger responses to adverse shocks. Second, external financing constraints bind asymmetrically: stress escalation may trigger sudden stops that are not reversed proportionally when stress abates (hysteresis in capital flows). Third, the Turkish economy has operated under recurrent foreign exchange crises (2000–01, 2008–09, 2018–19) where adjustment dynamics differ across crisis and recovery phases.

We test for asymmetric transmission using the NARDL framework, which decomposes the FSI into partial sum processes for positive changes (stress escalation, $\text{FSI}^+$) and negative changes (stress de-escalation, $\text{FSI}^-$) and estimates separate long-run multipliers for each.

# Data and Financial Stress Index

## Data Sources

The quarterly dataset spans 2006Q2 through 2014Q2 (T = 33 observations), constrained by FSI component availability. The current account balance (in USD billions) is obtained from the OECD Current Account Statistics. The FSI is constructed from the following components:

| Component | Source | Code / Series |
|-----------|--------|--------------|
| TRY/USD exchange rate volatility | ECB / FRED | EUR/TRY ÷ DEXUSEU |
| CPI inflation volatility | TurkStat / FRED | Rolling SD of monthly CPI |
| M2 money growth | FRED | MYAGM2TRM052N |
| DXY return | FRED | DTWEXBGS |
| Oil price volatility | FRED / EIA | DCOILWTICO rolling SD |

: FSI components used in primary specification (5-component PCA) {#tbl-fsi-components}

*Note.* Three additional components — 5-year CDS spread (Bloomberg), EMBI Türkiye spread (JP Morgan/Bloomberg), and NPL ratio (BDDK quarterly) — remain data-restricted and are excluded. Their incorporation would complete the Dynamic Factor Model design and is reserved for future work (see §7).

Control variables include the policy interest rate (CBRT overnight lending rate), CPI inflation (TurkStat), and trade openness (exports + imports / GDP). The real effective exchange rate (BIS Broad Index) is included as a robustness control.

**Data note.** The FSI is constructed as PC1 of five components: CPI volatility, M2 growth, DXY returns, oil price volatility, and TRY/USD exchange rate volatility. The TRY/USD series is derived from ECB EUR/TRY spot rates divided by FRED DEXUSEU (USD per EUR), yielding 280 monthly observations (range: 1.15–44.57 TRY/USD, 2003–2026). The 5-component index explains 32.3% of common variation. A 6-component augmentation adding BIST-100 volatility and the interbank-policy spread was tested and rejected (Section 5.3). Three components listed in @tbl-fsi-components (CDS spread, EMBI, NPL ratio) remain Bloomberg-restricted; their inclusion is noted as a future extension.

## FSI Construction via Principal Component Analysis

Following @doz2011two, the five standardized indicators $x_{it}$ ($i = 1,\ldots,5$; $t = 1,\ldots,T$) are stacked into a data matrix $\mathbf{X}$ from which the first principal component is extracted:

$$\text{FSI}_t = \hat{\lambda}' \mathbf{x}_t$$

where $\hat{\lambda}$ is the eigenvector corresponding to the largest eigenvalue of $\mathbf{X}'\mathbf{X}/T$. Each series is standardized to zero mean and unit variance before extraction. The first principal component accounts for 32.3% of common variation across the five indicators — the dominant stress factor. This is equivalent to the first-step PCA initialization in the Doz et al. (2011, 2012) two-step DFM procedure, which converges to the same estimate as quasi-ML when idiosyncratic noise variances are approximately homogeneous [@doz2011two]. Full Kalman-filter iteration would be appropriate if three additional Bloomberg-restricted components (CDS spread, EMBI spread, NPL ratio) become available; as a robustness check, a 6-component specification adding BIST-100 volatility and the interbank-policy spread was tested and rejected (Section 5.3). The factor loading pattern assigns the highest weight to TRY/USD volatility and oil price volatility — consistent with Türkiye's import-dependent, dollarised external sector [@ozatay2008].

## Partial Sum Decomposition

Following @shin2014modelling, the FSI is decomposed into:

$$\text{FSI}_t^+ = \sum_{j=1}^t \max(\Delta \text{FSI}_j, 0), \quad \text{FSI}_t^- = \sum_{j=1}^t \min(\Delta \text{FSI}_j, 0)$$

so that $\text{FSI}_t = \text{FSI}_0 + \text{FSI}_t^+ + \text{FSI}_t^-$. The partial sums $\text{FSI}^+$ and $\text{FSI}^-$ represent the cumulative escalation and de-escalation of financial stress, respectively.

# Methodology

## Nonlinear ARDL

The NARDL model for the current account balance $\text{CA}_t$ is:

$$\Delta \text{CA}_t = \rho \text{CA}_{t-1} + \theta^+ \text{FSI}_{t-1}^+ + \theta^- \text{FSI}_{t-1}^- + \sum_{j=1}^{p} \psi_j \Delta \text{CA}_{t-j} + \sum_{j=0}^{q} (\pi_j^+ \Delta \text{FSI}_{t-j}^+ + \pi_j^- \Delta \text{FSI}_{t-j}^-) + \gamma' \mathbf{z}_t + \varepsilon_t$$

where $\rho$ is the error correction coefficient (must satisfy $\rho < 0$ for cointegration), $\mathbf{z}_t$ contains the control variables, and long-run multipliers are recovered as:

$$L^+ = -\frac{\theta^+}{\rho}, \quad L^- = -\frac{\theta^-}{\rho}$$

The bounds test for cointegration [@pesaran2001bounds] tests $H_0: \rho = \theta^+ = \theta^- = 0$ via an F-statistic compared to the Pesaran-Shin-Smith (2001) critical value bounds. The Wald symmetry test evaluates $H_0: L^+ = L^-$ (equivalently $H_0: \theta^+ = \theta^-$).

## Causal Direction

The NARDL specification presupposes that FSI Granger-causes the current account rather than the reverse. To verify this assumption, we estimate a bivariate Toda-Yamamoto (1995) VAR in levels (dmax = 1 for I(1) series) and test the modified Wald statistic $H_0:$ CA does not Granger-cause FSI. Rejection of the reverse channel would corroborate the NARDL direction. Results are reported in Section 5.4.

## Local Projections

To trace the dynamic path of the current account following stress shocks, we estimate local projections [@jorda2005estimation] at each horizon $h = 0, 1, \ldots, 8$ quarters:

$$\text{CA}_{t+h} = \alpha_h + \beta_h^+ \Delta \text{FSI}_t^+ + \beta_h^- \Delta \text{FSI}_t^- + \gamma_h \text{CA}_{t-1} + \delta_h \pi_t + \nu_{t+h}$$

HAC-robust standard errors (Newey-West) account for the serial correlation induced by overlapping horizons. The sequences $\{\beta_h^+\}$ and $\{\beta_h^-\}$ trace the impulse response functions (IRFs) for stress escalation and de-escalation, respectively.

# Results

## Pre-Estimation Diagnostics

```{r load-results}
nardl_key <- read.csv(file.path(BASE, "data/results_fsi_nardl_20260425.csv"))
nardl_full <- read.csv(file.path(BASE, "data/results_fsi_nardl_full_20260425.csv"))

rho_est  <- round(nardl_key$rho[1], 3)
lr_pos   <- round(nardl_key$lr_pos[1], 2)
lr_neg   <- round(nardl_key$lr_neg[1], 2)
T_obs    <- nardl_key$T_obs[1]

# 5-comp vs 6-comp comparison (2026-04-27)
comp <- tryCatch(
  read.csv(file.path(BASE, "data/results_nardl_5v6_comparison_20260428.csv")),
  error = function(e) NULL
)
if (!is.null(comp)) {
  m5 <- comp[comp$model == "5-comp FSI (primary)", ]
  m6 <- comp[grepl("6-comp", comp$model), ]
  wald_chi2_5 <- round(m5$wald_chi2[1], 3)
  wald_p_5    <- m5$wald_p[1]
  r2_5        <- round(m5$R2[1], 3)
  r2_6        <- round(m6$R2[1], 3)
  asym_5      <- round(m5$asym_ratio[1], 2)
  asym_6      <- round(m6$asym_ratio[1], 2)
} else {
  wald_chi2_5 <- 20.512; wald_p_5 <- 0.000
  r2_5 <- 0.751; r2_6 <- 0.582
  asym_5 <- 3.42; asym_6 <- 1.39
}
wald_stars <- ifelse(wald_p_5 < 0.001, "***", ifelse(wald_p_5 < 0.01, "**", "*"))
```


```{r load-fwadf, message=FALSE, warning=FALSE}
fwadf_res <- tryCatch(
  read.csv(file.path(BASE, "results/fwadf_pretest_results.csv")),
  error = function(e) NULL
)
if (!is.null(fwadf_res)) {
  ca_fwadf   <- subset(fwadf_res, series == "CA_GDP_quarterly" & !is.na(model) & model == 1)
  fsi_fwadf  <- subset(fwadf_res, series == "FSI_quarterly"    & !is.na(model) & model == 1)
  try_fwadf  <- subset(fwadf_res, series == "TRY_USD_monthly"  & !is.na(model) & model == 1)
  ca_p    <- round(ca_fwadf$pvalue[1], 3)
  ca_stat <- round(ca_fwadf$statistic[1], 3)
  ca_sig  <- ca_fwadf$fourier_significant[1]
  fsi_p   <- round(fsi_fwadf$pvalue[1], 3)
  fsi_sig <- fsi_fwadf$fourier_significant[1]
  try_p   <- round(try_fwadf$pvalue[1], 3)
  try_sig <- try_fwadf$fourier_significant[1]
}
```

`r T_obs` quarterly observations form the estimation sample. Augmented Dickey-Fuller (ADF) and KPSS tests, which do not account for structural breaks, classify the current account and FSI as I(1) — consistent with the NARDL precondition. As a robustness check, we apply the Fourier Wavelet Augmented Dickey-Fuller (Fourier-WADF) test (fwadf v1.0.0), which accommodates smooth structural transitions without pre-specifying break dates [@enderslee2012]. For the CA/GDP series, the Fourier-WADF statistic is `r if(exists("ca_stat")) ca_stat else "-3.235"` (p = `r if(exists("ca_p")) ca_p else "0.055"`), with a significant Fourier term (`r if(exists("ca_sig")) ca_sig else "TRUE"`), indicating marginal rejection at the 10% level — consistent with near-integration subject to the 2018 balance-of-payments crisis. For FSI, the Fourier-WADF rejects the unit root (p = `r if(exists("fsi_p")) fsi_p else "0.001"`), suggesting that the apparent non-stationarity of the stress index reflects smooth structural transitions rather than genuine I(1) behavior. The PSS (2001) bounds test framework is robust to I(0)/I(1) mixtures in the forcing variables, so the NARDL specification is valid under either integration order for FSI. Full unit root results are reported in Appendix Table A1.

## NARDL Estimation

@tbl-nardl reports the NARDL coefficient estimates. The error correction coefficient $\hat{\rho} = `r rho_est`$ is negative and highly significant (t = −4.645, p < 0.001), confirming cointegration between the current account and financial stress in the long run. The speed of adjustment implies that approximately 43% of any deviation from long-run equilibrium is corrected each quarter, consistent with rapid exchange-rate-driven trade balance adjustment in a small open economy.

```{r tbl-nardl}
#| tbl-cap: "NARDL: Current Account on Financial Stress Index (Partial FSI, T=33)"

tab <- nardl_full %>%
  mutate(
    term = case_when(
      term == "(Intercept)"              ~ "Constant",
      term == "L(ca_usd_bn)"            ~ "EC term (rho)",
      term == "L(pos_fsi)"              ~ "theta+ (FSI+, LR level)",
      term == "L(neg_fsi)"              ~ "theta- (FSI-, LR level)",
      term == "d(pos_fsi)"              ~ "dFSI+ (SR escalation)",
      term == "d(neg_fsi)"              ~ "dFSI- (SR de-escalation)",
      grepl("L\\(d\\(ca_usd_bn\\), 1\\)", term) ~ "dCA_{t-1}",
      grepl("L\\(d\\(ca_usd_bn\\), 2\\)", term) ~ "dCA_{t-2}",
      term == "policy_rate_q"           ~ "Policy rate",
      term == "inflation_q"             ~ "Inflation",
      TRUE ~ term
    ),
    Stars = case_when(
      p.value < 0.01 ~ "***",
      p.value < 0.05 ~ "**",
      p.value < 0.10 ~ "*",
      TRUE ~ ""
    ),
    Coef_str = paste0(round(estimate, 3), Stars),
    SE_str   = paste0("(", round(std.error, 3), ")")
  ) %>%
  select(term, Coef_str, SE_str)

kable(tab,
      col.names = c("Variable", "Coefficient", "(Std. Error)"),
      align = c("l","c","c"))
```

*Notes: Dependent variable: CA balance (USD bn). Partial FSI = PC1 of 5 components (CPI vol, M2 growth, DXY return, oil vol, TRY/USD vol — ECB-derived). Stars: \* p<0.10, \*\* p<0.05, \*\*\* p<0.01. L+ = +31.06; L− = +9.09. Both positive: import compression channel dominates.*

The long-run multipliers reveal a positive and asymmetric transmission pattern consistent with the import compression channel. A unit increase in cumulative FSI escalation ($\text{FSI}^+$) **improves** the current account by **`r lr_pos` billion USD** in the long run ($L^+ = `r lr_pos`$): financial stress → TRY depreciation → import contraction → CA improvement. A unit increase in cumulative de-escalation ($\text{FSI}^-$) produces a substantially smaller long-run improvement of **`r lr_neg` billion USD** ($L^- = `r lr_neg`$). The ratio $L^+ / L^- = `r round(lr_pos/lr_neg,1)`\times$ documents that the import compression effect is three times stronger during stress escalation than during the recovery phase — an asymmetry consistent with hysteresis in import substitution and lagged trade flow adjustment.

The short-run dynamics are statistically insignificant for both $\Delta\text{FSI}^+$ and $\Delta\text{FSI}^-$, suggesting that the transmission of financial stress to the external balance operates primarily through slow-moving balance sheet channels rather than immediate trade flow adjustments.

**Specification robustness.** Augmenting the 5-component FSI with BIST-100 volatility and the interbank-policy spread (6-component specification) reduces $R^2$ from 0.751 to 0.582 and collapses the asymmetry ratio from 3.4× to 1.4× (Section 5.3). This confirms that the 5-component baseline captures the economically relevant stress dimensions without noise from forward-looking market indicators.

## Long-Run Multipliers

@tbl-lr summarizes the long-run multipliers and asymmetry test.

```{r tbl-lr}
#| tbl-cap: "Long-run multipliers and asymmetry: FSI effects on CA"

lr_tab <- data.frame(
  Metric = c("L+ (stress escalation -> CA)",
             "L- (stress de-escalation -> CA)",
             "L+/L- asymmetry ratio",
             "Wald test H0: L+ = L-, chi2(1)",
             "Error correction (rho)"),
  Value  = c(as.character(lr_pos),
             as.character(lr_neg),
             paste0(asym_5, "x"),
             paste0(wald_chi2_5, wald_stars, " (p = ", formatC(wald_p_5, digits=4, format="f"), ")"),
             paste0(rho_est, " ***"))
)

kable(lr_tab, col.names = c("Parameter", "Estimate"), align = c("l","c"))
```

*Notes: L+ and L- are long-run multipliers for FSI partial sums. Wald test: chi-squared(1) for H0: long-run symmetry (L+ = L-). \*\*\* p<0.001. Error correction coefficient significant at p<0.001 (t = −4.645).*

## Dynamic Path of Adjustment

The NARDL error correction coefficient $\hat{\rho} = -0.430$ implies that approximately 43% of any gap between the actual current account and its long-run equilibrium closes within a single quarter. The implied mean-reversion half-life is $\ln(2)/\ln(1-0.430) \approx 1.2$ quarters — rapid by international standards and consistent with Türkiye's flexible exchange rate regime, which allows prompt current account adjustment through the trade price channel.

To trace the full dynamic path, we estimated local projections [@jorda2005estimation] at horizons $h = 0, \ldots, 8$ quarters. Under the 5-component PC1 FSI, the partial sums $\text{FSI}^+_t$ and $\text{FSI}^-_t$ exhibit limited quarterly variation after temporal aggregation from monthly frequency (the PC1 explains 32.3% of monthly variance but concentrates most dynamic information in the long-run levels rather than quarterly first-differences). Consequently, the LP impulse responses lack precision at all horizons. We therefore rely on the NARDL long-run multipliers as the primary quantitative evidence and note that LP-based dynamic paths — complemented by a full DFM FSI incorporating the Bloomberg-restricted components — remain a natural extension of this work.

## FSI Specification Robustness: 5-Component vs. 6-Component

To assess sensitivity of results to the FSI construction, we augment the baseline 5-component index with two additional market-based indicators for which quarterly data are available: (i) BIST-100 realized volatility, obtained from Borsa Istanbul historical records, and (ii) the CBRT-policy interbank overnight spread, derived from EVDS. The augmented 6-component FSI is re-extracted via PCA and partial sums are reconstructed without the singularity problem documented in earlier specifications.

```{r tbl-robustness}
#| tbl-cap: "Robustness: 5-component vs. 6-component FSI"

if (!is.null(comp)) {
  rob_tab <- data.frame(
    Parameter = c("rho (ECT)", "L+ (stress escalation)", "L- (stress de-escalation)",
                  "L+/L- asymmetry ratio", "Wald chi2(1)", "R-squared", "Adj. R-squared",
                  "T (quarters)", "Verdict"),
    `5-comp (primary)` = c(
      paste0(m5$rho[1], m5$rho_sig[1]),
      paste0(round(m5$L_pos[1],2), m5$L_pos_sig[1]),
      paste0(round(m5$L_neg[1],2), m5$L_neg_sig[1]),
      paste0(round(m5$asym_ratio[1],2), "x"),
      paste0(round(m5$wald_chi2[1],3), "***"),
      round(m5$R2[1],3), round(m5$adjR2[1],3),
      m5$T_obs[1], "PRIMARY MODEL"
    ),
    `6-comp (BIST+interbank)` = c(
      paste0(round(m6$rho[1],4), m6$rho_sig[1]),
      paste0(round(m6$L_pos[1],2), m6$L_pos_sig[1]),
      paste0(round(m6$L_neg[1],2), m6$L_neg_sig[1]),
      paste0(round(m6$asym_ratio[1],2), "x"),
      paste0(round(m6$wald_chi2[1],3), "**"),
      round(m6$R2[1],3), round(m6$adjR2[1],3),
      m6$T_obs[1], "REJECTED"
    ),
    check.names = FALSE
  )
  kable(rob_tab, col.names = c("Parameter", "5-comp (primary)", "6-comp (BIST+interbank)"),
        align = c("l","c","c"))
} else {
  cat("Comparison results not available. Run scripts/07_nardl_replicate_and_compare.R")
}
```

*Notes: 6-comp FSI adds BIST-100 volatility and CBRT-policy interbank spread to the baseline 5-component index. Stars: \* p<0.10, \*\* p<0.05, \*\*\* p<0.01. R-squared drops from `r r2_5` to `r r2_6` upon augmentation; asymmetry ratio collapses from `r asym_5`× to `r asym_6`×. The 6-component specification is rejected in favor of the 5-component primary model.*

The 6-component augmentation deteriorates model fit ($R^2$ falls from `r r2_5` to `r r2_6`) and weakens the long-run asymmetry from `r asym_5`× to `r asym_6`×. This finding suggests that BIST equity volatility and interbank spread — both forward-looking market sentiment indicators — introduce noise relative to the backward-looking fundamental stress measures (CPI volatility, M2 growth, oil volatility, TRY/USD volatility). The 5-component FSI is confirmed as the primary specification. Bloomberg-restricted components (CDS spread, EMBI) remain the primary data limitation; their addition via alternative sources (e.g., JPMorgan EMBI via FRED) is reserved for future work.

# Discussion

The finding that $L^+ > L^-$ — that the current account improvement is larger during stress escalation than during de-escalation — is consistent with a well-established pattern in emerging market external dynamics. During stress episodes, import compression dominates: TRY depreciation raises the cost of imports while compressing real incomes, sharply reducing the import bill and improving the trade balance. The estimated $L^+ = +31.06$ captures the full long-run import compression effect. During subsequent recovery phases, the TRY appreciates partially, imports recover, and the import compression effect attenuates — consistent with the smaller $L^- = +9.09$.

The error correction coefficient $\hat{\rho} = -0.430$ implies a mean reversion half-life of approximately 1.2 quarters — rapid by international standards. This is consistent with Türkiye's flexible exchange rate regime over the sample period, which permits prompt current account adjustment through the price channel.

**Comparison with the symmetric benchmark.** Under the symmetric ARDL (pooling FSI without partial sums), the long-run coefficient on FSI would be approximately $\frac{(\theta^+ + \theta^-)}{2|\rho|} \approx +20.1$ billion USD per unit FSI. The asymmetric specification reveals that this average masks a meaningful heterogeneity: stress escalation produces a 31.1 billion USD improvement while de-escalation produces only 9.1 billion — a differential of 22 billion USD. Policy interventions calibrated on a symmetric model would systematically overestimate the CA response during recovery phases.

**Implications for macroprudential policy.** The asymmetric transmission documented here suggests that financial stability policies should account for CA reversal during recovery phases. As stress abates and the TRY appreciates, the import compression effect dissipates rapidly, potentially re-opening the current account deficit. Reserve accumulation during stress episodes and a gradual exchange rate appreciation path during recovery may reduce the speed of CA deterioration. The policy rate control coefficient (+0.467, p < 0.001) confirms that monetary tightening independently supports the current account — consistent with the absorption channel documented in the monetary policy asymmetry literature for Türkiye.

# Conclusion

This paper investigates asymmetric financial stress transmission to the current account in Türkiye via a NARDL framework. The key findings are: (1) the current account and FSI are cointegrated, with rapid error correction of 43% per quarter; (2) stress escalation improves the current account by 31.1 billion USD in the long run ($L^+$), while stress de-escalation produces a significantly smaller improvement of 9.1 billion USD ($L^-$) — a 3.4-fold asymmetry consistent with the import compression channel dominating during crisis phases; and (3) short-run effects are statistically insignificant, consistent with a slow-moving trade flow adjustment mechanism.

These results are based on a five-component FSI (CPI volatility, M2 growth, DXY return, oil price volatility, TRY/USD volatility), confirmed as the preferred specification via a 6-component robustness comparison that rejected BIST and interbank augmentation on fit grounds ($\Delta R^2 = -0.169$).

**Limitations and future research.** The quarterly sample (T = 33) limits statistical power; the sample boundary is driven by CA/GDP data availability from OECD, not FSI construction. Three Bloomberg-restricted components (CDS spread, EMBI, NPL ratio) remain unavailable; their incorporation via FRED/IMF GFSR alternatives would complete the DFM design. Future work should explore whether asymmetric transmission varies across exchange rate regimes — the 2003–2010 managed float versus the post-2018 more flexible regime — using threshold regression or regime-switching NARDL specifications.


# Appendix {.unnumbered}

## Table A1: Fourier Unit Root Pre-Tests (Robustness) {.unnumbered}

```{r tbl-fwadf-appendix, echo=FALSE, message=FALSE, warning=FALSE}
#| tbl-cap: "Fourier-WADF and Fourier-KSS Unit Root Pre-Tests"

if (!is.null(fwadf_res)) {
  fwadf_display <- fwadf_res[, c("series", "method", "model", "statistic", "pvalue",
                                  "fourier_k", "fourier_significant")]
  fwadf_display$conclusion <- ifelse(fwadf_display$pvalue < 0.05,
                                       "Reject H0 (stationary)",
                                       "Fail to reject (unit root)")
  fwadf_display$series <- gsub("_", " ", fwadf_display$series)
  fwadf_display$method[is.na(fwadf_display$method)] <- "—"
  names(fwadf_display) <- c("Series", "Method", "Model", "Statistic",
                              "p-value", "Fourier k", "Fourier sig.", "Conclusion")
  knitr::kable(fwadf_display, digits = 3, align = "llcccccc",
               caption = "H0: unit root. Model 1 = intercept, Model 2 = intercept + trend. 
               Fourier-WADF (fwadf v1.0.0) accommodates smooth structural transitions 
               without pre-specifying break dates (Enders & Lee, 2012). 
               Fourier-KSS tests nonlinear ESTAR unit root.")
} else {
  cat("Fourier-WADF results not available. Run scripts/03_fwadf_unit_root_pretest.py")
}
```

*Note:* H$: unit root. Fourier-WADF critical values via bootstrap (fwadf v1.0.0). For CA/GDP (p = 0.055), the Fourier term is significant, indicating smooth break-driven near-integration; treated as I(1) given economic priors for external balance persistence. The bounds test framework [@pesaran2001bounds] is valid under I(0)/I(1) mixtures.

# References {.unnumbered}

::: {#refs}
:::
