---
title: "Determinants of Sub-Saharan Africa's Agricultural Value Added and 2030 Projection: Multilayer Artificial Neural Networks and Scenario Analysis Across 29 Countries"
format: 
  docx:
    # reference-doc: null
    toc: false
    number-sections: true
    fig-width: 7
    fig-height: 5
  html: default
execute:
  echo: false
  warning: false
  eval: true 
---

```{r}
#| label: setup-and-calc
#| include: false
#| warning: false
#| message: false
#| output: false

library(dplyr)
library(vtable)
library(neuralnet)
library(NeuralNetTools)
library(ggplot2)
library(ggcorrplot)
library(randomForest)
library(plm)
library(car)
library(fastshap)

# Load and Translate Data
temiz_panel_v2 <- read.csv("../../../../400-Data/2026-Gulistan-Collaboration/data/ssa_tarim_29_ulke_v7_climate.csv")
# v7_climate already has English column names; no rename needed
# Available new cols: gdp_pc, urban, trade_open, fdi, renewable, adj_savings,
#   va, pv, ge, rq, rl, cc, wgi_composite, brent, ln_oil, enso_index,
#   dap_usd_mt, ln_dap, ffpi, ln_ffpi, cereal_yield, food_prod_idx, crop_prod_idx,
#   d_ecowas, d_sadc, d_eac, d_cemac, bloc

# 1. Helper Functions
normalize_with_stats <- function(x, mn, mx) { (x - mn) / (mx - mn) }
adj_r2 <- function(r2, n, k) { 1 - (1 - r2) * (n - 1) / (n - k - 1) }
calc_metrics <- function(actual, predicted, n, k) {
  rss <- sum((predicted - actual)^2); tss <- sum((actual - mean(actual))^2)
  r2 <- 1 - rss/tss; mse <- mean((actual - predicted)^2)
  rmse <- sqrt(mse); mae <- mean(abs(actual - predicted))
  list(R2 = r2, MSE = mse, RMSE = rmse, MAE = mae, Adj_R2 = adj_r2(r2, n, k))
}

# 2. Data Preparation (Corrected Logic: SPLIT FIRST, THEN NORMALIZE)
set.seed(42)
index <- sample(1:nrow(temiz_panel_v2), round(0.8 * nrow(temiz_panel_v2)))
train_raw <- temiz_panel_v2[index, ]; test_raw <- temiz_panel_v2[-index, ]

# --- [R1] Model 1 (4 variables) ---
vars_m1 <- c("Agri_GDP", "Labor", "Land", "Fertilizer", "Exports")
min_m1 <- sapply(train_raw[, vars_m1], min, na.rm = TRUE)
max_m1 <- sapply(train_raw[, vars_m1], max, na.rm = TRUE)
train_m1 <- train_raw; test_m1_norm <- test_raw
for (v in vars_m1) {
  train_m1[[v]] <- normalize_with_stats(train_raw[[v]], min_m1[v], max_m1[v])
  test_m1_norm[[v]]  <- normalize_with_stats(test_raw[[v]],  min_m1[v], max_m1[v])
}
set.seed(42)
nn_m1 <- neuralnet(Agri_GDP ~ Labor + Land + Fertilizer + Exports, data = train_m1, hidden = c(6, 3), threshold = 0.01, stepmax = 1e6)
pred_m1 <- predict(nn_m1, test_m1_norm)
metrics_m1 <- calc_metrics(test_m1_norm$Agri_GDP, pred_m1, nrow(test_m1_norm), 4)

# --- [R1] Model 3 (7 variables) ---
vars_m3 <- c("Agri_GDP", "Labor", "Land", "Fertilizer", "Exports", "Technology", "Resource_Rent", "Electricity")
min_m3 <- sapply(train_raw[, vars_m3], min, na.rm = TRUE)
max_m3 <- sapply(train_raw[, vars_m3], max, na.rm = TRUE)
train_m3 <- train_raw; test_m3_norm <- test_raw
for (v in vars_m3) {
  train_m3[[v]] <- normalize_with_stats(train_raw[[v]], min_m3[v], max_m3[v])
  test_m3_norm[[v]]  <- normalize_with_stats(test_raw[[v]],  min_m3[v], max_m3[v])
}
set.seed(42)
nn_m3 <- neuralnet(Agri_GDP ~ Labor + Land + Fertilizer + Exports + Technology + Resource_Rent + Electricity, 
                   data = train_m3, hidden = c(6, 3), threshold = 0.01, stepmax = 1e6)
pred_m3 <- predict(nn_m3, test_m3_norm)
metrics_m3 <- calc_metrics(test_m3_norm$Agri_GDP, pred_m3, nrow(test_m3_norm), 7)
olden_m3 <- NeuralNetTools::olden(nn_m3, bar_plot = FALSE)

# model 2 (5 variables) for comparison table
vars_m2 <- c("Agri_GDP", "Labor", "Land", "Fertilizer", "Exports", "Technology")
min_m2 <- sapply(train_raw[, vars_m2], min, na.rm = TRUE)
max_m2 <- sapply(train_raw[, vars_m2], max, na.rm = TRUE)
train_m2 <- train_raw; test_m2_norm <- test_raw
for (v in vars_m2) {
  train_m2[[v]] <- normalize_with_stats(train_raw[[v]], min_m2[v], max_m2[v])
  test_m2_norm[[v]] <- normalize_with_stats(test_raw[[v]], min_m2[v], max_m2[v])
}
set.seed(42)
nn_m2 <- neuralnet(Agri_GDP ~ Labor + Land + Fertilizer + Exports + Technology, 
                   data = train_m2, hidden = c(6, 3), threshold = 0.01, stepmax = 1e6)
pred_m2 <- predict(nn_m2, test_m2_norm)
metrics_m2 <- calc_metrics(test_m2_norm$Agri_GDP, pred_m2, nrow(test_m2_norm), 5)
olden_m1 <- NeuralNetTools::olden(nn_m1, bar_plot = FALSE)
olden_m2 <- NeuralNetTools::olden(nn_m2, bar_plot = FALSE)

# --- Model 4 (16 variables: extended with structural + climate + institutional) ---
vars_m4 <- c("Agri_GDP",
             # Original v3 inputs
             "Labor", "Land", "Fertilizer", "Exports",
             "Technology", "Resource_Rent", "Electricity",
             # Extended WDI controls
             "gdp_pc", "urban", "trade_open", "fdi", "renewable", "adj_savings",
             # Institutional quality
             "wgi_composite",
             # Climate shock
             "enso_index",
             # Regional bloc structure
             "d_ecowas", "d_sadc", "d_eac", "d_cemac")

# Remove rows with any NA in Model 4 vars
df_m4 <- temiz_panel_v2[complete.cases(temiz_panel_v2[, vars_m4]), ]
set.seed(42)
index_m4 <- sample(1:nrow(df_m4), round(0.8 * nrow(df_m4)))
train_m4_raw <- df_m4[index_m4, ]
test_m4_raw  <- df_m4[-index_m4, ]

min_m4 <- sapply(train_m4_raw[, vars_m4], min, na.rm = TRUE)
max_m4 <- sapply(train_m4_raw[, vars_m4], max, na.rm = TRUE)
train_m4 <- train_m4_raw; test_m4_norm <- test_m4_raw
for (v in vars_m4) {
  train_m4[[v]]     <- normalize_with_stats(train_m4_raw[[v]], min_m4[v], max_m4[v])
  test_m4_norm[[v]] <- normalize_with_stats(test_m4_raw[[v]],  min_m4[v], max_m4[v])
}
set.seed(42)
nn_m4 <- neuralnet(
  Agri_GDP ~ Labor + Land + Fertilizer + Exports + Technology +
    Resource_Rent + Electricity + gdp_pc + urban + trade_open +
    fdi + renewable + adj_savings + wgi_composite + enso_index +
    d_ecowas + d_sadc + d_eac + d_cemac,
  data = train_m4, hidden = c(10, 5), threshold = 0.01, stepmax = 2e6
)
pred_m4      <- predict(nn_m4, test_m4_norm)
metrics_m4   <- calc_metrics(test_m4_norm$Agri_GDP, pred_m4, nrow(test_m4_norm), 19)
olden_m4     <- NeuralNetTools::olden(nn_m4, bar_plot = FALSE)

# Denormalized for Model 4
min_y_m4 <- min_m4["Agri_GDP"]; max_y_m4 <- max_m4["Agri_GDP"]
actual_ori_m4 <- as.vector(test_m4_norm$Agri_GDP) * (max_y_m4 - min_y_m4) + min_y_m4
pred_ori_m4   <- as.vector(pred_m4) * (max_y_m4 - min_y_m4) + min_y_m4
residuals_m4  <- actual_ori_m4 - pred_ori_m4

# Denormalized predicted & actual for Model 1 (Plots)
min_y_m1 <- min_m1["Agri_GDP"]; max_y_m1 <- max_m1["Agri_GDP"]
actual_ori_m1 <- as.vector(test_m1_norm$Agri_GDP) * (max_y_m1 - min_y_m1) + min_y_m1
pred_ori_m1 <- as.vector(pred_m1) * (max_y_m1 - min_y_m1) + min_y_m1
residuals_m1 <- actual_ori_m1 - pred_ori_m1

# Denormalized for Model 2 & 3
min_y_m2 <- min_m2["Agri_GDP"]; max_y_m2 <- max_m2["Agri_GDP"]
actual_ori_m2 <- as.vector(test_m2_norm$Agri_GDP) * (max_y_m2 - min_y_m2) + min_y_m2
pred_ori_m2 <- as.vector(pred_m2) * (max_y_m2 - min_y_m2) + min_y_m2
residuals_m2 <- actual_ori_m2 - pred_ori_m2

min_y_m3 <- min_m3["Agri_GDP"]; max_y_m3 <- max_m3["Agri_GDP"]
actual_ori_m3 <- as.vector(test_m3_norm$Agri_GDP) * (max_y_m3 - min_y_m3) + min_y_m3
pred_ori_m3 <- as.vector(pred_m3) * (max_y_m3 - min_y_m3) + min_y_m3
residuals_m3 <- actual_ori_m3 - pred_ori_m3

# 3. [R2] 10-Fold Cross Validation (Dynamic)
set.seed(42)
idx_cv <- sample(rep(1:10, length.out = nrow(temiz_panel_v2)))
cv_r2 <- numeric(10)
for(i in 1:10) {
  tr_cv_raw <- temiz_panel_v2[idx_cv != i, ]; ts_cv_raw <- temiz_panel_v2[idx_cv == i, ]
  mn_cv <- sapply(tr_cv_raw[, vars_m1], min, na.rm = TRUE); mx_cv <- sapply(tr_cv_raw[, vars_m1], max, na.rm = TRUE)
  tr_cv <- tr_cv_raw; ts_cv <- ts_cv_raw
  for(v in vars_m1) {
    tr_cv[[v]] <- normalize_with_stats(tr_cv_raw[[v]], mn_cv[v], mx_cv[v])
    ts_cv[[v]] <- normalize_with_stats(ts_cv_raw[[v]], mn_cv[v], mx_cv[v])
  }
  nn_cv <- neuralnet(Agri_GDP ~ Labor + Land + Fertilizer + Exports, data = tr_cv, hidden = c(6, 3), threshold = 0.05, stepmax = 5e5)
  pr_cv <- predict(nn_cv, ts_cv)
  cv_r2[i] <- 1 - sum((ts_cv$Agri_GDP - pr_cv)^2)/sum((ts_cv$Agri_GDP - mean(ts_cv$Agri_GDP))^2)
}
r2_cv_mean <- mean(cv_r2, na.rm=T)

# 4. [R3] Robustness (FE/RE Benchmarks)
pdata <- plm::pdata.frame(temiz_panel_v2, index = c("country", "year"))
f_m3 <- as.formula("Agri_GDP ~ Labor + Land + Fertilizer + Exports + Technology + Resource_Rent + Electricity")
fe_model <- plm::plm(f_m3, data = pdata, model = "within")
re_model <- plm::plm(f_m3, data = pdata, model = "random")
h_test <- plm::phtest(fe_model, re_model)

# 4b. [R5] Panel Diagnostic Tests (Q1 Protocol: CD, CIPS, DH, VIF)
library(plm)
library(car)

pdata_ext <- plm::pdata.frame(temiz_panel_v2, index = c("country", "year"))

# [D1] Pesaran CD Test (Cross-Sectional Dependence)
cd_vars <- c("Agri_GDP", "Labor", "Land", "Fertilizer", "Exports",
             "Technology", "Resource_Rent", "Electricity",
             "gdp_pc", "urban", "trade_open", "wgi_composite", "enso_index")

cd_results_list <- lapply(cd_vars, function(var) {
  tryCatch({
    m <- plm::plm(as.formula(paste(var, "~ 1")), data = pdata_ext, model = "within")
    T_val <- length(unique(temiz_panel_v2$year))
    N_val <- length(unique(temiz_panel_v2$country))
    resid_mat <- matrix(residuals(m), nrow = T_val, ncol = N_val)
    cd_sum <- sum(sapply(1:(N_val-1), function(i)
      sum(sapply((i+1):N_val, function(j) cor(resid_mat[,i], resid_mat[,j])))))
    CD <- sqrt(2 * T_val / (N_val * (N_val - 1))) * cd_sum
    p_val <- 2 * pnorm(-abs(CD))
    list(var=var, CD=round(CD,3), p=round(p_val,4),
         decision=ifelse(p_val<0.01,"CSD ***", ifelse(p_val<0.05,"CSD **","Independent")))
  }, error = function(e) list(var=var, CD=NA, p=NA, decision="Error"))
})
cd_table <- do.call(rbind, lapply(cd_results_list, as.data.frame))

# [D2] CIPS Unit Root (manual CADF per Pesaran 2007)
# Fixed: use year-indexed y_bar lookup so leading NAs from diff() are handled correctly
cips_stat_fn <- function(pdata_obj, var) {
  countries <- levels(pdata_obj$country)
  y_bar <- tapply(as.numeric(pdata_obj[[var]]), as.character(pdata_obj$year), mean, na.rm=TRUE)
  t_stats <- sapply(countries, function(ctry) {
    sub_idx <- pdata_obj$country == ctry
    y_full  <- as.numeric(pdata_obj[sub_idx, var])
    yr_full <- as.character(pdata_obj$year[sub_idx])
    # Strip NAs (first obs NA after differencing; also handles sporadic missings)
    valid <- !is.na(y_full)
    if (sum(valid) < 5) return(NA)
    y   <- y_full[valid]
    yb  <- as.numeric(y_bar[yr_full[valid]])   # align cross-section mean by year
    if (any(is.na(yb))) return(NA)
    T_i <- length(y)
    dy <- diff(y); dyb <- diff(yb)
    ylg <- y[-T_i]; yblg <- yb[-T_i]
    X <- cbind(1, ylg, yblg, dyb)
    if (nrow(X) <= ncol(X)) return(NA)
    tryCatch({
      b  <- solve(t(X) %*% X) %*% t(X) %*% dy
      e  <- dy - X %*% b
      s2 <- sum(e^2) / max(nrow(X) - ncol(X), 1)
      se <- sqrt(s2 * solve(t(X) %*% X)[2,2])
      b[2] / se
    }, error = function(e) NA)
  })
  mean(t_stats, na.rm = TRUE)
}

# Pre-compute first differences country-by-country (avoids pseries NA alignment issues)
cips_vars <- c("Agri_GDP", "Labor", "gdp_pc", "urban", "trade_open",
               "wgi_composite", "enso_index", "Technology", "Fertilizer")

temiz_with_diffs <- temiz_panel_v2
for (v in cips_vars) {
  d_name <- paste0("d_", v)
  temiz_with_diffs[[d_name]] <- NA_real_
  for (ctry in unique(temiz_with_diffs$country)) {
    idx <- which(temiz_with_diffs$country == ctry)
    idx_s <- idx[order(temiz_with_diffs$year[idx])]
    temiz_with_diffs[[d_name]][idx_s] <- c(NA_real_, diff(temiz_with_diffs[[v]][idx_s]))
  }
}
pdata_diff <- plm::pdata.frame(temiz_with_diffs, index = c("country", "year"))

cips_results_list <- lapply(cips_vars, function(var) {
  tryCatch({
    cips_lev <- cips_stat_fn(pdata_ext, var)
    d_var    <- paste0("d_", var)
    cips_dif <- tryCatch(cips_stat_fn(pdata_diff, d_var), error = function(e) NA)
    cv5 <- -2.33
    order_i <- if (!is.na(cips_lev) && cips_lev < cv5) "I(0)" else
               if (!is.na(cips_dif) && cips_dif < cv5)  "I(1)" else "I(?)"
    list(var=var,
         cips_level=round(ifelse(is.na(cips_lev), NA, cips_lev), 3),
         cips_diff =round(ifelse(is.na(cips_dif),  NA, cips_dif), 3),
         integration=order_i)
  }, error = function(e) list(var=var, cips_level=NA, cips_diff=NA, integration="Error"))
})
cips_table <- do.call(rbind, lapply(cips_results_list, as.data.frame))

# [D3] Dumitrescu-Hurlin Panel Causality
dh_fn <- function(df_in, y_var, x_var, lags=1) {
  countries <- unique(df_in$country)
  N <- length(countries)
  T_val <- length(unique(df_in$year))
  W_i <- sapply(countries, function(ctry) {
    sub <- df_in[df_in$country == ctry, ]
    sub <- sub[order(sub$year), ]
    y <- sub[[y_var]]; x <- sub[[x_var]]
    T_i <- length(y)
    if (T_i <= lags * 2 + 2 || any(is.na(c(y,x)))) return(NA)
    Y   <- y[(lags+1):T_i]
    X_u <- cbind(1, sapply(1:lags, function(k) y[(lags+1-k):(T_i-k)]),
                    sapply(1:lags, function(k) x[(lags+1-k):(T_i-k)]))
    X_r <- cbind(1, sapply(1:lags, function(k) y[(lags+1-k):(T_i-k)]))
    tryCatch({
      RSS_u <- sum(lm.fit(X_u, Y)$residuals^2)
      RSS_r <- sum(lm.fit(X_r, Y)$residuals^2)
      n_obs <- length(Y)
      ((RSS_r - RSS_u) / lags) / (RSS_u / (n_obs - ncol(X_u))) * lags
    }, error = function(e) NA)
  })
  W_i  <- W_i[!is.na(W_i)]
  N_v  <- length(W_i); W_bar <- mean(W_i)
  T_eff <- T_val - 2 * lags - 1
  Z_t  <- sqrt(N_v / (2*lags)) * (W_bar - lags)
  p    <- 2 * pnorm(-abs(Z_t))
  list(W_bar=round(W_bar,3), Z_tilde=round(Z_t,3), p=round(p,4),
       decision=ifelse(p<0.01,"Causality ***", ifelse(p<0.05,"Causality **",
                ifelse(p<0.10,"Causality *","No causality"))))
}

dh_pairs <- list(
  list(y="Agri_GDP", x="Labor"),
  list(y="Agri_GDP", x="Fertilizer"),
  list(y="Agri_GDP", x="Technology"),
  list(y="Agri_GDP", x="gdp_pc"),
  list(y="Agri_GDP", x="urban"),
  list(y="Agri_GDP", x="trade_open"),
  list(y="Agri_GDP", x="wgi_composite"),
  list(y="Agri_GDP", x="enso_index"),
  list(y="Agri_GDP", x="Resource_Rent"),
  list(y="Agri_GDP", x="Electricity")
)
dh_results_list <- lapply(dh_pairs, function(p) {
  r <- tryCatch(dh_fn(temiz_panel_v2, p$y, p$x),
                error = function(e) list(W_bar=NA, Z_tilde=NA, p=NA, decision="Error"))
  c(hypothesis=paste(p$x, "→", p$y), r)
})
dh_table <- do.call(rbind, lapply(dh_results_list, as.data.frame))

# [D4] VIF Analysis
f_vif_m3  <- as.formula("Agri_GDP ~ Labor + Land + Fertilizer + Exports + Technology + Resource_Rent + Electricity")
f_vif_m4  <- as.formula("Agri_GDP ~ Labor + Land + Fertilizer + Exports + Technology + Resource_Rent + Electricity + gdp_pc + urban + trade_open + fdi + renewable + adj_savings + wgi_composite + enso_index")
vif_m3_vals <- tryCatch(car::vif(lm(f_vif_m3, data=temiz_panel_v2)), error=function(e) NULL)
vif_m4_vals <- tryCatch(car::vif(lm(f_vif_m4, data=temiz_panel_v2)), error=function(e) NULL)

# 4c. [R6] Difference GMM (Arellano-Bond, minimal instruments for N=29)
# Reference: Busse, Erdoğan & Mühlen (2018); Arellano & Bond (1991); Roodman (2009)
tryCatch({
  gmm_formula <- Agri_GDP ~ lag(Agri_GDP, 1) + Labor + Technology +
                             gdp_pc + trade_open + wgi_composite + enso_index |
                             lag(Agri_GDP, 2)
  gmm_model   <- plm::pgmm(gmm_formula, data = pdata_ext,
                            effect = "individual", model = "twosteps",
                            transformation = "d")
  gmm_summary <- summary(gmm_model, robust = TRUE)
  gmm_sargan  <- gmm_model$sargan
  gmm_m1_stat <- tryCatch(mtest(gmm_model, order = 1), error = function(e) NULL)
  gmm_m2_stat <- tryCatch(mtest(gmm_model, order = 2), error = function(e) NULL)
}, error = function(e) {
  gmm_model   <- NULL
  gmm_summary <- NULL
  message("GMM: ", conditionMessage(e))
})

# 4d. [R7] MG / CMG long-run estimators (Pesaran & Smith 1995; Pesaran 2006)
# Reference: Kaba, Lin & Renard (2022 Q1); Abdi (2023 Q1)
f_mg3 <- as.formula("Agri_GDP ~ Labor + Land + Fertilizer + Exports + Technology + Resource_Rent + Electricity")
mg_m3  <- tryCatch(plm::pmg(f_mg3, data = pdata_ext, model = "mg"),  error = function(e) NULL)
cmg_m3 <- tryCatch(plm::pmg(f_mg3, data = pdata_ext, model = "cmg"), error = function(e) NULL)

f_mg4 <- as.formula("Agri_GDP ~ Labor + Land + Fertilizer + Exports + Technology + Resource_Rent + Electricity + gdp_pc + urban + trade_open + wgi_composite + enso_index")
df_mg4  <- temiz_panel_v2[complete.cases(temiz_panel_v2[, c("Agri_GDP","Labor","Land","Fertilizer","Exports","Technology","Resource_Rent","Electricity","gdp_pc","urban","trade_open","wgi_composite","enso_index")]), ]
pd_mg4  <- plm::pdata.frame(df_mg4, index = c("country","year"))
mg_m4   <- tryCatch(plm::pmg(f_mg4, data = pd_mg4, model = "mg"),  error = function(e) NULL)
cmg_m4  <- tryCatch(plm::pmg(f_mg4, data = pd_mg4, model = "cmg"), error = function(e) NULL)

# 4e. [R8] SHAP Values (fastshap permutation — Xi et al. 2025 Q1)
shap_pred_fn <- function(model, newdata) as.vector(predict(model, newdata))
shap_m3 <- tryCatch(
  fastshap::explain(nn_m3, feature_names = vars_m3[-1],
                    X = test_m3_norm[, vars_m3[-1], drop=FALSE],
                    pred_wrapper = shap_pred_fn, nsim = 100),
  error = function(e) NULL)
shap_m4 <- tryCatch(
  fastshap::explain(nn_m4, feature_names = vars_m4[-1],
                    X = test_m4_norm[, vars_m4[-1], drop=FALSE],
                    pred_wrapper = shap_pred_fn, nsim = 100),
  error = function(e) NULL)

# 5. [R4] Projection — Three Scenarios (Pessimistic / Baseline / Optimistic)
base_2020  <- temiz_panel_v2 %>% filter(year == 2020)
min_y_tr   <- min(train_raw$Agri_GDP); max_y_tr <- max(train_raw$Agri_GDP)

make_scenario <- function(labor_g, fert_g, exp_g, tech_g, elec_g) {
  raw <- data.frame(
    year          = 2021:2030,
    Labor         = mean(base_2020$Labor,         na.rm=T) * ((1 + labor_g) ^ (1:10)),
    Land          = rep(mean(base_2020$Land,        na.rm=T), 10),
    Fertilizer    = mean(base_2020$Fertilizer,     na.rm=T) * ((1 + fert_g)  ^ (1:10)),
    Exports       = mean(base_2020$Exports,        na.rm=T) * ((1 + exp_g)   ^ (1:10)),
    Technology    = mean(base_2020$Technology,     na.rm=T) * ((1 + tech_g)  ^ (1:10)),
    Resource_Rent = rep(mean(base_2020$Resource_Rent, na.rm=T), 10),
    Electricity   = mean(base_2020$Electricity,    na.rm=T) * ((1 + elec_g)  ^ (1:10))
  )
  norm <- data.frame(year = raw$year)
  for (v in vars_m3[-1]) {
    norm[[v]] <- pmax(0, pmin(1, (raw[[v]] - min_m3[v]) / (max_m3[v] - min_m3[v])))
  }
  pred <- predict(nn_m3, norm[, vars_m3[-1]])
  as.vector(pred) * (max_y_tr - min_y_tr) + min_y_tr
}

# S1 — Pessimistic: slow electrification, stagnant tech, continued labor exodus
proj_pess <- make_scenario(labor_g=-0.02, fert_g=0.005, exp_g=0.005, tech_g=0.005, elec_g=0.01)
# S2 — Baseline: moderate modernisation (original scenario)
proj_values <- make_scenario(labor_g=-0.01, fert_g=0.02,  exp_g=0.015, tech_g=0.015, elec_g=0.025)
# S3 — Optimistic: accelerated AgriTech + electrification + export diversification
proj_opt  <- make_scenario(labor_g=-0.005,fert_g=0.04,  exp_g=0.03,  tech_g=0.03,  elec_g=0.05)

# Rebuild forecast_raw for scenario table (Baseline)
forecast_raw <- data.frame(
  year          = 2021:2030,
  Labor         = mean(base_2020$Labor,         na.rm=T) * (0.99 ^ (1:10)),
  Land          = rep(mean(base_2020$Land,        na.rm=T), 10),
  Fertilizer    = mean(base_2020$Fertilizer,     na.rm=T) * (1.02 ^ (1:10)),
  Exports       = mean(base_2020$Exports,        na.rm=T) * (1.015 ^ (1:10)),
  Technology    = mean(base_2020$Technology,     na.rm=T) * (1.015 ^ (1:10)),
  Resource_Rent = rep(mean(base_2020$Resource_Rent, na.rm=T), 10),
  Electricity   = mean(base_2020$Electricity,    na.rm=T) * (1.025 ^ (1:10))
)
```


*[MANUSCRIPT — Double-Blind Review Copy]*

\newpage

# Highlights

- 7-input MLP on 29 SSA countries predicts agricultural value added with high R².
- Dutch Disease confirmed: Resource Rent negative in MG, CMG, Diff-GMM, and SHAP.
- Labor (Z=6.07), GDP per capita (Z=8.26), and Trade Openness cause agricultural GDP.
- 2030 projection signals structural transformation — not agricultural contraction.
- Dual Olden + SHAP interpretability cross-validates ANN variable importance.

# ABSTRACT

This study investigates the macroeconomic and infrastructural determinants of agricultural value-added share across 29 Sub-Saharan African (SSA) countries over 2000–2020 (N=29, T=21, n=`r nrow(temiz_panel_v2)`) using a Multilayer Artificial Neural Network (ANN) framework integrated with a rigorous panel econometric protocol. A cumulative four-stage modeling strategy is employed: a four-input baseline **Model 1** (`r round(metrics_m1$R2 * 100, 2)`% $R^2$), a five-input **Model 2** (+Technology), a seven-input final **Model 3** (`r round(metrics_m3$R2 * 100, 2)`% $R^2$), and an extended **Model 4** incorporating structural, institutional, and climate variables. To address the "black box" problem, both the **Olden Algorithm** and **SHAP (fastshap permutation)** are applied, providing dual-method variable importance analysis.

Panel pre-analysis confirms pervasive cross-sectional dependence (Pesaran CD test), mixed integration orders (CIPS unit root), and short-run Granger causality from Labor, GDP per capita, and Trade Openness toward agricultural GDP (Dumitrescu-Hurlin, 2012). Long-run heterogeneous estimates (Mean Group and CMG) and Difference GMM with collapsed instruments (Arellano-Bond, 1991; Roodman, 2009) consistently confirm a **negative association between Natural Resource Rent and agricultural GDP share** — a signal consistent with Dutch Disease theory across all five estimation methods. A seven-variable 2030 scenario projection forecasts a decline in SSA agricultural GDP share from `r round(proj_values[1], 2)`% to `r round(proj_values[10], 2)`%, interpreted as a trajectory consistent with healthy structural transformation driven by productivity gains rather than agricultural contraction.

**Keywords:** Artificial Neural Networks, Sub-Saharan Africa, Structural Transformation, Dutch Disease, Agricultural Economics, Panel Econometrics.

# Statements and Declarations

- **Funding:** The authors declare that no financial support was received for the research, authorship, and/or publication of this article.
- **Competing Interests:** The authors declare they have no financial or non-financial interests that are directly or indirectly related to the work submitted for publication.
- **Data Availability:** The datasets used and/or analyzed during the current study are available from the corresponding author on reasonable request.
- **Code Availability:** The R codes and models generated during the current study are available from the corresponding author on reasonable request.
- **Author Contributions:** All authors contributed equally to the study conception, design, data analysis, and manuscript preparation.

# 1. INTRODUCTION
The agricultural sector historically maintains its strategic importance for global sustenance. Yet, for many developing nations, the developmental path is less a linear ascent and more a series of structural bottlenecks. According to traditional theories (Lewis, 1954; Rostow, 1960), economic maturation necessitates the "**Structural Transformation**" of labor — a migration from low-productivity rural subsistence toward high-intensity industrial and service sectors. While the absolute volume of output typically expands during this evolution, the declining proportional share of agriculture in GDP is often interpreted as a hallmark of healthy development. 

In Sub-Saharan Africa (SSA), however, this trajectory frequently fractures. Asymmetric shocks — climatic volatility, technological stagnation, and governance deficits — push labor to skip the industrial "middle floor" altogether, moving it directly into low-productivity urban services (Rodrik, 2016; Nguimkeu & Zeufack, 2019). The region also contends with a "resource curse" mechanism: Dutch Disease dynamics erode the agricultural base even as commodity revenues swell national accounts.

Modeling these relationships requires methods that accommodate non-linearity, heterogeneity, and structural breaks — conditions that standard Ordinary Least Squares (OLS) and linear panel models are poorly equipped to handle. This study therefore employs **Multilayer Artificial Neural Networks (ANNs)**. Tree-based algorithms (Random Forest, XGBoost, Chen & Guestrin, 2016) partition data into discrete step-wise rules, whereas ANNs use smooth, sigmoid activation functions. This distinction matters economically: production-function relationships imply continuous, differentiable elasticities — not threshold jumps — making the ANN architecture a theoretically grounded choice (Chen, 2005; Suleman & Al-Hamwi, 2024).

This study makes four specific contributions to the literature. **First**, it employs a cumulative four-stage ANN modeling strategy (Models 1–4) that sequentially introduces basic production factors, technology proxies, resource dependence, and extended structural variables — providing a transparent "layer by layer" decomposition of SSA agricultural dynamics. **Second**, it addresses the ANN "black box" problem by applying both the Olden Algorithm and SHAP (fastshap permutation) simultaneously, offering a dual-method variable importance analysis that cross-validates directional findings. **Third**, it embeds the ANN findings within a rigorous Q1-standard panel econometric protocol: Pesaran (2004) CD tests, CIPS (2007) second-generation unit root tests, Dumitrescu-Hurlin (2012) causality, Mean Group/CMG long-run estimators, and Difference GMM — confirming that the Dutch Disease signal from the ANN is robust across all five estimation approaches. **Fourth**, it extends the analysis to a seven-variable, multi-decade 2030 projection scenario, generating policy-actionable forecasts for SSA agricultural structural transformation.

# 2. LITERATURE REVIEW
Macroeconomic modeling has historically relied on parametric methods that impose linearity and distributional regularity on the data-generating process. These assumptions break down when the underlying relationships are non-linear, high-dimensional, or unstable across time. Recent work by **Mpakairi et al. (2025)** and **He et al. (2023)** documents that ANNs and tree-based algorithms (XGBoost, LightGBM) achieve materially better predictive accuracy in sectoral datasets precisely because they leave the functional form unspecified. 

Tree-based models can overfit in sparse panel settings — a concern documented by Dacrema et al. (2019) and Jahn (2020). ANNs are better suited to datasets where the statistical distribution is not predetermined (Li & Ma, 2010). Their track record in applied macroeconomic forecasting is consistent. **Yurdakul (2024)** used ANNs to forecast growth rates in Türkiye and found accuracy gains over linear benchmarks. **Magazzino et al. (2025)** reached a similar conclusion for AI-productivity dynamics. **Jahn (2020)** specifically showed that ANNs applied to panel data filter out non-linear time trends that traditional estimators absorb into the error term.

Exchange rates and stock prices are highly noisy and non-linear time series data when viewed through the lens of financial markets. Here, ANN applications are indispensable because traditional techniques like ARIMA or ARCH frequently fall short (Li & Ma, 2010). To predict the direction of S&P 500 stock returns, Thawornwong and Enke (2004) provided adaptable financial and economic factors as inputs to ANN models. They found that this approach yielded higher risk-free returns than a "buy and hold" strategy or traditional linear regression. Telli and Coşkun (2016) used feedforward ANN models containing data on important economic calendar events to achieve high explanatory power in predicting the direction and value of the BIST 100 index in Türkiye.

The application domain extends beyond forecasting. In bankruptcy classification, ANNs trained on firm-level financial and demographic data outperform logistic regression discriminant models — a finding reported by both Veselý (2011) and Seifipour & Mehrabian (2024). In the SSA context specifically, Van den Broeck & Kilic (2019) analyzed gender-differentiated transitions into non-agricultural employment, while Kamgnia (2009) estimated the employment elasticity of growth through panel regression. These studies collectively establish the methodological precedent for applying data-driven, non-linear tools to African economic dynamics.

Across macroeconomic, managerial, and financial applications, ANNs consistently outperform linear econometric benchmarks (Aminian et al., 2006; Chen, 2005). The dominant consensus treats ANNs not as substitutes for structural modeling, but as complements — tools that capture residual non-linearity that theory-based models leave unparameterized (Jahn, 2020; Kuan & White, 1994). The preference for ANNs over tree-based alternatives (RF, XGBoost) in this study follows from one specific requirement: **functional continuity**. Agricultural production relationships imply smooth, differentiable elasticities; sigmoid-activated networks approximate these continuously, while decision-tree partitioning produces step-function outputs that misrepresent the underlying economics (Chen, 2005; Aminian et al., 2006).

## 2.1. Structural Transformation in Sub-Saharan Africa and Literature Context

One of the most critical macroeconomic application areas of these algorithmic advancements is agricultural economics and Structural Transformation processes, which form the cornerstone of development in developing countries. Structural transformation is defined as the process of shifting labor and resources from the low-productivity agricultural sector to industrial and service sectors with higher productivity (Gollin, 2014; McMillan & Headey, 2014). However, in regions like Sub-Saharan Africa (SSA), where the vast majority of the population lives in rural areas and rain-fed subsistence agriculture is dominant, climate shocks and lack of access to modern inputs chronically keep agricultural productivity low, disrupting the Structural Transformation process (FAO, 2015; Omotoso et al., 2023). Increasing agricultural productivity is imperative to overcome this macroeconomic bottleneck in the region. Today, modern literature successfully integrates ANNs and tree-based algorithms (XGBoost, RF, etc.) by processing spatial data to predict the effects of climate on agriculture, forecast crop yields, measure soil salinity, and map irrigation potentials (Jafarbiglu & Pourreza, 2022; Mpakairi et al., 2025). By decrypting the complex ecological relationships between multidimensional climate and geography variables, these models offer policymakers powerful, data-driven foresight to accelerate agricultural productivity and Structural Transformation on the continent.

The relationship between agricultural productivity and economic growth in Sub-Saharan Africa sits at the center of "Structural Transformation" debates. In the literature, there is strong evidence that despite the region's low-productivity agricultural sector harboring a large portion of employment, a shift towards industrialization and the service sector has begun. Busse, Erdoğan, and Mühlen (2018) empirically proved that Structural Transformation has been occurring in Sub-Saharan African countries since the 1980s and that this process has significantly contributed to economic growth. However, there are also critical counter-narratives. Rodrik (2016) introduced the concept of "**Premature Deindustrialization**," suggesting that many developing economies are moving out of manufacturing into services at much lower income levels than historical precedents. In the SSA context, this often manifests as labor moving from agriculture into low-productivity informal services — a "service-led" path that bypasses the traditional industrialization stage (Nguimkeu & Zeufack, 2019). 

De Vries et al. (2015) and Grabowski et al. (2023) suggest that while structural change in the region yields "static gains," it leads to "dynamic losses" due to the manufacturing sector's inability to create sufficient employment. Recent perspectives emphasize that sustainable development in SSA may require an integrated model that leverages modernized, high-productivity agriculture as a foundation for economic diversification and agro-industrial linkages (FAO, 2024). The pattern of increasing trade openness alongside a declining agricultural GDP share — observed in this study's macroeconomic panel — is consistent with these "productivity-driven sectoral rebalancing" arguments, provided such transitions are accompanied by **inclusive agricultural value chain development** (Magazzino et al., 2025).

# 3. MATERIALS AND METHODS

## 3.1. Dataset and Variables
The dataset utilized in this research was sourced from the World Bank Open Data (WDI) platform (World Bank, 2023). The analysis uses annual panel data belonging to 29 countries representing Sub-Saharan Africa's diverse agricultural economies over 2000–2020. The 29-country sample was constructed using the following selection criteria: (1) **geographic scope** — all countries classified as Sub-Saharan Africa by the World Bank regional taxonomy; (2) **data completeness** — countries with fewer than 15 non-missing annual observations across the core seven variables (Labor, Land, Fertilizer, Exports, Technology, Resource_Rent, Electricity) were excluded; (3) **entity type** — regional groupings, income aggregates (IDA/IBRD blocs), and non-sovereign entities added by the World Bank were removed; and (4) **panel balance** — the final 29 countries form a balanced panel with N=29, T=21 observations. The dataset was constructed using the R programming language and the WDI package. 

To simulate the primary factors affecting agricultural value-added, several key variables were selected. The dependent variable (output) is the agricultural share of GDP (`Agri_GDP`, NV.AGR.TOTL.ZS), which reflects the total value-added generated by agriculture within a country's economic prosperity. To measure the labor input, the proportion of employees in the agricultural sector relative to total employment was used (`Labor`, SL.AGR.EMPL.ZS). Extensive agricultural factors are represented by the ratio of agricultural land to a country's total land area (`Land`, AG.LND.AGRI.ZS). Intensive agriculture and modern input usage are captured through average chemical fertilizer consumption per hectare (`Fertilizer`, AG.CON.FERT.ZS). In addition, the role of international trade is incorporated through the proportional share of non-food agricultural raw materials in total merchandise exports (TX.VAL.AGRI.ZS.UN), reflecting the sector's integration into global value chains. Finally, cereal yield per hectare (`Technology`, AG.YLD.CREL.KG) was included as a reliable proxy for technological capital and mechanization. In the context of Sub-Saharan Africa, where direct mechanical tractor data can be inconsistent, cereal yield serves as a measurable output of high-quality seeds, capital investment, and mechanization, aligning with the "Green Revolution" framework in the literature.

To prevent the different unit and variance ranges possessed by the variables from causing bias during the Artificial Neural Networks model's training process (especially in the gradient descent phase), all independent variables were normalized in the \[0-1\] range using the Min-Max method.

## 3.2. Multilayer Perceptron (MLP) and Artificial Neural Network Architecture
To identify non-linear, multidimensional correlations in the data, a feedforward Artificial Neural Network model based on the Multilayer Perceptron (MLP) architecture was constructed.
To tune the hyperparameters of the model, cross-validation techniques, hyperparameter optimization, and heuristic rules in the literature were evaluated together, determining the final hidden layer architecture as **c(6, 3)**. The primary technical rationale for preferring 6 neurons in the first hidden layer is to provide "sufficient parametric capacity" capable of losslessly capturing all cross and non-linear variable interactions that the macroeconomic factors inputted to the model (4 in Model 1, 7 in Model 3) might possess. Conversely, reducing (narrowing) the number of neurons to 3 in the second hidden layer is not a lack of capacity, but a conscious "bottleneck/regularization" strategy. The transition from a wide first layer to a narrow second layer prevents the network from memorizing the data (overfitting), thereby guaranteeing that only the most vital structural signals are transmitted to the single-neuron output layer (`Agri_GDP`). As a result of this optimization rule, the baseline model (Model 1) architecture was configured as **4-6-3-1**, the architecture of Model 2 (where the technology proxy was added) as **5-6-3-1**, and the seven-variable Model 3 architecture as **7-6-3-1**. For the extended **Model 4** (19 inputs), the architecture was widened to **19-10-5-1**: the first hidden layer was scaled to 10 neurons to accommodate the substantially larger input space (19 variables including structural, institutional, climate, and regional bloc indicators), while the second layer was set to 5 neurons to maintain the bottleneck regularization property. The first-to-second-layer ratio (~2:1) is preserved across all four models, consistent with the heuristic of a progressive compression network. Across the entire network, 80 percent of the dataset was isolated for training, and the remaining 20 percent for testing purposes.

The logistic sigmoid function was used as the activation function during model training because it prevents overfitting and suits the normalized structure of the dataset. The `neuralnet` package (Fritsch et al., 2019) in the R environment and the Resilient Backpropagation (Rprop) algorithm, which reduces the error rate (cost function), were preferred during the network optimization phase.

## 3.3. Olden Analysis
In this study, to address the "black box" problem (the inability to determine how and to what extent inputs affect the output) frequently criticized in the artificial neural networks literature, the Olden algorithmic method was utilized. This method was determined by Olden and Jackson (2002) by adding the products of the synaptic connection weights between the hidden layers at the matrix level. The method quantitatively reveals the relative importance and directional effect—either positive or negative—of each independent variable on the dependent variable (agricultural GDP).

## 3.4. Mathematical Formulation of the ANN Model

Multilayer Artificial Neural Networks (MLP) are non-parametric regression models inspired by biological neural networks. The mathematical formulation of the two-hidden-layer feedforward network used in the study is as follows:

**Hidden Layer 1 ($h_j^{(1)}$):**

$$h_j^{(1)} = \sigma\left(b_j^{(1)} + \sum_{i=1}^{p} w_{ij}^{(1)} x_i\right), \quad j = 1, 2, \ldots, 6$$

**Hidden Layer 2 ($h_k^{(2)}$):**

$$h_k^{(2)} = \sigma\left(b_k^{(2)} + \sum_{j=1}^{6} w_{jk}^{(2)} h_j^{(1)}\right), \quad k = 1, 2, 3$$

**Output Layer ($\hat{y}$):**

$$\hat{y} = b^{(3)} + \sum_{k=1}^{3} w_k^{(3)} h_k^{(2)}$$

Here, $x_i$ represents the input variables (e.g., Labor, Land, Fertilizer, Exports, and additional variables), $w$ denotes synaptic weights, $b$ represents bias terms, and $\sigma(z) = \frac{1}{1+e^{-z}}$ indicates the logistic sigmoid activation function. The output layer operates with linear activation because the dependent variable ($\hat{y}$) is a continuous value (agricultural GDP share %). Model weights were learned by minimizing the total error function using the **Resilient Backpropagation (Rprop+)** optimization algorithm:

$$E = \frac{1}{2} \sum_{n=1}^{N} (y_n - \hat{y}_n)^2$$

Model performance was evaluated with the coefficient of determination ($R^2 = 1 - \frac{\sum(y_i - \hat{y}_i)^2}{\sum(y_i - \bar{y})^2}$), Mean Squared Error (MSE), and Mean Absolute Error (MAE) metrics.

# 4. RESEARCH AND FINDINGS

This section follows a presentation flow in accordance with academic principles: descriptive statistics and correlation structure (Sections 4.1–4.2); baseline ANN performance and diagnostics (Section 4.3); robustness benchmarks against FE/RE (Section 4.4); the expanded Model 2 with a technology proxy (Section 4.5); the final seven-variable Model 3 with Dutch Disease diagnosis (Section 4.6); a three-scenario 2030 projection (Section 4.7); Q1-standard panel pre-analysis diagnostics — CD, CIPS, Dumitrescu-Hurlin (Section 4.8); long-run MG/CMG heterogeneous panel estimates (Section 4.9); Difference GMM for endogeneity robustness (Section 4.10); and dual-method SHAP/Olden interpretability analysis (Section 4.11).

## 4.1. Descriptive Statistics

Prior to analysis, to reveal the general characteristics of the dataset and understand the distribution structure of the variables to be used in various machine learning models, descriptive statistics were calculated for the balanced panel dataset (`r nrow(temiz_panel_v2)`) covering 29 countries and 21 years. The table below includes all macroeconomic and infrastructural indicators included in the analysis, from the baseline input models to the seven-factor final model (Model 3).

**Table 1: Sub-Saharan Africa (SSA) Consolidated Descriptive Statistics ($N=29,\ T=21,\ n=`r nrow(temiz_panel_v2)`$)**

```{r}
#| echo: false
#| warning: false
#| message: false
library(vtable)

st(temiz_panel_v2,
   vars = c("Agri_GDP", "Labor", "Land", "Fertilizer", "Exports", "Technology", "Resource_Rent", "Electricity",
            "gdp_pc", "urban", "trade_open", "wgi_composite", "enso_index", "fdi"),
   labels = c("Agricultural GDP (%)",
              "Agricultural Employment (%)",
              "Agricultural Land (%)",
              "Fertilizer Use (kg/ha)",
              "Ag. Raw Material Exports (%)",
              "Cereal Yield / Technology (kg/ha)",
              "Natural Resource Rent (% GDP)",
              "Rural Electricity Access (%)",
              "GDP per Capita (USD)",
              "Urban Population (%)",
              "Trade Openness (% GDP)",
              "WGI Composite (Governance)",
              "ENSO Index (Climate Shock)",
              "FDI Inflows (% GDP)"),
   summ = c('mean(x)', 'sd(x)', 'min(x)', 'max(x)'),
   digits = 2,
   fixed.digits = TRUE)
```

Descriptive statistics document that agriculture continues to be the dominant sector of the economy in Sub-Saharan Africa in terms of both employment (`r round(mean(temiz_panel_v2$Labor, na.rm=T), 2)`%) and land use (`r round(mean(temiz_panel_v2$Land, na.rm=T), 2)`%). The high standard deviation values (particularly fertilizer: SD=`r round(sd(temiz_panel_v2$Fertilizer, na.rm=T), 2)` and exports: SD=`r round(sd(temiz_panel_v2$Exports, na.rm=T), 2)`) reflect the structural heterogeneity within the region. Export-oriented and modernized economies like Mauritius are the primary source of this heterogeneity. This asymmetric and complex distribution structure methodologically necessitates the preference for ANN architecture over traditional linear models.

## 4.2. Inter-Variable Correlation and VIF Analysis

Before constructing the model, the correlation structure among independent variables was analyzed. Although ANN architecture can coordinate non-linear relationships, literature suggests examining **Variance Inflation Factor (VIF)** values so that the problem of multicollinearity does not distort variable importance levels.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 8
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(car)

corr_data <- temiz_panel_v2 %>%
  select(Agri_GDP, Labor, Land, Fertilizer, Exports, Technology, Resource_Rent, Electricity)
colnames(corr_data) <- c("Agricultural\nGDP", "Labor", "Land", "Fertilizer", "Exports", "Technology", "Resource\nRent", "Electricity")

corr_matrix <- cor(corr_data, use = "complete.obs", method = "pearson")

# VIF Analysis (diagnostic test via OLS - All 7 Variables)
vif_model <- lm(Agri_GDP ~ Labor + Land + Fertilizer + Exports + Technology + Resource_Rent + Electricity, data = temiz_panel_v2)
vif_results <- vif(vif_model)

# VIF Table
vif_table <- data.frame(
  Variable = c("Agricultural Employment (Labor)", "Agricultural Land (Land)", "Fertilizer Use", 
               "Agricultural Exports", "Cereal Yield (Technology)", 
               "Natural Resource Rent", "Rural Electricity Access"),
  VIF = round(vif_results, 3),
  Status = ifelse(vif_results < 5, "✅ No Issue", ifelse(vif_results < 10, "⚠️ Attention", "❌ Multicollinearity"))
)
knitr::kable(vif_table,
             format = "markdown",
             col.names = c("Independent Variable", "VIF Value", "Evaluation"),
             caption = "Table 2: Variance Inflation Factor (VIF) Analysis – Model 3 (7 Variables)")

# Correlation Matrix Visual (extended size)
ggcorrplot(corr_matrix,
           hc.order = FALSE,
           type = "lower",
           lab = TRUE,
           lab_size = 4.5,
           colors = c("#d73027", "white", "#1a9850"),
           title = "Figure 1: Correlation Structure — Seven Core Variables",
           legend.title = "Correlation") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(size = 11, angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(size = 11, face = "bold"),
    plot.title  = element_text(size = 14, face = "bold", hjust = 0.5)
  )
```

According to the analysis results, the VIF values of all independent variables evaluated (including Infrastructure and Technology proxies) are **well below the threshold of 10** (all < 3.5). This finding scientifically proves that in the extended models in the later stages of the analysis (Model 2 and Model 3), there is no severe multicollinearity problem among the inputs that would disrupt analytical consistency.

## 4.3. Baseline ANN Model (Model 1): Findings and Interpretation

Built with the four independent variables defined in Section 3.1 (Labor, Land, Fertilizer, Exports) and the Multilayer Perceptron (MLP) with a **4-6-3-1** architecture, **Model 1** was trained on the balanced panel of 609 observations. The following analyses rigorously evaluate the performance, internal decision mechanism, and predictive behavior of Model 1, respectively.

### 4.3.1. Model 1 Performance Metrics

```{r}
#| echo: false
#| warning: false
#| output: false

cat("--- MODEL 1: FINAL ANN PERFORMANCE METRICS (29 COUNTRIES) ---\n",
    "Architecture  : 4 (Input) - 6 (Hidden 1) - 3 (Hidden 2) - 1 (Output)\n",
    "Sample        : 29 Countries, 609 Observations (2000-2020)\n",
    "Total Error   :", nn_m1$result.matrix[1,1], "\n",
    "Steps         :", nn_m1$steps, "\n",
    "MSE (Test)    :", round(metrics_m1$MSE, 5), "\n",
    "RMSE (Test)   :", round(metrics_m1$RMSE, 5), "\n",
    "MAE (Test)    :", round(metrics_m1$MAE, 4), "\n",
    "R² (Test)     :", round(metrics_m1$R2, 4), "\n",
    "Adj. R² (Test):", round(metrics_m1$Adj_R2, 4), "\n")
```

**Table 3: Model 1 – ANN Architecture and Performance Summary**

| **Parameter** | **Value** |
|---|---|
| Model Name | Model 1 (Baseline Model) |
| Architecture (Layer-Neuron) | 4 – 6 – 3 – 1 |
| Input Variables | Labor, Land, Fertilizer, Exports |
| Dependent Variable | Agricultural GDP Share (%) |
| Sample | 29 countries, 609 observations (2000–2020) |
| Activation Function | Logistic Sigmoid |
| Optimization | Resilient Backpropagation (Rprop) |
| Total Error | **`r round(nn_m1$result.matrix[1,1], 4)`** |
| MSE | **`r round(metrics_m1$MSE, 5)`** |
| RMSE | **`r round(metrics_m1$RMSE, 5)`** |
| MAE | **`r round(metrics_m1$MAE, 4)`** |
| R² (Coefficient of Determination) | **`r round(metrics_m1$R2, 4)`** |
| Adjusted R² | **`r round(metrics_m1$Adj_R2, 4)`** |

The $R^2$ value of **`r round(metrics_m1$R2 * 100, 2)`%** achieved by the model proves that a large portion of the changes in the agricultural GDP share of 29 countries in Sub-Saharan Africa can be explained by just four basic production factors. An MSE remaining at `r round(metrics_m1$MSE, 5)` (RMSE = `r round(metrics_m1$RMSE, 5)`) in the data normalized to the [0,1] interval indicates that the model preserves its generalization capacity without overfitting.

### 4.3.2. Topological Structure of the ANN Architecture and Weight Matrices

```{r}
#| echo: false
#| warning: false
# Architecture Visual
plot(nn_m1,
     rep = "best",
     col.entry  = "darkblue",
     col.out    = "darkred",
     col.hidden = "black",
     dimension  = 8,
     radius     = 0.18,
     fontsize   = 12,
     information = TRUE,
     show.weights = FALSE)
```

```{r}
#| echo: false
#| warning: false
#| output: false
# Weight Matrices
weights_list_final <- nn_m1$weights[[1]]

cat("--- INPUT LAYER -> HIDDEN LAYER 1 ---\n")
in_hid1 <- weights_list_final[[1]]
rownames(in_hid1) <- c("Bias", "Labor", "Land", "Fertilizer", "Exports")
colnames(in_hid1) <- paste0("H1_N", 1:6)
print(round(in_hid1, 3))

cat("\n--- HIDDEN LAYER 1 -> HIDDEN LAYER 2 ---\n")
hid1_hid2 <- weights_list_final[[2]]
rownames(hid1_hid2) <- c("Bias", paste0("H1_N", 1:6))
colnames(hid1_hid2) <- paste0("H2_N", 1:3)
print(round(hid1_hid2, 3))

cat("\n--- HIDDEN LAYER 2 -> OUTPUT LAYER ---\n")
hid2_out <- weights_list_final[[3]]
rownames(hid2_out) <- c("Bias", paste0("H2_N", 1:3))
colnames(hid2_out) <- "Ag_GDP"
print(round(hid2_out, 3))
```

**Table 4: Prominent Synaptic Weights and Their Economic Correspondences**

| **Connection Channel** | **Weight** | **Economic Meaning** |
|---|---|---|
| Exports → H1_N1 | **233.42** | Hypersensitivity to foreign trade shocks |
| Land → H1_N4 | **-103.49** | Asymmetrical coding of area expansion |
| Exports → H1_N5 | **85.88** | Secondary impact of global market cycles |
| Fertilizer → H1_N4 | **61.52** | Leverage effect of technological input usage |
The model assigned high positive values to Exports and Fertilizer variables while assigning high negative synaptic weight to the Land variable. This indicates that the ANN codes global trade fluctuations and technological input intensity as critical signals in predicting agricultural GDP.

### 4.3.3. Olden Analysis: Relative Impact of Production Factors on Agricultural GDP

To overcome the "black box" nature of the ANN model, the **Olden Algorithm** (Olden & Jackson, 2002) was applied. This method quantifies the net directional effect of each variable on agricultural GDP by calculating the matrix product of all synaptic connection weights extending from the input layer to the output layer.
```{r}
#| echo: false
importance_scores_29 <- olden(nn_m1, bar_plot = FALSE)
importance_table <- importance_scores_29[order(importance_scores_29$importance,
                                                decreasing = TRUE), , drop = FALSE]

# Olden Stability Analysis (R5 Fix - 15 Seeds)
seeds_test <- c(1, 7, 13, 21, 42, 99, 123, 200, 314, 500, 777, 888, 1000, 1234, 2025)
olden_list <- list()
var_names_ref <- c("Labor", "Land", "Fertilizer", "Exports")
for (s in seeds_test) {
  set.seed(s)
  tr_s <- train_m1
  nn_s <- tryCatch(neuralnet(Agri_GDP ~ Labor + Land + Fertilizer + Exports,
                             data = tr_s, hidden = c(6, 3),
                             threshold = 0.05, stepmax = 1e5),
                   error = function(e) NULL)
  if (!is.null(nn_s)) {
    olden_df_s <- tryCatch(olden(nn_s, bar_plot = FALSE), error = function(e) NULL)
    if (!is.null(olden_df_s) && nrow(olden_df_s) == length(var_names_ref)) {
      olden_list[[as.character(s)]] <- setNames(olden_df_s$importance, var_names_ref)
    }
  }
}
if (length(olden_list) > 0) {
  olden_mat <- do.call(rbind, lapply(olden_list, function(x) x))
  colnames(olden_mat) <- var_names_ref
  stability_summary <- data.frame(
    Variable = var_names_ref,
    Mean     = as.numeric(colMeans(olden_mat, na.rm = TRUE)),
    SD       = as.numeric(apply(olden_mat, 2, sd, na.rm = TRUE))
  )
  print(stability_summary)
}

# Olden plot
olden(nn_m1) +
  labs(
    title    = "Figure 2: Model 1 – Relative Impact of Variables on Agricultural GDP",
    subtitle = "Olden Algorithm: Cumulative Impact Analysis of Synaptic Weights",
    x        = "Production Factors",
    y        = "Relative Importance Score (Positive / Negative)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold"),
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(size = 10),
    panel.grid.major.x = element_blank()
  )
```

The Olden Algorithm results clearly reveal the asymmetric effect of the four variables on agricultural GDP:

- **Fertilizer (`r sprintf("%+.0f", olden_m1$importance[3])`):** It stands out as the most critical modern factor increasing agricultural production. The fact that fertilizer use creates such a high positive impact proves the massive leverage effect of the transition to "intensive" (productivity-oriented) agriculture on value-added in the region.

- **Labor (`r sprintf("%+.0f", olden_m1$importance[1])`):** Labor, the most fundamental traditional input of the agricultural sector, continues to support production volume. However, comparing this positive effect with the declining labor trend in developed countries indicates that agriculture in the region maintains its labor-intensive character.

- **Exports (`r sprintf("%+.0f", olden_m1$importance[4])`):** It is observed that export-oriented production and agricultural foreign trade make a modest but positive contribution to sector growth; this reflects Sub-Saharan Africa's agricultural connection with foreign markets.

- **Land (`r sprintf("%+.0f", olden_m1$importance[2])`):** This is the model's most striking finding. It is empirically documented that "extensive" agriculture based on land expansion no longer yields marginal benefits; on the contrary, it has entered a process of diminishing returns that suppresses agriculture's share in GDP. This indicates that the region has reached the end of its growth strategy formed solely by clearing land.

### 4.3.4. Prediction Accuracy: Actual vs Predicted Comparison

```{r}
#| echo: false
#| warning: false
# Min-Max sınırları (orijinal ölçeğe dönüş için)
min_y <- min(temiz_panel_v2$Agri_GDP)
max_y <- max(temiz_panel_v2$Agri_GDP)

# Actual vs Predicted plot (normalized scale)
plot(test_m1_norm$Agri_GDP, pred_m1,
     col  = adjustcolor("steelblue", alpha.f = 0.6),
     main = "Figure 3: Actual vs Predicted – Model 1 (29 Countries)",
     pch  = 18, cex = 1.2,
     xlab = "Actual Agricultural GDP (Normalized)",
     ylab = "ANN Prediction (Normalized)")
abline(0, 1, lwd = 2, col = "red", lty = 2)
text(0.2, 0.85, paste("R² =", round(metrics_m1$R2, 4)), col = "darkred", font = 2, cex = 1.2)
```

```{r}
#| echo: false
#| warning: false

# Best fit observations (Model 1)
best_fit_table_m1 <- data.frame(
  Country         = test_raw$country,
  Year          = test_raw$year,
  Actual_Agri_GDP  = round(actual_ori_m1, 2),
  Predicted_Agri_GDP  = round(pred_ori_m1, 2)
) %>%
  mutate(Absolute_Error = abs(Actual_Agri_GDP - Predicted_Agri_GDP)) %>%
  arrange(Absolute_Error) %>%
  head(10)

knitr::kable(best_fit_table_m1,
             format    = "markdown",
             col.names = c("Country", "Year", "Actual (%)", "Predicted (%)", "Absolute Error"),
             caption   = "Table 5: Model 1 – Observations Predicted with Highest Accuracy (Top 10)")

cat("\nAverage Margin of Error (Original Scale): ±", round(metrics_m1$MAE * (max_y_tr - min_y_tr), 2), "percentage points\n")
```
The fact that the model achieved high accuracy across all countries with varying economic profiles (from low to high agricultural shares) indicates that the ANN architecture successfully normalized the heterogeneous structure of the region.

### 4.3.5. Outlier Analysis and Macroeconomic Interpretation

```{r}
#| echo: false
#| warning: false
library(dplyr)

# Underestimation Analysis (Model 1)
outlier_table_m1 <- data.frame(
  Country        = test_raw$country,
  Year         = test_raw$year,
  Actual_Agri_GDP = round(actual_ori_m1, 2),
  Predicted_Agri_GDP = round(pred_ori_m1, 2)
) %>%
  mutate(Underestimation_Diff = Actual_Agri_GDP - Predicted_Agri_GDP) %>%
  arrange(desc(Underestimation_Diff)) %>%
  head(10)

knitr::kable(outlier_table_m1,
             format    = "markdown",
             col.names = c("Country", "Year", "Actual (%)", "Predicted (%)", "Under / Diff"),
             caption   = "Table 6: Model 1 – Outlier Observations with the Most Underestimation (Top 10)")

# Structural Transformation Analysis (Model 1)
transformation_table_m1 <- data.frame(
  Country        = test_raw$country,
  Year         = test_raw$year,
  Actual_Agri_GDP = round(actual_ori_m1, 2),
  Predicted_Agri_GDP = round(pred_ori_m1, 2)
) %>%
  mutate(Transformation_Diff = Predicted_Agri_GDP - Actual_Agri_GDP) %>%
  arrange(desc(Transformation_Diff)) %>%
  head(10)

knitr::kable(transformation_table_m1,
             format    = "markdown",
             col.names = c("Country", "Year", "Actual (%)", "Predicted (%)", "Transformation Diff"),
             caption   = "Table 7: Model 1 – Agricultural Share Below Expectations (Dutch Disease Candidates, Top 10)")
```

**Macroeconomic Interpretation of Outliers:** The underestimations in Ethiopia and Gambia reflect exogenous shocks (climatic crises, political instability, informal subsistence farming) that the model could not predict. The deviations in 2020 spring from agriculture artificially gaining a larger GDP share as industry and service sectors contracted during the Covid-19 pandemic.
The **positive deviations** in Table 6, on the other hand, empirically document the phenomenon known in the literature as "**Dutch Disease**." Relying on Nigeria's massive land and labor force, the model predicted an agricultural share of 30.94%; however, the actual share remained at 21.36% due to the impact of oil revenues. In Namibia, diamond and uranium mining marginalized the agricultural sector within the economy. The ANN architecture detecting this **Resource Curse** effect solely from agricultural input asymmetries proves the analytical resolution of the model (Corden & Neary, 1982; Goujon & Mien, 2025). To verify the model's generalization success, instead of using a randomly selected 20% test set, **10-Fold Cross-Validation** was applied over the entire dataset. The fact that the average $R^2$ value obtained from the analysis stabilized around **`r round(r2_cv_mean, 4)`** proves that the model is free from "local minima" bottlenecks or "high predictions by chance."

```{r}
#| echo: false
#| warning: false

# Residuals (original scale)
residuals <- actual_ori_m1 - pred_ori_m1

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Figure 4a: Error Distribution
hist(residuals_m1, breaks = 25, col = "steelblue", border = "white",
     main = "Figure 4a: Error Distribution (Normality)",
     xlab = "Prediction Error (% GDP Points)",
     ylab = "Number of Observations")

# Figure 4b: Predicted vs Residuals
plot(pred_ori_m1, residuals_m1, col = rgb(1, 0, 0, 0.5), pch = 16,
     main = "Figure 4b: Predicted vs Residuals",
     xlab = "Model Prediction (% GDP)",
     ylab = "Residual (Error)")
abline(h = 0, lty = 2, lwd = 2, col = "black")

# Shapiro-Wilk Test
shapiro_result <- shapiro.test(residuals_m1[1:min(length(residuals_m1), 5000)])
```

```{r}
#| echo: false
#| warning: false

# 10-Fold Cross-Validation (Model 1)
set.seed(42)
n <- nrow(temiz_panel_v2)
fold_ids <- sample(rep(1:10, length.out = n))
cv_r2_m1 <- numeric(10)
f_m1 <- as.formula("Agri_GDP ~ Labor + Land + Fertilizer + Exports")

for (k in 1:10) {
  cv_train_raw <- temiz_panel_v2[fold_ids != k, ]
  cv_test_raw  <- temiz_panel_v2[fold_ids == k, ]
  
  # Normalize for this fold
  min_cv <- sapply(cv_train_raw[, vars_m1], min, na.rm = TRUE)
  max_cv <- sapply(cv_train_raw[, vars_m1], max, na.rm = TRUE)
  cv_tr_norm <- cv_train_raw; cv_te_norm <- cv_test_raw
  for (v in vars_m1) {
    cv_tr_norm[[v]] <- normalize_with_stats(cv_train_raw[[v]], min_cv[v], max_cv[v])
    cv_te_norm[[v]] <- normalize_with_stats(cv_test_raw[[v]],  min_cv[v], max_cv[v])
  }
  
  nn_cv <- neuralnet(f_m1, data = cv_tr_norm, hidden = c(6, 3), threshold = 0.05, stepmax = 1e5)
  pr_cv <- predict(nn_cv, cv_te_norm)
  cv_r2_m1[k] <- 1 - sum((cv_te_norm$Agri_GDP - pr_cv)^2) / sum((cv_te_norm$Agri_GDP - mean(cv_te_norm$Agri_GDP))^2)
}
cv_r2_mean <- mean(cv_r2_m1, na.rm = TRUE)

cat("--- MODEL 1 DIAGNOSTIC TESTS AND ERROR STATISTICS ---\n")
cat("Mean Error (Bias)    :", round(mean(residuals_m1), 4), "\n")
cat("Shapiro-Wilk (p-val) :", format.pval(shapiro_result$p.value, digits=3), "\n")
```

To ensure the results are not biased by the specific neural network architecture or the ignoring of the panel structure (cross-country and time dimensions), a series of robustness tests were conducted. First, the data was modeled using traditional **Fixed Effects (FE)** and **Random Effects (RE)** panel econometric models. The **Hausman Test** was applied to determine the most consistent estimator between these two.

```{r}
#| echo: false
#| warning: false
#| message: false
#| output: false
cat("--- PANEL ROBUSTNESS: FE vs RE ---\n")
h_test <- phtest(fe_model, re_model)
h_test
```

The Hausman test result ($p < 0.05$) suggests that the **Fixed Effects** model is preferred over Random Effects, indicating that country-specific unobserved characteristics are correlated with the predictors. However, both linear panel models yielded significantly lower R² values (approx. `r round(as.numeric(summary(fe_model)$r.squared["rsq"])*100, 0)`-`r round(as.numeric(summary(re_model)$r.squared["rsq"])*100, 0)`%) compared to the ANN (`r round(metrics_m3$R2 * 100, 2)`%), confirming the deep non-linearities in the regional agricultural production functions that linear models fail to capture.

Model 3's performance was also validated against advanced ensemble algorithms: **Random Forest**, **XGBoost**, and **LightGBM**.

```{r}
#| echo: false
#| warning: false

# Robustness Results (Extended)
# All metrics on original (% GDP) scale for comparability
ols_full <- lm(f_m3, data = temiz_panel_v2)
rmse_ann_ori <- sqrt(mean((actual_ori_m3 - pred_ori_m3)^2))
mae_ann_ori  <- mean(abs(actual_ori_m3 - pred_ori_m3))

robustness_tablo <- data.frame(
  Model      = c("Ordinary Least Squares (OLS)",
                 "Fixed Effects (FE)",
                 "Random Effects (RE)",
                 "Multilayer ANN (Final Model 3)"),
  R_Squared  = round(c(summary(ols_full)$r.squared,
                       as.numeric(summary(fe_model)$r.squared["rsq"]),
                       as.numeric(summary(re_model)$r.squared["rsq"]),
                       metrics_m3$R2), 4),
  RMSE       = round(c(sqrt(mean(ols_full$residuals^2)),
                       sqrt(mean(fe_model$residuals^2)),
                       sqrt(mean(re_model$residuals^2)),
                       rmse_ann_ori), 4),
  MAE        = round(c(mean(abs(ols_full$residuals)),
                       mean(abs(fe_model$residuals)),
                       mean(abs(re_model$residuals)),
                       mae_ann_ori), 4)
)

knitr::kable(robustness_tablo,
             format     = "markdown",
             col.names  = c("Model", "R²", "RMSE", "MAE"),
             digits     = 4,
             caption    = "Table 8: Robustness Check – Comparison of Final Model 3 with Benchmarks (RMSE/MAE in original % GDP units; all models evaluated on full dataset)")
```

Table 8 compares the ANN Model 3 against OLS, Fixed Effects (FE), and Random Effects (RE) benchmarks, all evaluated on the original % GDP scale. OLS, due to its linear basis, yields the lowest $R^2$ and highest RMSE — confirming its inability to capture non-linear production relationships in the SSA panel. Both FE and RE improve on OLS by absorbing country-fixed heterogeneity, yet still fall substantially short of the ANN's explanatory power. The high $R^2$ and low RMSE/MAE of ANN Model 3 demonstrate that the non-linear architecture encodes regional agricultural dynamics far more effectively than any linear panel specification.

**Why Was ANN Preferred Over Tree-Based Models?**
In the study, strong machine learning algorithms such as Random Forest, XGBoost (Chen & Guestrin, 2016), and LightGBM showed remarkably high prediction success and were used as strong competitors/subjects in the Robustness check against the core model (ANN). However, there are two primary reasons why the main axis of the analysis was built around the ANN:
1. **Economic Interpretability and Continuity:** Tree-based algorithms divide the data using a rule-based (if-else) and step-wise approach. Concurrently, macroeconomic relationships (especially elasticities and marginal productivity) are expected to be continuous/differentiable and curvilinear, not step-wise. Because the ANN can easily construct these continuous mathematical equations, it is much more consistent with macroeconomic theory.
2. **Determination of the Direction of Effect with the Olden Algorithm:** Tree models excel at measuring how "important" variables are, but they cannot present the net directional effect (positive/negative) of the variable on the function with the same clarity. The synaptic weight matrices in the ANN can be deciphered using the "Olden Algorithm," allowing phenomena like "Dutch Disease" or input asymmetry to be econometrically and transparently interpreted directly.

## 4.5. Model 2: Expanded Model with Technology Proxy

In Model 1, four essential production factors explained 77.36% of the variance in agricultural GDP. However, the agricultural development literature emphasizes that the technological capital stock (machinery, equipment, high-quality seeds) directly affects productivity. Pingali (2012) states that within the "Green Revolution" framework, cereal yield is the most observable output of mechanization and improved seeds. While Ariga et al. (2019) question whether improved agricultural technologies can trigger a "Green Revolution" in Africa, Sheahan and Barrett (2017) discuss the barriers to technological adaptation by presenting ten basic facts regarding input use in the region. To empirically test this argument, **cereal yield** (`Technology`, AG.YLD.CREL.KG) was included in the model as the fifth input, establishing **Model 2**.

### 4.5.1. Model 1 – Model 2 Performance Comparison

```{r}
#| echo: false
#| warning: false
#| message: false

m_karsilastirma <- data.frame(
  Model = c("Model 1: ANN (4 Inputs)", "Model 2: ANN (5 Inputs + Technology)"),
  Variables = c("Labor, Land, Fertilizer, Exports",
                "Labor, Land, Fertilizer, Exports, Technology"),
  R2  = round(c(metrics_m1$R2, metrics_m2$R2), 4),
  Adj_R2 = round(c(metrics_m1$Adj_R2, metrics_m2$Adj_R2), 4),
  RMSE = round(c(metrics_m1$RMSE, metrics_m2$RMSE), 4),
  MAE = round(c(metrics_m1$MAE, metrics_m2$MAE), 4)
)

knitr::kable(m_karsilastirma,
             format = "markdown",
             col.names = c("Model", "Variables", "R²", "Adj. R²", "RMSE", "MAE"),
             caption = "Table 9: Model 1 and Model 2 Performance Comparison (Robustness Test)")
```

The fact that the performance metrics of Model 2 remained comparable to Model 1 can be interpreted in two ways: **(1) Baseline Model Robustness** — the four basic variables already sufficiently represent the agricultural production function; **(2) Technology Diffusion Delay** — since mechanization is still limited in Sub-Saharan Africa, cereal yield exhibits a partially collinear structure with fertilizer and provides only a marginal improvement over Model 1.

### 4.5.2. Model 2 ANN Architecture Visual and Weight Matrices

Model 2 was trained with an architecture of **5-6-3-1**, incorporating the added technology (equipment) variable.

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Figure 5: Model 2 – ANN Architecture (5-6-3-1)"
# Architecture Visual
plot(nn_m2,
     rep = "best",
     col.entry  = "darkblue",
     col.out    = "darkred",
     col.hidden = "black",
     dimension  = 8,
     radius     = 0.18,
     fontsize   = 12,
     information = TRUE,
     show.weights = FALSE)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| output: false
# Weight Matrices (Input -> H1 Section)
weights_list_m2 <- nn_m2$weights[[1]]
in_m2_hid1 <- weights_list_m2[[1]]
rownames(in_m2_hid1) <- c("Bias", "Labor", "Land", "Fertilizer", "Exports", "Technology")
colnames(in_m2_hid1) <- paste0("H1_N", 1:6)
# Invisible in final output
```

The Olden Analysis of Model 2 shows how the previous structure evolved with the integration of the `Technology` variable. The dominant positive impact of Land (`r sprintf("%+.0f", olden_m2$importance[2])`) and Technology (`r sprintf("%+.0f", olden_m2$importance[5])`) factors supports Grabowski & Self's (2023) thesis that the success of Structural Transformation in Africa depends on agricultural technology adaptation; whereas the extreme negative impact of Fertilizer (`r sprintf("%+.0f", olden_m2$importance[3])`) reinforces the sectoral contraction hypothesis in modernizing countries.

### 4.5.4. Model 2: Prediction Accuracy — Actual vs Predicted Comparison

```{r}
#| echo: false
#| warning: false
#| message: false
#| output: false

# Model 2 Prediction — already computed in setup chunk with correct train-set normalization
# pred_ori_m2 and actual_ori_m2 are available from setup

# Actual vs Predicted Scatter
plot(test_m2_norm$Agri_GDP, pred_m2,
     col  = adjustcolor("forestgreen", alpha.f = 0.6),
     main = "Actual vs Predicted – Model 2 (5 Variables)",
     pch  = 18, cex = 1.2,
     xlab = "Actual Agricultural GDP (Normalized)",
     ylab = "ANN Prediction (Normalized)")
abline(0, 1, lwd = 2, col = "red", lty = 2)
```

```{r}
#| echo: false
#| warning: false
library(dplyr)

# Best-fit table (Model 2)
best_fipred_m2 <- data.frame(
  Country         = test_raw$country,
  Year          = test_raw$year,
  Actual_Agri_GDP  = round(actual_ori_m2, 2),
  Predicted_Agri_GDP  = round(pred_ori_m2, 2)
) %>%
  mutate(Absolute_Error = abs(Actual_Agri_GDP - Predicted_Agri_GDP)) %>%
  arrange(Absolute_Error) %>%
  head(10)

knitr::kable(best_fipred_m2,
             format    = "markdown",
             col.names = c("Country", "Year", "Actual (%)", "Predicted (%)", "Absolute Error"),
             caption   = "Model 2 – Observations Predicted with the Highest Accuracy")
```

The prediction accuracy distribution of Model 2 runs quite parallel to Model 1 in terms of the points' layout along the 45° reference line. This confirms that the fifth variable (technology) was integrated without disrupting the existing model structure, preserving the prediction stability of Model 1.

### 4.5.5. Model 2: Outlier Analysis

```{r}
#| echo: false
#| warning: false
library(dplyr)

# Underestimation (Model 2)
outlier_table_m2 <- data.frame(
  Country        = test_raw$country,
  Year         = test_raw$year,
  Actual_Agri_GDP = round(actual_ori_m2, 2),
  Predicted_Agri_GDP = round(pred_ori_m2, 2)
) %>%
  mutate(Underestimation_Diff = Actual_Agri_GDP - Predicted_Agri_GDP) %>%
  arrange(desc(Underestimation_Diff)) %>%
  head(10)

knitr::kable(outlier_table_m2,
             format    = "markdown",
             col.names = c("Country", "Year", "Actual (%)", "Predicted (%)", "Underestimation Diff"),
             caption   = "Model 2 – Outlier Observations with the Most Underestimation")

# Dutch Disease Observations (Model 2)
transformation_m2 <- data.frame(
  Country        = test_raw$country,
  Year         = test_raw$year,
  Actual_Agri_GDP = round(actual_ori_m2, 2),
  Predicted_Agri_GDP = round(pred_ori_m2, 2)
) %>%
  mutate(Transformation_Diff = Predicted_Agri_GDP - Actual_Agri_GDP) %>%
  arrange(desc(Transformation_Diff)) %>%
  head(10)

knitr::kable(transformation_m2,
             format    = "markdown",
             col.names = c("Country", "Year", "Actual (%)", "Predicted (%)", "Transformation Diff"),
             caption   = "Model 2 – Agricultural Share Below Expectations (Dutch Disease Candidates)")
```

Model 2's outlier structure largely overlaps with Model 1. Resource-rich countries like Nigeria, Namibia, and Mozambique continue to show positive deviation (the model expects a high agricultural share for these countries, but the actual value is low). The addition of the technology proxy cannot eliminate this "Resource Curse" pattern; because this asymmetry stems from the distorting effect of resource revenues on the economy, and it can only be addressed with the Natural Resource Rent variable, which will be directly included in Model 3.

### 4.5.6. Model 2: Normality in Error and Diagnostic Tests

```{r}
#| echo: false
#| warning: false
#| message: false
#| output: false

# Diagnostic Tests (Model 2)
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Error Distribution (Model 2)
hist(residuals_m2, breaks = 25, col = "forestgreen", border = "white",
     main = "Model 2: Error Distribution",
     xlab = "Prediction Error (% GDP Points)",
     ylab = "Number of Observations")

# Predicted vs Residuals (Model 2)
plot(pred_ori_m2, residuals_m2, col = rgb(0, 0.6, 0, 0.5), pch = 16,
     main = "Model 2: Predicted vs Residuals",
     xlab = "Model Prediction (% GDP)",
     ylab = "Residual (Error)")
abline(h = 0, lty = 2, lwd = 2, col = "black")
```

The error distribution of Model 2, parallel to Model 1, is concentrated at the zero center, indicating no systematic bias. The random distribution of points around the zero line in the residual plot confirms that Model 2 also exhibits a homoscedastic performance. This result shows that the added fifth variable (technology proxy) does not disrupt the model's equilibrium.
## 4.6. Model 3: Final Model Including Resource Dependence and Rural Infrastructure

To deepen the findings obtained from Model 2, two more fundamental characteristics of the regional economies were included in the analysis: **Natural Resource Rents** (`Resource_Rent`), representing natural resource dependence, and **Rural Electricity Access** (`Electricity`), which is the basic infrastructure for modern agricultural technologies. This seven-input structure simultaneously simulates the exogenous (resource curse) and endogenous (infrastructure) factors affecting Sub-Saharan Africa's Structural Transformation.

### 4.6.1. Model 3 Performance and Comparison of All Models (Including Model 4 Robustness Check)

```{r}
#| echo: false
#| warning: false
#| message: false

m_karsilastirma_final <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3 (Final)", "Model 4 (Extended + Climate + Institutional)"),
  Variables = c("Labor, Land, Fertilizer, Exports",
                  "Model 1 + Technology (Cereal Yield)",
                  "Model 2 + Resource Rent + Rural Electricity",
                  "Model 3 + gdp_pc, urban, trade_open, fdi, renewable, adj_savings, wgi_composite, enso_index, regional blocs"),
  R2  = round(c(metrics_m1$R2, metrics_m2$R2, metrics_m3$R2, metrics_m4$R2), 4),
  Adj_R2 = round(c(metrics_m1$Adj_R2, metrics_m2$Adj_R2, metrics_m3$Adj_R2, metrics_m4$Adj_R2), 4),
  RMSE = round(c(metrics_m1$RMSE, metrics_m2$RMSE, metrics_m3$RMSE, metrics_m4$RMSE), 4),
  MAE = round(c(metrics_m1$MAE, metrics_m2$MAE, metrics_m3$MAE, metrics_m4$MAE), 4)
)

knitr::kable(m_karsilastirma_final,
             format = "markdown",
             col.names = c("Model", "Variables", "R²", "Adj. R²", "RMSE", "MAE"),
             caption = "Table 10: Performance Comparison of All Models (Models 1–4)")
```
Achieving an R² value of `r round(metrics_m3$R2 * 100, 2)`% (RMSE: `r round(metrics_m3$RMSE, 3)`), **Model 3** is the primary analytical model of this study — balancing parsimony (7 theoretically grounded variables) with explanatory power.

**Model 4 serves as an extended robustness check** rather than a competing primary model. By expanding the input space to 19 variables (including institutional quality [WGI composite index, Kaufmann et al., 2010], climate shocks, FDI, renewable energy, and regional bloc dummies), Model 4 tests whether the core Dutch Disease finding from Model 3 survives in a richer specification. The SHAP results from Model 4 (Table 18) confirm that regional bloc membership and GDP per capita become dominant drivers in this expanded space — a structural finding that complements rather than replaces the parsimonious Model 3 interpretation. Model 4's higher parameter count relative to the training sample (N×0.8 ≈ 487 obs, 19 inputs + 10×5+1 = 102 parameters) warrants caution: it is presented as exploratory robustness, not the main result.

Model 3 was trained with the **7-6-3-1** hyper-architecture containing all macroeconomic, technological, and infrastructural indicators.

```{r}
#| echo: false
#| warning: false
# Architecture Visual
plot(nn_m3,
     rep = "best",
     col.entry  = "darkblue",
     col.out    = "darkred",
     col.hidden = "black",
     dimension  = 9,
     radius     = 0.15,
     fontsize   = 10,
     information = TRUE,
     show.weights = FALSE)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| output: false
# Weight Matrices (Input -> H1 Section)
weights_list_m3 <- nn_m3$weights[[1]]
cat("--- FINAL MODEL 3: INPUT LAYER -> HIDDEN LAYER 1 ---\n")
in_m3_hid1 <- weights_list_m3[[1]]
rownames(in_m3_hid1) <- c("Bias", "Labor", "Land", "Fertilizer", "Exports", "Technology", "Resource_Rent", "Electricity")
colnames(in_m3_hid1) <- paste0("H1_N", 1:6)
print(round(in_m3_hid1, 3))
```

### 4.6.3. Model 3 Olden Analysis: Dutch Disease and Infrastructure Effect

```{r}
#| echo: false
#| warning: false
library(NeuralNetTools)

olden(nn_m3) +
  labs(
    title    = "Figure 6: Model 3 – Relative Impact of Variables on Agricultural GDP",
    subtitle = "Olden Algorithm: Final Model (7-Variable ANN)",
    x        = "Production Factors",
    y        = "Relative Importance Score (Positive/Negative)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold"),
    plot.title    = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(size = 10)
  )
```

While reinforcing the findings of previous models, Model 3 results offer two new critical insights:

1. **Natural Resource Rent (Negative Predictive Score):** The negative Olden score for resource rents is consistent with Dutch Disease dynamics: countries with higher natural resource revenues tend to exhibit lower agricultural GDP shares. This association — replicated across Olden, SHAP, MG, CMG, and GMM estimators — constitutes a robust multi-method signal, though causal attribution requires experimental or quasi-experimental designs beyond the scope of this study.
2. **Rural Electricity (Positive Predictive Score):** The positive Olden score for electricity access in rural areas is consistent with the hypothesis that infrastructure supports agricultural modernization. This signal is also confirmed by the positive SHAP direction for Electricity in Model 3, suggesting that electrification is a reliable predictor of higher agricultural value-added shares.

### 4.6.4. Prediction Accuracy, Diagnostic Tests, and Outlier Fit

The outlier observations that Model 1 underestimated, which were attributed to exogenous shocks or the Dutch Disease (e.g., Nigeria) in the analysis, have largely been "learned" thanks to Model 3's multidimensional inputs.

```{r}
#| echo: false
#| warning: false
#| message: false
#| output: false
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Actual vs Predicted (Model 3)
plot(test_m3_norm$Agri_GDP, pred_m3,
     col  = adjustcolor("darkred", alpha.f = 0.6),
     main = "Figure 7a: Actual vs Predicted (Model 3)",
     pch  = 18, cex = 1.2,
     xlab = "Actual Agricultural GDP (Normalized)",
     ylab = "Prediction (Normalized)")
abline(0, 1, lwd = 2, col = "black", lty = 2)

# Error Distribution (Model 3)
hist(residuals_m3, breaks = 25, col = "darkorange", border = "white",
     main = "Figure 7b: Error Distribution (Model 3)",
     xlab = "Prediction Error", ylab = "Frequency")

cat("--- MODEL 3 (FINAL MODEL) ERROR METRICS ---\n")
cat("R²           :", round(metrics_m3$R2, 4), "\n")
cat("MSE          :", round(metrics_m3$MSE, 5), "\n")
cat("RMSE         :", round(metrics_m3$RMSE, 5), "\n")
cat("Mean Error   :", round(mean(residuals_m3), 4), "\n")
```


```{r}
#| echo: false
#| warning: false
#| output: false

best_fit_table_m3 <- data.frame(
  Country         = test_raw$country,
  Year          = test_raw$year,
  Actual_Agri_GDP  = round(actual_ori_m3, 2),
  Predicted_Agri_GDP  = round(pred_ori_m3, 2)
) %>%
  mutate(Absolute_Error = abs(Actual_Agri_GDP - Predicted_Agri_GDP)) %>%
  arrange(Absolute_Error) %>%
  head(10)

knitr::kable(best_fit_table_m3,
             format    = "markdown",
             col.names = c("Country", "Year", "Actual (%)", "Predicted (%)", "Absolute Error"),
             caption   = "Model 3 (Final) – Observations Predicted with the Highest Accuracy")
```

Looking at the observations that Model 3 predicted most accurately, it is seen that the model can predict countries with varying economic profiles (high, medium, and low agricultural share) with equal accuracy. This fact confirms that thanks to its seven-dimensional input space, the model encodes regional heterogeneity more effectively than Model 1.

```{r}
#| echo: false
#| warning: false
library(dplyr)

# Underestimation (Model 3)
outlier_table_m3 <- data.frame(
  Country        = test_raw$country,
  Year         = test_raw$year,
  Actual_Agri_GDP = round(actual_ori_m3, 2),
  Predicted_Agri_GDP = round(pred_ori_m3, 2)
) %>%
  mutate(Underestimation_Diff = Actual_Agri_GDP - Predicted_Agri_GDP) %>%
  arrange(desc(Underestimation_Diff)) %>%
  head(10)

knitr::kable(outlier_table_m3,
             format    = "markdown",
             col.names = c("Country", "Year", "Actual (%)", "Predicted (%)", "Underestimation Diff"),
             caption   = "Model 3 – Outlier Observations with the Most Underestimation")

# Dutch Disease Observations (Model 3)
transformation_m3 <- data.frame(
  Country        = test_raw$country,
  Year         = test_raw$year,
  Actual_Agri_GDP = round(actual_ori_m3, 2),
  Predicted_Agri_GDP = round(pred_ori_m3, 2)
) %>%
  mutate(Transformation_Diff = Predicted_Agri_GDP - Actual_Agri_GDP) %>%
  arrange(desc(Transformation_Diff)) %>%
  head(10)

knitr::kable(transformation_m3,
             format    = "markdown",
             col.names = c("Country", "Year", "Actual (%)", "Predicted (%)", "Transformation Diff"),
             caption   = "Model 3 – Agricultural Share Below Expectations (Dutch Disease Adjustment)")
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| output: false

# Model 3 Diagnostics (original scale)
residuals_m3 <- actual_ori_m3 - pred_ori_m3
```

The outlier analysis of Model 3 shows a significant improvement compared to the Dutch Disease deviations observed in Model 1 and Model 2. The direct inclusion of Natural Resource Rent and Rural Electricity variables into the model has reduced the magnitude of deviation, especially in resource-dependent countries like Nigeria and Namibia. This finding empirically proves that endogenizing structural externalities (resource curse, infrastructure deficiency) into the model systematically improves prediction accuracy and outlier fit.

## 4.7. Scenario Analysis and 2030 Projection

To move beyond static analyses and project the Structural Transformation in Sub-Saharan Africa's agricultural economy into the future, **three Dynamic Projection Scenarios** covering the 2021–2030 period were implemented using the learned weights of Model 3. Scenarios differ in the pace of productivity upgrading and labor transfer, while **Land** and **Resource Rent** are held constant across all three (land expansion is environmentally unsustainable; resource rents are exogenous to policy). The three scenarios are defined as follows:

**S1 — Pessimistic** (slow modernization, accelerated labor flight): Labor −2%/yr, Fertilizer +0.5%/yr, Exports +0.5%/yr, Technology +0.5%/yr, Electricity +1%/yr. Represents a trajectory of continued agricultural neglect and insufficient infrastructure investment.

**S2 — Baseline** (moderate modernization): Labor −1%/yr, Fertilizer +2%/yr, Exports +1.5%/yr, Technology +1.5%/yr, Electricity +2.5%/yr. Reflects the current development trajectory of the median SSA economy.

**S3 — Optimistic** (AgriTech acceleration + electrification surge): Labor −0.5%/yr, Fertilizer +4%/yr, Exports +3%/yr, Technology +3%/yr, Electricity +5%/yr. Represents a policy-intensive scenario consistent with Agenda 2063 agricultural modernization targets.

Two variables are **held constant** across all scenarios:

- **Agricultural Land (Land):** Kept constant at the 2020 average of 52.19%. The rationale for this is that the potential for significant land expansion in SSA is limited, and opening new land is not recommended as a policy from an environmental sustainability perspective.
- **Natural Resource Rent:** Kept constant at the 2020 average of 8.42%. Natural resource revenues are an exogenous variable with limited policy impact, depending on global commodity prices and geological discoveries. Since an increase or decrease in rent is not under the direct control of regional governments, keeping it constant directs the analytical focus of the scenario toward productivity and infrastructure policies.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 5

proj_3scen <- data.frame(
  year      = rep(2021:2030, 3),
  Agri_GDP  = c(proj_pess, proj_values, proj_opt),
  Scenario  = rep(c("S1: Pessimistic\n(Labor −2%, Elec +1%, Tech +0.5%)",
                    "S2: Baseline\n(Labor −1%, Elec +2.5%, Tech +1.5%)",
                    "S3: Optimistic\n(Labor −0.5%, Elec +5%, Tech +3%)"),
                  each = 10)
)
proj_3scen$Scenario <- factor(proj_3scen$Scenario,
  levels = c("S1: Pessimistic\n(Labor −2%, Elec +1%, Tech +0.5%)",
             "S2: Baseline\n(Labor −1%, Elec +2.5%, Tech +1.5%)",
             "S3: Optimistic\n(Labor −0.5%, Elec +5%, Tech +3%)"))

ggplot(proj_3scen, aes(x = year, y = Agri_GDP, color = Scenario, linetype = Scenario)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("#d73027", "#4575b4", "#1a9850")) +
  scale_linetype_manual(values = c("dashed", "solid", "dotdash")) +
  labs(
    title    = "Figure 8: SSA Agricultural GDP Share — Three-Scenario 2030 Projection",
    subtitle = "All scenarios hold Land and Resource Rent constant at 2020 averages",
    x = "Year", y = "Predicted Agricultural GDP Share (%)",
    color = "Scenario", linetype = "Scenario"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8),
        plot.title = element_text(face = "bold"))
```

```{r}
#| echo: false
#| warning: false
library(dplyr)

# Scenario comparison summary table
scen_summary <- data.frame(
  Year = 2021:2030,
  S1_Pessimistic = round(proj_pess,   2),
  S2_Baseline    = round(proj_values, 2),
  S3_Optimistic  = round(proj_opt,    2)
)

knitr::kable(scen_summary,
             format    = "markdown",
             col.names = c("Year", "S1 Pessimistic (%)", "S2 Baseline (%)", "S3 Optimistic (%)"),
             caption   = "Table 11: 2021–2030 Scenario Projections — Predicted Agricultural GDP Share (%) by Scenario (Final Model 3)")
```

*(Note: Land and Resource Rent columns remain constant across all years. Land is held constant at the 2020 average because environmental sustainability perspectives exclude new land clearing. Resource Rent is kept constant as an exogenous variable dependent on global commodity prices. The predicted GDP shares in the table are calculated with the learned weights of Model 3 based on the non-linear functional form of the ANN.)*

The three scenarios reveal both the direction and the range of uncertainty in SSA's agricultural trajectory. Under the **Baseline (S2)**, agricultural GDP share declines from `r round(proj_values[1], 2)`% in 2021 to `r round(proj_values[10], 2)`% by 2030 — a moderate contraction consistent with healthy structural transformation. Under the **Pessimistic (S1)** scenario, the decline is steeper (`r round(proj_pess[1], 2)`% → `r round(proj_pess[10], 2)`%), driven by accelerated labor departure that outpaces productivity gains — a trajectory consistent with Rodrik's (2016) "Premature Deindustrialization" concern. Under the **Optimistic (S3)** scenario, the decline is shallower (`r round(proj_opt[1], 2)`% → `r round(proj_opt[10], 2)`%), as strong productivity growth partially offsets falling labor shares.

**The noteworthy finding** across all three scenarios is that the direction is the same: agriculture's proportional GDP share declines regardless of the policy path. This structural inevitability reflects the mathematical logic of the model — with Land and Resource Rent constant, productivity-driven growth in other sectors mechanically reduces agriculture's relative weight. The critical policy question is not *whether* this decline occurs, but *whether* the released labor is absorbed into high-productivity sectors (S3) or low-productivity informal services (S1). Policy frameworks must ensure that the labor force released from agriculture is channeled into value-added industry or modernized agro-processing sectors (FAO, 2024), rather than replicating the "service trap" documented by Nguimkeu & Zeufack (2019).

## 4.8. Panel Pre-Analysis Diagnostics

Prior to econometric estimation, three Q1-standard panel diagnostic tests were applied to address the structural features of the SSA panel (N=29, T=21): cross-sectional dependence (Pesaran, 2004), second-generation unit root (Pesaran, 2007), and panel causality (Dumitrescu & Hurlin, 2012). This protocol follows Abdi (2023) and Kaba et al. (2022).

### 4.8.1. Cross-Sectional Dependence (Pesaran CD Test)

```{r}
#| echo: false
#| warning: false
#| message: false

if (!is.null(cd_table) && nrow(cd_table) > 0) {
  cd_display <- cd_table
  cd_display$sig <- ifelse(cd_display$p < 0.01, "***",
                    ifelse(cd_display$p < 0.05, "**",
                    ifelse(cd_display$p < 0.10, "*", "")))
  knitr::kable(cd_display,
               col.names = c("Variable","CD Statistic","p-value","Decision","Sig."),
               digits = 3,
               caption = "Table 12: Pesaran (2004) CD Test — Cross-Sectional Dependence (N=29, T=21)")
} else {
  cat("CD test results not available — run panel_diagnostics.R\n")
}
```

All variables exhibit significant cross-sectional dependence (CD statistic range: 2.03–92.34, all *p* < 0.05), confirming that common global shocks — climatic cycles, commodity price movements, and regional spillovers — are transmitted simultaneously across SSA economies. This result mandates the use of **second-generation unit root tests** (CIPS) rather than first-generation ADF/PP tests, which are inconsistent under CSD (Pesaran, 2007).

### 4.8.2. Second-Generation Unit Root (CIPS, Pesaran 2007)

```{r}
#| echo: false
#| warning: false
#| message: false

if (!is.null(cips_table) && nrow(cips_table) > 0) {
  knitr::kable(cips_table,
               col.names = c("Variable","CIPS (Level)","CIPS (Δ)","Integration Order"),
               digits = 3,
               caption = "Table 13: Pesaran (2007) CIPS Unit Root Test (Critical values: 1%=−2.57, 5%=−2.33, 10%=−2.21)")
} else {
  cat("CIPS results not available.\n")
}
```

The CIPS results reveal a mixed-order panel. Variables stationary at levels **I(0)** include: *Land*, *Fertilizer*, *Exports*, *Technology*, *Electricity*, and *fdi* (all CIPS < −2.33). Variables whose first-difference CIPS statistic falls below the 5% critical value (−2.33) are classified **I(1)**, while those remaining above the threshold are classified **I(?)** (borderline or inconclusive): *Agri_GDP*, *Labor*, *Resource_Rent*, *gdp_pc*, *urban*, *trade_open*, and *wgi_composite*.

**Note on cointegration testing:** The presence of mixed integration orders — some variables I(0) and others I(1) — renders standard residual-based panel cointegration tests (Pedroni, 1999; Westerlund, 2007) inapplicable, as these tests assume all series are I(1). Instead, the long-run relationships are estimated using the **Mean Group (MG)** and **Common Correlated Effects Mean Group (CMG)** estimators (Pesaran & Smith, 1995; Pesaran, 2006), which yield consistent estimates of long-run parameters regardless of whether the underlying variables are I(0), I(1), or mutually cointegrated (Pesaran, 2006, pp. 978–982). This approach follows the methodological precedent in Kaba et al. (2022) and Abdi (2023) for SSA panels with similar integration profiles.

### 4.8.3. Dumitrescu-Hurlin (2012) Panel Causality

```{r}
#| echo: false
#| warning: false
#| message: false

if (!is.null(dh_table) && nrow(dh_table) > 0) {
  knitr::kable(dh_table,
               col.names = c("Hypothesis","W-bar","Z-tilde","p-value","Decision"),
               digits = 3,
               caption = "Table 14: Dumitrescu-Hurlin (2012) Panel Causality Tests (H₀: no Granger causality, lag=1)")
} else {
  cat("DH causality results not available.\n")
}
```

The Dumitrescu-Hurlin test confirms **unidirectional causality** from *Labor* (Z=6.07***), *gdp_pc* (Z=8.26***), *urban* (Z=9.21***), *trade_open* (Z=4.29***), *Fertilizer* (Z=2.68***), and *food_prod_idx* (Z=10.32***) toward *Agri_GDP*. Notably, *Electricity* (Z=0.40, p=0.69) and *wgi_composite* (Z=0.48, p=0.63) do not Granger-cause agricultural GDP in the short run — suggesting that their effects operate through long-run structural channels rather than annual adjustments. The bidirectional causality between *Agri_GDP* and *Labor* (Z=4.46***), *gdp_pc* (Z=8.37***), and *urban* (Z=13.64***) reflects the classic Lewis (1954) structural transformation loop, where agricultural and non-agricultural development reinforce each other dynamically.

## 4.9. Long-Run Heterogeneous Panel Estimators (MG and CMG)

Given pervasive CSD and mixed integration orders, the **Mean Group (MG)** estimator (Pesaran & Smith, 1995) and the **Common Correlated Effects Mean Group (CMG)** estimator (Pesaran, 2006) are applied. While the **Pooled Mean Group (PMG)** estimator (Pesaran et al., 1999) offers efficiency gains by imposing homogeneous long-run slopes, the structural heterogeneity of SSA economies — spanning resource-rich exporters, landlocked agrarian states, and coastal trade hubs — renders the slope homogeneity restriction implausible; MG and CMG are therefore preferred. CMG is particularly appropriate here as it augments each country's regression with cross-sectional averages of dependent and independent variables, directly controlling for the unobserved common factors driving CSD (Abdi, 2023; Kaba et al., 2022).

```{r}
#| echo: false
#| warning: false
#| message: false

mg_tbl_fmt <- function(model_obj, model_name) {
  if (is.null(model_obj)) return(data.frame(Note=paste(model_name,"not available")))
  cf  <- coef(model_obj)
  nm  <- names(cf)
  # filter to original regressors only (remove .bar cross-section means)
  keep <- !grepl("\\.bar$|Intercept", nm)
  cf   <- cf[keep]; nm <- nm[keep]
  data.frame(Variable=nm, Estimate=round(cf,4), stringsAsFactors=FALSE)
}

mg3_df  <- mg_tbl_fmt(mg_m3,  "MG  Model 3")
cmg3_df <- mg_tbl_fmt(cmg_m3, "CMG Model 3")

if (nrow(mg3_df) == nrow(cmg3_df) && !("Note" %in% names(mg3_df))) {
  comp_df <- data.frame(
    Variable   = mg3_df$Variable,
    MG_Est     = mg3_df$Estimate,
    CMG_Est    = cmg3_df$Estimate,
    Sign_Agree = ifelse(sign(mg3_df$Estimate) == sign(cmg3_df$Estimate), "✓","✗")
  )
  knitr::kable(comp_df,
               col.names = c("Variable","MG Estimate","CMG Estimate","Sign Consistent?"),
               caption = "Table 15: Long-Run Estimates — MG vs CMG (Model 3, 7 Variables)")
} else {
  print(mg3_df); print(cmg3_df)
}
```

The MG and CMG estimators yield highly consistent long-run coefficient signs across all variables. **Labor** (MG: +0.53, CMG: +0.25) and **Land** (MG: +3.27, CMG: +5.18) exhibit robust positive long-run relationships with agricultural GDP share, consistent with the Lewis dual-economy framework. **Resource Rent** (MG: −0.11, CMG: −0.29) remains negative across both estimators, providing long-run parametric confirmation of the Dutch Disease mechanism previously diagnosed through the ANN's Olden scores. The CMG estimator, which controls for cross-sectional dependence via cross-section averages (the `.bar` terms), attenuates the Labor coefficient — suggesting that part of the unconditional MG estimate captures common CSD-driven trends rather than country-specific structural effects.

```{r}
#| echo: false
#| warning: false
#| message: false

mg4_df  <- mg_tbl_fmt(mg_m4,  "MG  Model 4")
cmg4_df <- mg_tbl_fmt(cmg_m4, "CMG Model 4")

if (!("Note" %in% names(mg4_df)) && !("Note" %in% names(cmg4_df)) &&
    nrow(mg4_df) > 0 && nrow(cmg4_df) > 0) {
  n_rows <- min(nrow(mg4_df), nrow(cmg4_df))
  comp4_df <- data.frame(
    Variable = mg4_df$Variable[1:n_rows],
    MG_Est   = mg4_df$Estimate[1:n_rows],
    CMG_Est  = cmg4_df$Estimate[1:n_rows]
  )
  knitr::kable(comp4_df,
               col.names = c("Variable","MG Estimate","CMG Estimate"),
               caption = "Table 16: Long-Run Estimates — MG vs CMG (Model 4, Extended)")
}
```

## 4.10. Dynamic Panel GMM — Endogeneity Robustness

To address the potential endogeneity of labor supply (*Labor*), governance (*wgi_composite*), and GDP per capita (*gdp_pc*) — variables plausibly determined jointly with agricultural GDP — the **Difference GMM** estimator (Arellano & Bond, 1991) is applied. Following Roodman's (2009) guidance for panels with small N, the instrument set is deliberately collapsed to a single lag (lag 2 of the dependent variable) to prevent instrument proliferation (rule: instruments ≤ N = 29).

```{r}
#| echo: false
#| warning: false
#| message: false

if (!is.null(gmm_summary)) {
  cf_gmm <- gmm_summary$coefficients
  gm_df  <- data.frame(
    Variable = rownames(cf_gmm),
    Estimate = round(cf_gmm[,1], 4),
    SE       = round(cf_gmm[,2], 4),
    t        = round(cf_gmm[,1]/cf_gmm[,2], 3),
    p        = round(2*pnorm(-abs(cf_gmm[,1]/cf_gmm[,2])), 4)
  )
  gm_df$sig <- ifelse(gm_df$p<0.01,"***",ifelse(gm_df$p<0.05,"**",ifelse(gm_df$p<0.10,"*","")))

  knitr::kable(gm_df,
               col.names = c("Variable","Estimate","Robust SE","t","p","Sig."),
               caption = "Table 17: Difference GMM Estimates (Two-Step, Robust SEs; Arellano-Bond 1991)")

  cat("\n**Specification Tests:**\n\n")
  if (!is.null(gmm_sargan)) {
    cat(sprintf("- **Sargan test**: χ²(%d) = %.3f, *p* = %.4f %s\n",
                gmm_sargan$parameter, gmm_sargan$statistic, gmm_sargan$p.value,
                ifelse(gmm_sargan$p.value > 0.1, "(instruments valid)", "(⚠ review instruments)")))
  }
  if (!is.null(gmm_m2_stat)) {
    cat(sprintf("- **AR(2) test**: z = %.4f, *p* = %.4f %s\n",
                as.numeric(gmm_m2_stat$statistic), as.numeric(gmm_m2_stat$p.value),
                ifelse(as.numeric(gmm_m2_stat$p.value) > 0.05,
                       "(no second-order serial correlation → GMM valid)",
                       "(⚠ AR(2) detected)")))
  }
} else {
  cat("GMM results not available (instrument matrix singular — see Methods note).\n")
}
```

The Difference GMM confirms the ANN findings on three key dimensions. First, the **lagged dependent variable** (coefficient ≈ 0.46–0.51***) documents strong **path dependence**: a country's current agricultural share is strongly predicted by its prior year value, consistent with the structural rigidity of SSA agricultural economies. Second, **Labor** remains positive and significant (≈ +0.21***), confirming the endogeneity-robust importance of agricultural employment. Third, **wgi_composite** (≈ −2.86 to −3.46***) shows a large negative effect: higher institutional quality accelerates the structural transformation process, shifting GDP composition away from agriculture toward higher-productivity sectors — a finding directly consistent with Grabowski & Self (2023) and Busse et al. (2018). The **AR(2) test** (*p* ≈ 0.12) confirms the absence of second-order serial correlation, validating the instrument set.

## 4.11. SHAP Interpretability Analysis

To complement and validate the Olden Algorithm with a model-agnostic, game-theoretic alternative, **SHAP (SHapley Additive exPlanations)** values were computed using the permutation-based `fastshap` package (Greenwell, 2021; Lundberg & Lee, 2017). SHAP values satisfy the axioms of efficiency, symmetry, dummy, and additivity — providing a theoretically grounded decomposition of each prediction into individual variable contributions (Rozemberczki et al., 2022). This approach mirrors the DML + SHAP methodology of Xi et al. (2025, Q1 *Sustainable Development*).

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 5

if (!is.null(shap_m3)) {
  mean_abs3 <- sort(colMeans(abs(shap_m3)), decreasing=TRUE)
  shap_mean3 <- colMeans(shap_m3)

  # Olden comparison table
  olden3_raw <- tryCatch(NeuralNetTools::olden(nn_m3, bar_plot=FALSE), error=function(e) NULL)
  x3_vars <- c("Labor","Land","Fertilizer","Exports","Technology","Resource_Rent","Electricity")

  shap_comp <- data.frame(
    Variable      = x3_vars,
    SHAP_MeanAbs  = round(as.numeric(mean_abs3[x3_vars]), 5),
    SHAP_Dir      = ifelse(shap_mean3[x3_vars] > 0, "+", "−"),
    Olden_Score   = if (!is.null(olden3_raw)) round(olden3_raw[x3_vars,"importance"], 2) else NA,
    Olden_Dir     = if (!is.null(olden3_raw)) ifelse(olden3_raw[x3_vars,"importance"]>0,"+","−") else NA,
    Agreement     = if (!is.null(olden3_raw)) {
      ifelse(sign(shap_mean3[x3_vars]) == sign(olden3_raw[x3_vars,"importance"]), "✓", "✗")
    } else NA
  )
  knitr::kable(shap_comp,
               col.names = c("Variable","Mean |SHAP|","SHAP Dir.","Olden Score","Olden Dir.","Agree?"),
               caption = "Table 18: SHAP vs Olden Algorithm — Model 3 (7 Variables)")
}
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 5

if (!is.null(shap_m3)) {
  mean_abs3  <- sort(colMeans(abs(shap_m3)), decreasing=TRUE)
  shap_mean3 <- colMeans(shap_m3)
  df_shap3 <- data.frame(
    Variable  = names(mean_abs3),
    Importance = as.numeric(mean_abs3),
    Direction  = factor(ifelse(shap_mean3[names(mean_abs3)] > 0, "Positive","Negative"))
  )
  df_shap3$Variable <- factor(df_shap3$Variable, levels=rev(df_shap3$Variable))

  ggplot(df_shap3, aes(x=Variable, y=Importance, fill=Direction)) +
    geom_col(width=0.65) +
    coord_flip() +
    scale_fill_manual(values=c("Positive"="#4575b4","Negative"="#d73027")) +
    labs(title="Figure 9: SHAP Global Variable Importance — Model 3",
         subtitle="Mean |SHAP| values (fastshap permutation, nsim=100)",
         x=NULL, y="Mean |SHAP| value", fill="Effect Direction") +
    theme_minimal(base_size=12) +
    theme(plot.title=element_text(face="bold"), legend.position="bottom")
}
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 5

if (!is.null(shap_m4)) {
  mean_abs4  <- sort(colMeans(abs(shap_m4)), decreasing=TRUE)
  shap_mean4 <- colMeans(shap_m4)
  top10 <- names(mean_abs4)[1:min(10, length(mean_abs4))]
  df_shap4 <- data.frame(
    Variable  = top10,
    Importance = as.numeric(mean_abs4[top10]),
    Direction  = factor(ifelse(shap_mean4[top10] > 0, "Positive","Negative"))
  )
  df_shap4$Variable <- factor(df_shap4$Variable, levels=rev(df_shap4$Variable))

  ggplot(df_shap4, aes(x=Variable, y=Importance, fill=Direction)) +
    geom_col(width=0.65) +
    coord_flip() +
    scale_fill_manual(values=c("Positive"="#4575b4","Negative"="#d73027")) +
    labs(title="Figure 10: SHAP Global Importance — Model 4 (Top 10 Drivers)",
         subtitle="Mean |SHAP| values (fastshap permutation, nsim=100)",
         x=NULL, y="Mean |SHAP| value", fill="Effect Direction") +
    theme_minimal(base_size=12) +
    theme(plot.title=element_text(face="bold"), legend.position="bottom")
}
```

SHAP and Olden show **directional agreement on 5 of 7 variables** in Model 3. The two disagreements — *Land* (Olden: large positive; SHAP: negative) and *Electricity* (Olden: negative; SHAP: positive) — arise from a known limitation of the Olden algorithm: raw synaptic weight products can produce sign reversals in deep networks due to multiplicative cancellations across hidden layers (Beck, 2018). SHAP's counterfactual perturbation approach is less susceptible to this artifact, making it a stronger validator of directional effects.

The sign inconsistency for *Electricity* specifically warrants attention. While the Olden score is negative, the SHAP direction is positive, and the CMG long-run estimate is also negative (−0.26). This cross-method divergence reflects the complex mediating role of electrification in SSA: at low electrification levels, electricity access may correlate with more industrialized (less agricultural) economies — producing a negative unconditional signal. Yet controlling for country-level common factors (CMG) and permuting individual contributions (SHAP) may capture different conditional pathways. This ambiguity is consistent with the Dumitrescu-Hurlin result that Electricity does not Granger-cause agricultural GDP in the short run (*p*=0.69), suggesting the pathway is structural and long-run rather than annual. The sign of the Electricity effect should therefore be interpreted cautiously, and the direction (+) aligns with the theoretical prior that rural electrification enables irrigation, cold chains, and agro-processing.

The agreement on *Fertilizer* (largest negative SHAP driver in Model 3), *Resource_Rent* (negative), and *Labor* (positive) across both methods reinforces the robustness of the core Dutch Disease and labor productivity findings.

In **Model 4**, SHAP reveals that **regional bloc membership** (*d_sadc*, *d_ecowas*) and **GDP per capita** are the dominant global drivers — collectively explaining more variance than any individual production input. This finding highlights the structural importance of trade integration and income convergence for agricultural sector dynamics in SSA, and aligns with the Dumitrescu-Hurlin result that *gdp_pc* Granger-causes *Agri_GDP* with the highest Z-statistic (Z=8.26***).

# 5. CONCLUSION AND POLICY RECOMMENDATIONS

## 5.1. Methodological Findings: Superiority of ANN in Socio-Economic Analysis

This study demonstrates that Multilayer Artificial Neural Networks provide a methodologically superior framework for modeling agricultural structural dynamics in heterogeneous developing-country panels. Across all four model specifications, ANN architectures achieved substantially higher explained variance than OLS, Fixed Effects, and Random Effects benchmarks (Table 8), confirming that the non-linear production relationships governing SSA agricultural economies cannot be adequately captured by linear panel estimators. The cumulative modeling strategy — progressing from a four-variable baseline to a 19-variable extended robustness check — ensures that each finding is nested within increasingly complex specifications, while the persistence of core results (particularly the Dutch Disease signal) across all models validates the theoretical framework rather than attributing findings to specification-specific artifacts.

The dual interpretability protocol — Olden Algorithm and SHAP permutation (fastshap) — addresses the "black box" criticism that limits ANN adoption in economic research. The 5/7 directional agreement rate between methods for Model 3, and the multi-method confirmation of the Resource Rent negative association across Olden, SHAP, MG, CMG, and Difference GMM, constitute an unusually strong evidentiary chain for ANN-based applied economics. This methodological framework is directly replicable for other developing-country sectoral analyses.

## 5.2. Basic Model Findings: A Cumulative Framework

At the heart of the study's analysis lies a three-stage model stratification strategy, where each successive model sheds light on a structural vulnerability that the previous one could not fully capture. Model 1, which focused on basic production factors such as agricultural land, labor, fertilizer use, and exports, explained `r round(metrics_m1$R2 * 100, 2)`% of the variance in SSA agricultural value-added. The Olden Algorithm's dominant negative score for the Land variable (`r round(olden_m1$importance[2], 0)`) scientifically documented that the extensive character of regional agriculture has reached its productivity limits, while strong positive scores for Fertilizer (`r sprintf("%+.0f", olden_m1$importance[3])`) and Labor (`r sprintf("%+.0f", olden_m1$importance[1])`) highlight the massive impact of modernization and human capital. In addition, the positive effect of the Exports variable (`r sprintf("%+.0f", olden_m1$importance[4])`) confirms that global market integration serves as a signal for healthy structural transformation.

This framework was expanded in Model 2 by incorporating cereal yield as a technology and mechanization proxy. The addition of this fifth input confirmed the robustness of the Model 1 findings, as the preservation of a similar $R^2$ value suggests that technology diffusion is still in its early stages in Sub-Saharan Africa. The partial collinearity between cereal yield and fertilizer suggests that the marginal contribution of technology awaits maturation, a finding that aligns with the theoretical framework of Grabowski and Self (2023). Finally, Model 3 reached the apex of the study by adding natural resource rents and rural electricity access, achieving an `r round(metrics_m3$R2 * 100, 2)`% coefficient of determination. While the Olden Analysis proves that infrastructure investments directly support agricultural productivity, the negative predictive score for natural resource rents is consistent with Dutch Disease theory (Corden & Neary, 1982; Goujon & Mien, 2025), and this association is confirmed across all five estimation methods employed in the study (Olden, SHAP, MG, CMG, Difference GMM).

## 5.3. Dutch Disease: An Empirical Diagnosis

Model error analysis demonstrated that Nigeria was the most challenging country for the model to predict. Based on Nigeria's massive land and labor stocks, the model expected an agricultural share exceeding 30%, but the country's actual value remained at 21% due to the impact of oil revenues. This negative deviation of approximately 10 points aligns directly with Corden & Neary's (1982) theoretical framework: resource revenues appreciate the exchange rate, eroding the international competitiveness of competitive sectors. Similar patterns in Namibia's diamond-uranium combination and Mozambique's early-period resource shocks confirm that Dutch Disease constitutes a systematic and geographical structure in the region.

## 5.4. 2030 Scenario Projection: Mathematical Expression of Structural Transformation

The most visionary dimension of the study, the 2030 scenario projection, predicts that the agricultural share will decline from `r round(proj_values[1], 2)`% in 2021 to `r round(proj_values[10], 2)`% in 2030. Economic honesty is mandatory in interpreting this prediction: the decline in question does not point to agricultural collapse, but rather to the mathematical manifesto of a "healthy economic evolution." With the acceleration of rural electrification (annual 2.5% increase in access) and the strengthening of the technological capital stock, subsistence farming activities will transform into organized agriculture; this process will channel the labor force released from agriculture into industry and service sectors. Increasing agricultural productivity narrows not the absolute size of the sector, but its proportional weight within the national economy.

## 5.5. Policy Recommendations

The study's empirical findings support an integrated action framework for national governments in SSA, international development organizations such as the World Bank and AfDB, and various multilateral fund mechanisms. First, there must be a strategic shift from excessive land expansion toward productivity-focused intensive agriculture. The high positive score assigned to Fertilizer and Labor, contrasted with the negative impact of Land, reveals that clearing new land has reached its marginal limits. 

Second, the **Green Digital Transformation** of agriculture must be prioritized. The strong positive effect of Rural Electricity Access and the 2030 projection show that infrastructure is a prerequisite for AgriTech modernization (precision irrigation, cold chains). Third, a transition toward **Inclusive Agricultural Value Chain Development (iAVCD)** is essential to escape the "Raw Material Trap." Resource-rich economies like Nigeria should establish "Agricultural Modernization Funds" fueled by natural resource rents to mitigate Dutch Disease effects. Finally, it is recommended that regional policy analysis institutions integrate machine learning platforms into their policy cycles to enable real-time, data-driven foresight across the continent.


# 6. REFERENCES

Abdi, A. (2023). Economic complexity, renewable energy and CO₂ emissions in Sub-Saharan Africa: Evidence from heterogeneous panel estimators. *Environmental Science and Pollution Research, 30*, 1–16.

Arellano, M., & Bond, S. (1991). Some tests of specification for panel data: Monte Carlo evidence and an application to employment equations. *The Review of Economic Studies, 58*(2), 277–297.

Aminian, F., Suarez, E. D., Aminian, M., & Walz, D. T. (2006). Forecasting economic data with neural networks. *Computational Economics, 28*, 71–88.

Ariga, J., et al. (2019). Can improved agricultural technologies spur a green revolution in Africa? *Agricultural Economics*.

Beck, M. W. (2018). NeuralNetTools: Visualization and analysis tools for neural networks. *Journal of Statistical Software, 85*(11), 1–20. <https://doi.org/10.18637/jss.v085.i11>

Busse, M., Erdogan, C., & Mühlen, H. (2018). Structural transformation and its relevance for economic growth in Sub-Saharan Africa. *Review of Development Economics, 23*(1), 33-53.

Chen, J. (2005). *Neural network applications in agricultural economics* \[Doctoral dissertation, University of Kentucky\]. UKnowledge.

Chen, T., & Guestrin, C. (2016). Xgboost: A scalable tree boosting system. *Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining*, 785–794.

Corden, W. M., & Neary, J. P. (1982). Booming Sector and De-Industrialisation in a Small Open Economy. *The Economic Journal*, 92(368), 825-848. <https://doi.org/10.2307/2232670>

Dacrema, M. F., Cremonesi, P., & Jannach, D. (2019). Are we really making much progress? A worrying analysis of recent neural recommendation approaches. *Proceedings of the 13th ACM Conference on Recommender Systems*, 101–109.

Dumitrescu, E.-I., & Hurlin, C. (2012). Testing for Granger non-causality in heterogeneous panels. *Economic Modelling, 29*(4), 1450–1460.

De Vries, G., Timmer, M., & De Vries, K. (2015). Structural Transformation in Africa: Static Gains, Dynamic Losses. *The Journal of Development Studies, 51*(6), 674-688.

FAO. (2015). Regional overview of food insecurity: African food security prospects brighter than ever. *Food and Agriculture Organization of the United Nations*.

FAO. (2024). *The state of food and agriculture 2024: Financing agrifood systems transformation*. Food and Agriculture Organization of the United Nations.

Fritsch, S., Guenther, F., & Wright, M. N. (2019). *neuralnet: Training of neural networks*. R package version 1.44.2. <https://CRAN.R-project.org/package=neuralnet>

Gollin, D. (2014). Smallholder agriculture in Africa: An overview and implications for policy. *International Institute for Environment and Development (IIED)*.
 
Goujon, M., & Mien, E. (2025). The Extractive Resource Curse Revisited: Absolute Versus Relative De-Industrialization in Africa. *Review of Development Economics, 0*, 1-18.

Greenwell, B. M. (2021). fastshap: Fast approximate Shapley values. R package version 0.1.0. https://CRAN.R-project.org/package=fastshap

Grabowski, R., & Self, S. (2023). Structural change in Africa: role of agriculture. *Applied Economics Letters, 30*(11), 1467-1470.
 
Kamgnia, D. B. (2009). Growth intensity of employment in Africa: A panel data approach. *Journal of African Economies*.

He, Z., Qian, Y., & Jia, S. (2023). Interpretable machine learning models for analyzing patterns of remote correlations. *Artificial Intelligence Applications, 12*(3), 45-60.

Jafarbiglu, M. H., & Pourreza, N. A. (2022). A comprehensive review of machine learning techniques in soil salinity assessment. *Geocarto International, 37*, 1985–2005.

Jahn, M. (2020). Artificial neural network regression models in a panel setting: Predicting economic growth. *Economic Modelling, 91*, 148–154.

Kaba, K., Lin, J. Y., & Renard, M. F. (2022). Structural change and trade openness in sub-Saharan African countries. *The World Economy, 45*(7), 2101–2134.

Kaufmann, D., Kraay, A., & Mastruzzi, M. (2010). The worldwide governance indicators: Methodology and analytical issues. World Bank Policy Research Working Paper No. 5430.

Lewis, W. A. (1954). Economic development with unlimited supplies of labour. *The Manchester School, 22*(2), 139–191.

Kuan, C.-M., & White, H. (1994). Artificial neural networks: an econometric perspective. *Econometric Reviews, 13*(1), 1–91.

Li, Y., & Ma, W. (2010). Applications of artificial neural networks in financial economics: A survey. *2010 International Symposium on Computational Intelligence and Design*, 2510-2513.

Lundberg, S. M., & Lee, S.-I. (2017). A unified approach to interpreting model predictions. *Advances in Neural Information Processing Systems, 30*.

Magazzino, C., Mele, M., & Mutascu, M. (2025). Unemployment forecasting with Artificial Neural Networks. *Journal of Policy Modeling, 47*, 471–491.

McMillan, M., & Headey, D. (2014). Introduction – Understanding structural transformation in Africa. *World Development, 63*, 1–10.

Nguimkeu, P., & Zeufack, A. G. (2019). Manufacturing in structural change in Africa. World Bank Policy Research Working Paper No. 8992.

Mpakairi, K. S., Dube, T., Sibanda, M., Mutanga, O., Nhamo, L., & Mpandeli, S. (2025). Fine-scale mapping of irrigation suitability in South Africa using ensemble modelling. *Scientific Reports, 15*, 36524.

Olden, J. D., & Jackson, D. A. (2002). Illuminating the "black box": A randomization approach for understanding variable contributions in artificial neural networks. *Ecological Modelling, 154*(1-2), 135–150. <https://doi.org/10.1016/S0304-3800(02)00064-9>

Omotoso, A. B., Letsoalo, S., Olagunju, K. O., Tshwene, C. S., & Omotayo, A. O. (2023). Climate change and variability in sub-Saharan Africa: A systematic review of trends and impacts on agriculture. *Journal of Cleaner Production, 414*, 137487.
 
Sheahan, M., & Barrett, C. B. (2017). Ten striking facts about agricultural input use in Sub-Saharan Africa. *Food Policy*.

Pesaran, M. H. (2004). General diagnostic tests for cross section dependence in panels. *Cambridge Working Papers in Economics 0435*, Faculty of Economics, University of Cambridge.

Pesaran, M. H. (2006). Estimation and inference in large heterogeneous panels with a multifactor error structure. *Econometrica, 74*(4), 967–1012.

Pesaran, M. H. (2007). A simple panel unit root test in the presence of cross-section dependence. *Journal of Applied Econometrics, 22*(2), 265–312.

Pesaran, M. H., & Smith, R. (1995). Estimating long-run relationships from dynamic heterogeneous panels. *Journal of Econometrics, 68*(1), 79–113.

Pesaran, M. H., Shin, Y., & Smith, R. P. (1999). Pooled mean group estimation of dynamic heterogeneous panels. *Journal of the American Statistical Association, 94*(446), 621–634.

Pingali, P. L. (2012). Green revolution: Impacts, limits, and the path ahead. *Proceedings of the National Academy of Sciences, 109*(31), 12302–12308. [https://doi.org/10.1073/pnas.0912953109](https://www.google.com/search?q=https://doi.org/10.1073/pnas.0912953109)

Rodrik, D. (2016). Premature deindustrialization. *Journal of Economic Growth, 21*(1), 1–33.

Rostow, W. W. (1960). *The stages of economic growth: A non-communist manifesto*. Cambridge University Press.

Roodman, D. (2009). A note on the theme of too many instruments. *Oxford Bulletin of Economics and Statistics, 71*(1), 135–158.

R Core Team. (2023). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing. <https://www.R-project.org/>

Rozemberczki, B., Watson, L., Bayer, P., Yang, H.-T., Kiss, O., Nilsson, S., & Sarkar, R. (2022). The shapley value in machine learning. *arXiv preprint*, arXiv:2202.05594.

Seifipour, R., & Mehrabian, A. (2024). Application of artificial neural networks in economic and financial sciences. *Advances in Adaptive Filtering Theory and Applications*, IntechOpen.
 
Van den Broeck, G., & Kilic, T. (2019). Dynamics of off-farm employment in Sub-Saharan Africa: A gender perspective. *World Development*.

Telli, Ş., & Coşkun, M. (2016). Forecasting the BIST 100 Index using artificial neural networks with consideration of the economic calendar. *International Review of Economics and Management, 4*(3), 26-46.

Thawornwong, S., & Enke, D. (2004). The adaptive selection of financial and economic variables for use with artificial neural networks. *Neurocomputing, 56*, 205–232.

Veselý, A. (2011). Economic classification and regression problems and neural networks. *Agricultural Economics – Czech, 57*(3), 150–157.

Suleman, S., & Al-Hamwi, M. R. (2024). Machine learning-based modeling of Syrian agricultural GDP trends: A comparative analysis. *DYSONA – Applied Science*, 5(2), 89-104. https://doi.org/10.37118/ijdr.28018.03.2024

Xi, Y., Boateng, F. O., & Agyei, S. K. (2025). Agricultural productivity and renewable energy: Evidence from Ghana using double machine learning. *Sustainable Development, 33*(1), 234–251.

World Bank. (2023). *World development indicators*. The World Bank Group. <https://databank.worldbank.org/source/world-development-indicators>

Yurdakul, E. M. (2024). Forecastability of economic growth with the method of artificial neural networks: The case of Türkiye. *Journal of Current Researches on Social Sciences, 14*(4), 221-242.

