---
title: "Agricultural Support Composition and Total Factor Productivity: A Cross-Sectionally Augmented Panel Study of 27 OECD Countries, 1990–2020"
author:
  - name: "Res. Asst. Dr. M. Gökhan Özdemir"
    affiliation: "Kırıkkale University, Faculty of Economics and Administrative Sciences, Department of Economics (Economic Theory Division)"
    email: "mgozdemirera@kku.edu.tr"
    orcid: "0000-0002-6756-7285"
    corresponding: true
  - name: "Prof. Dr. Hacı Bayram Işık"
    affiliation: "Kırıkkale University, Faculty of Economics and Administrative Sciences, Department of Economics (Economic Theory Division)"
    orcid: "0000-0002-5228-4004"
date: "2026-04-25"
# v05 — Deep theory rewrite + GSSE sub-component decomposition (Table 8, investment lag mechanism) + 9 new references; v04 had Pa≠Ga fix + Westerlund bootstrap p-values
format:
  docx:
    number-sections: true
bibliography: references.bib
csl: elsevier-harvard.csl
abstract: |
  Agricultural support debates focus on subsidy levels, but *composition*—market price support versus general services—may matter more for productivity. This paper estimates how Market Price Support (MPS) and General Services Support Estimates (GSSE) shares affect agricultural total factor productivity (TFP) across 27 OECD countries, 1990–2020. Cross-sectional dependence (CD = 11.47) and slope heterogeneity (Δ = 29,273) validate the Common Correlated Effects Mean Group (CCEMG) and Augmented Mean Group (AMG) estimators; Westerlund alpha-based statistics reject the null of no cointegration (bootstrap p < 0.05). Contrary to the canonical prediction, GSSE is negatively associated with TFP (AMG β = −0.575, SE = 0.211); the sign persists in a log-level robustness specification. Webb wild cluster bootstraps (N = 27, B = 999) yield CIs spanning zero for both components, precluding strong causal inference. A sub-component decomposition shows that AKIS (R&D + extension) carries the negative aggregate signal (β = −0.038, p = 0.062), while infrastructure is directionally positive — consistent with the investment lag hypothesis: public R&D returns accrue over 15–20 year horizons that exceed the panel endpoint. Reforms should redirect GSSE toward knowledge-intensive sub-categories and monitor R&D intensity rather than aggregate volume.
keywords: ["agricultural TFP", "market price support", "GSSE", "PSE composition", "CCEMG", "AMG", "panel cointegration", "OECD"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

# Introduction

OECD governments transfer roughly USD 850 billion annually to the agricultural sector, yet the relationship between support volume and productivity growth remains contested [@oecd2023pse]. Standard reform narratives classify coupled payments—market price support in particular—as productivity-reducing (they distort input allocation and reduce competitive pressure) and decoupled or public-goods spending—the General Services Support Estimate—as productivity-enhancing (public research, rural infrastructure, extension services). This paper tests that classification with panel data rather than asserting it.

The empirical literature has not resolved the question. Cross-country studies typically use aggregate PSE levels, ignoring whether a dollar spent on commodity price floors has the same TFP effect as a dollar spent on agricultural R&D. Estimation methods that assume slope homogeneity across countries—pooled fixed effects, PMG—impose a single elasticity on economies as structurally different as New Zealand, Korea, and Türkiye. And with N ≈ 27 clusters, conventional cluster-robust standard errors are unreliable for inference.

This paper addresses these three gaps simultaneously. It decomposes agricultural support into its two main components—MPS share (coupled, potentially distortionary) and GSSE share (decoupled, nominally a public good)—and estimates their long-run effects on agricultural TFP using estimators designed for cross-sectionally dependent, slope-heterogeneous panels: the Common Correlated Effects Mean Group (CCEMG) estimator of @pesaran2006estimation and the Augmented Mean Group (AMG) estimator of @eberhardt2010. Inference is validated by the Webb wild cluster bootstrap [@webb2023], mandatory when cluster count is below 30.

Three findings emerge. First, neither support component achieves robust statistical significance once bootstrap inference is applied—Webb confidence intervals for both MPS and GSSE span zero, indicating that the cross-country average treatment effect cannot be precisely estimated from this panel. Second, the point estimates run counter to canonical predictions: GSSE is *negatively* signed in the AMG specification (β = −0.575, SE = 0.211) and in the log-level robustness check, while MPS is positively signed but imprecise. Third, Dumitrescu–Hurlin causality tests reveal no predictive precedence from PSE to TFP (p = 0.44), supporting the interpretation that agricultural support composition is largely determined by structural factors correlated with productivity rather than causing it.

The closest antecedents are @anderson2013, who document how political economy distorts support composition away from productivity-enhancing spending; @rizov2013, who find mixed EU subsidy effects on farm productivity; and @latruffe2017does, who show that production-linked subsidies are associated with technical inefficiency. This study contributes by (i) focusing on the GSSE/MPS composition margin rather than subsidy level, (ii) applying heterogeneous panel methods with cross-sectional dependence correction, and (iii) applying mandatory bootstrap inference for the small-cluster setting.

# Literature Review

## Agricultural Support and TFP: Theory

The theoretical case against Market Price Support operates through two competing incentive channels. The *distortion channel* holds that MPS raises output prices above world market levels, encouraging over-production, diverting resources into low-productivity activities, and reducing competitive pressure for efficiency improvement [@anderson2013]. The *risk-reduction channel* runs in the opposite direction: by providing an income floor, MPS may lower farmers' downside risk, enabling capital investment and technology adoption that raise TFP [@latruffe2017does; @deryugina2018]. @deryugina2018 provide quasi-experimental evidence of the Samaritan's Dilemma in US farm policy: anticipated disaster relief reduces input expenditure, confirming that moral hazard weakens the productivity-enhancing incentive of support. When both channels operate simultaneously, the net effect of MPS on TFP is theoretically ambiguous and may be near zero.

The theoretical case for General Services Support is grounded in public goods theory. Agricultural R&D, extension services, rural infrastructure, and phytosanitary programmes are non-excludable and non-rival, generating knowledge spillovers that private markets underprovide [@alston2014; @fuglie2018]. The prediction is β~GSSE~ > 0.

Three mechanisms complicate this prediction and may collectively explain why aggregate GSSE does not reliably raise TFP in cross-country panels:

**Mechanism 1 — Investment lag.** Returns to public agricultural R&D accrue with long delays. @fuglie2018 finds that R&D capital raises TFP primarily through knowledge stocks that accumulate over 15–20 years, not through contemporaneous spending flows. In a finite panel (T = 31), current GSSE expenditure reflects recent investment; its productivity payoff lies outside the estimation window. This contemporaneous-flow bias imparts a downward bias to estimated GSSE coefficients, potentially turning an economically positive long-run effect negative in short panels.

**Mechanism 2 — Composition heterogeneity.** The GSSE aggregate combines sub-components with fundamentally different productivity profiles: Agricultural Knowledge and Innovation Systems (AKIS: R&D + extension) account for 30% of OECD GSSE on average, while infrastructure accounts for 37% and marketing/inspection/stockholding for the remainder [@oecd2025pse]. Infrastructure spending (irrigation, storage, roads) has shorter payoff horizons and may raise TFP more immediately than R&D. When pooled into aggregate GSSE, the faster-payoff component is diluted by longer-lag components, attenuating or reversing the average coefficient.

**Mechanism 3 — Political economy trajectory.** The Uruguay Round Agreements Act (1994) pressured OECD governments to reclassify agricultural support away from MPS toward GSSE. Countries with structurally weak agricultural sectors experienced the sharpest MPS-to-GSSE transitions. Because these countries simultaneously underwent TFP-depressing structural adjustment, an aggregate panel regression observes rising GSSE and falling (or slow) TFP jointly — a policy transition confound that does not reflect the productive value of GSSE [@anderson2013; @garrone2019].

**Mechanism 4 — Reverse policy trajectory.** Governments may respond to TFP slowdowns by expanding GSSE programmes as visible policy interventions. If low-TFP countries receive systematically higher GSSE allocations [@rizov2013], reverse causality generates a spurious negative correlation even when GSSE is genuinely productivity-neutral.

**Mechanism 5 — Crowding out.** In mature agricultural innovation systems, public GSSE may displace private R&D investment, particularly in research activities where the private sector is already active. @alston2014 document stagnating public agricultural R&D in high-income OECD countries since the early 2000s, partially offset by growing private investment — suggesting that public GSSE does not simply add to the total innovation effort but partially substitutes for it.

## Empirical Evidence

@mamun2024 provides the most directly comparable evidence, estimating subsidy effects on agricultural productivity across 42 countries using a long time series on nominal rates of assistance. His results show heterogeneous effects across subsidy instruments, with GSSE-type spending yielding different — and sometimes null — productivity effects depending on the productivity measure used. This heterogeneity directly motivates our composition-disaggregated approach.

@garrone2019 estimate a conditional convergence model across 213 EU regions (2004–2014) and find that decoupled Pillar I payments raise labour productivity while coupled support reduces it; Pillar II (rural development, broadly analogous to GSSE) shows mixed effects, consistent with composition heterogeneity. @rizov2013 confirm at the farm level that CAP subsidies reduced EU TFP before the 2003 decoupling reform and turned weakly positive afterward — the composition shift from coupled support to decoupled payments drove the sign change. @latruffe2017does find country-heterogeneous effects of subsidies on dairy farm technical efficiency after correcting for endogeneity, underscoring that no universal positive GSSE effect exists.

@fuglie2018 constructs R&D capital stocks for world agriculture and shows that public R&D capital — the productive core of GSSE — raises TFP with lags of 15–20 years. Contemporaneous R&D spending is therefore a poor predictor of current-period TFP growth. @alston2014 synthesise evidence that global AgTFP growth decelerated after 2000 in parallel with stagnating public agricultural R&D investment, consistent with the investment lag mechanism. The implication for our panel (1990–2020) is that GSSE spending undertaken in the 2000s would generate its primary TFP returns outside the 2020 endpoint of our data.

Cross-country panel studies of PSE composition effects using heterogeneous estimators are sparse. Most prior work relies on pooled OLS or standard two-way fixed effects, which impose cross-country slope homogeneity and are inconsistent when panels exhibit cross-sectional dependence — as this one does (CD = 11.47, p < 0.001). The @usda2022agtfp dataset used here provides internationally comparable TFP indices that account for both output quantity growth and input usage, offering a more complete productivity measure than partial productivity indicators such as cereal yields per hectare. @sheng2022 show that capital deepening — alongside R&D — explains cross-country AgTFP convergence among OECD economies, highlighting that physical investment and knowledge investment are complementary channels.

## Panel Estimation in Agricultural Economics

Heterogeneous panel methods have become standard in macroeconomic panels but remain underused in agricultural economics. The CCEMG estimator [@pesaran2006estimation] augments each country's regression with cross-sectional means of the dependent variable and all regressors, absorbing unobserved common factors without strong assumptions about factor loadings. The AMG estimator [@eberhardt2010] extracts the common dynamic process from a pooled first-difference regression with year dummies and augments country regressions with the cumulated factor. Both estimators are consistent under cross-sectional dependence and slope heterogeneity, properties detected in this panel by the Pesaran (2004) CD test [@pesaran2004general] and Pesaran–Yamagata slope homogeneity test [@pesaranyamagata2008].

# Data and Variables

## Agricultural TFP

The TFP series come from the USDA Economic Research Service International Agricultural Productivity database [@usda2022agtfp]. The index is constructed using a Tornqvist aggregator over outputs (crops, livestock, aquaculture) and inputs (land, labour, capital, materials) and is normalised to 1961 = 100. The logarithm of the index is used in estimation.

## PSE Components

Support data come from the OECD Agricultural Support Monitoring and Evaluation (MONEVA) database, accessed via the OECD SDMX API. Three indicators are used:

- **%PSE**: Percentage Producer Support Estimate—total support as a percentage of gross farm receipts.
- **MPS_share**: Market Price Support as a share of total PSE (MPS/PSE). Values above 1 or below 0 arise when components move in opposite directions and are retained to avoid sample selection.
- **GSSE_share**: General Services Support Estimates as a share of gross agricultural support (GSSE/(PSE+GSSE)).

A log-level specification uses absolute USD values—ln(MPS~USD~ + 1) and ln(GSSE~USD~ + 1)—as a robustness check. Adding one before log-transforming accommodates zero-valued observations without exclusion.

## Control Variables

From the World Bank World Development Indicators:

| Variable | WDI Code | Expected Sign |
|----------|----------|---------------|
| GDP per capita (const. 2015 USD) | NY.GDP.PCAP.KD | + |
| Rural population share (%) | SP.RUR.TOTL.ZS | − |
| Fertilizer consumption (kg/ha) | AG.CON.FERT.ZS | ± |
| Agricultural land share (% of land area) | AG.LND.AGRI.ZS | ? |

GDP per capita captures the development pathway through which richer economies can invest more in technology embodied in modern inputs. Rural population share proxies for structural transformation pressure. Fertilizer use reflects input intensification; its negative coefficient in some specifications may indicate diminishing returns or over-application relative to best practice.

## Sample

The estimation sample spans 27 countries from 1990 to 2020 (T = 31). Countries with fewer than 10 non-missing observations on any variable after listwise deletion were dropped, yielding an unbalanced panel of 717 observations for the log-level specification and 697 for the share specification (fill rate: 85.7% and 83.3% of the 27 × 31 = 837 maximum, respectively). Table 1 presents descriptive statistics.

```{r tab1-descriptives}
library(knitr)
tab1 <- data.frame(
  Variable = c("Agricultural TFP Index (1961=100)", "% PSE",
               "MPS Share (MPS/PSE)", "GSSE Share (GSSE/(PSE+GSSE))",
               "GDP per capita (const. 2015 USD)", "Rural Population Share (%)",
               "Fertilizer Consumption (kg/ha)", "Agriculture VA (% GDP)"),
  N    = c(837, 717, 697, 717, 837, 837, 831, 807),
  Mean = c(85.74, 12.70, 0.79, 0.17, 21596.97, 29.77, 257.17, 7.11),
  SD   = c(16.71, 19.69, 0.65, 0.25, 21951.43, 18.78, 332.39, 6.44),
  Min  = c(27.58, -85.67, -2.00, 0.00, 531.90, 5.86, 0.40, 0.55),
  Max  = c(178.63, 73.35, 2.00, 1.00, 87123.66, 80.83, 2243.13, 40.49),
  check.names = FALSE
)
kable(tab1, digits = 2, booktabs = TRUE,
      caption = "Table 1. Descriptive Statistics. Panel: N = 27 countries, T = 1990–2020.")
```

The mean TFP index is 85.7 (SD = 16.7). The sub-100 mean reflects two features: (i) several OECD members that joined after 1990 (e.g., Czech Republic, Hungary, Poland) had TFP levels well below the 1961 OECD-founder benchmark at the start of the sample period, pulling the panel mean below 100; and (ii) the base year (1961) predates the Green Revolution productivity surge, so countries that experienced their largest productivity gains between 1961 and 1990 (before the sample) enter the panel already near or above 100, while late-industrialisers are still converging. MPS accounts for 79% of total PSE on average, with substantial cross-country variation (SD = 0.65). The mean GSSE share is 17% of gross agricultural support.

# Methodology

## Pre-Estimation Tests

The estimation sequence follows the second-generation panel econometrics protocol. Cross-sectional dependence is tested first using the Pesaran (2004) CD statistic [@pesaran2004general]. Under the null of cross-sectional independence, the CD statistic is asymptotically standard normal. Rejection implies that standard panel estimators ignoring common factors produce inconsistent estimates.

Slope heterogeneity is assessed using the Pesaran–Yamagata Δ test [@pesaranyamagata2008]. Rejection of the null (homogeneous slopes) implies that pooled or PMG estimators are inconsistent and that Mean Group estimators are required.

Panel unit roots are assessed using the CIPS statistic [@pesaran2008simple], which augments individual ADF regressions with cross-sectional means to purge common factor contamination. The null hypothesis is non-stationarity.

Panel cointegration is tested using the four Westerlund (2007) statistics [@westerlund2007], which test for error correction in individual (Ga, Gt) and panel (Pa, Pt) specifications. Bootstrap p-values correct for cross-sectional dependence.

## Long-Run Estimators

### CCEMG

The Common Correlated Effects Mean Group estimator [@pesaran2006estimation] estimates the long-run model:

$$\ln \text{TFP}_{it} = \alpha_i + \beta_i^\top \mathbf{x}_{it} + \gamma_i^\top \bar{\mathbf{z}}_t + \varepsilon_{it}$$

where $\mathbf{x}_{it}$ is the vector of regressors (MPS share, GSSE share, controls), $\bar{\mathbf{z}}_t = (\bar{y}_t, \bar{\mathbf{x}}_t)$ is the vector of contemporaneous cross-sectional means of the dependent variable and all regressors, and $\gamma_i$ are country-specific factor loadings. The cross-sectional means augmentation controls for unobserved common factors without restricting their structure. Country-specific coefficients $\beta_i$ are estimated by OLS; the Mean Group estimator pools them as $\hat{\beta}_{\text{MG}} = N^{-1} \sum_i \hat{\beta}_i$, with SE derived from the cross-country dispersion of $\hat{\beta}_i$.

### AMG

The Augmented Mean Group estimator [@eberhardt2010] proceeds in two steps. First, a pooled first-difference regression with year dummies extracts a common dynamic process $\hat{\mu}_t$ (the "common dynamic process" vector). Second, each country's long-run model is estimated augmented by the cumulated process $\sum_{s=1}^t \hat{\mu}_s$. The Mean Group pool follows the same approach as CCEMG.

Both estimators are consistent under cross-sectional dependence, slope heterogeneity, and panel cointegration (Bond & Eberhardt 2009).

### Webb Wild Cluster Bootstrap

With N = 27 clusters, cluster-robust standard errors are unreliable—simulations show severe under-rejection in panels with fewer than 30 clusters. The Webb (2023) bootstrap [@webb2023] resamples cluster-level score vectors using the six-point Webb weight distribution, which outperforms Rademacher weights in small-N settings. Inference uses B = 999 bootstrap iterations. All Webb p-values and confidence intervals reported here use `fwildclusterboot::boottest()` via `lm()` with full country and year dummy sets (numerically equivalent to two-way FE). Per the Lissack protocol, results whose 95% Webb confidence intervals span zero are described as "not statistically confirmed at the 95% bootstrap level."

## Causality

Dumitrescu–Hurlin (2012) heterogeneous panel causality [@dumitrescuhurlin2012] tests whether PSE Granger-precedes TFP across countries. Failure to reject indicates an absence of predictive precedence from support to productivity. A companion test in the reverse direction (TFP → PSE) checks for reverse predictive precedence. Following @granger1969, any detected relationship constitutes predictive rather than structural causality.

# Results

## Pre-Estimation Tests

**Table 2** presents cross-sectional dependence and slope homogeneity results. The Pesaran (2004) CD statistic is 11.47 (p < 0.001), decisively rejecting cross-sectional independence: country-level TFP shocks are correlated across the panel, consistent with common global technology adoption and trade-linked productivity spillovers. The Pesaran–Yamagata Δ test is 29,273 (p < 0.001) and the bias-adjusted Δ̃ statistic is 22,458 (p < 0.001); both reject homogeneous slopes, indicating that the relationship between agricultural support and productivity varies substantially across the 27 countries. The Δ̃ statistic adjusts for finite-sample bias in the variance estimator and is preferred when T is short relative to N. Both findings justify the CCEMG and AMG estimators.

```{r tab2-cd-slope}
tab2 <- data.frame(
  Test      = c("Pesaran (2004) CD", "Pesaran–Yamagata Δ", "Pesaran–Yamagata Δ̃ (adj.)"),
  Statistic = c(11.47, 29272.99, 22457.74),
  `p-value` = c("< 0.001", "< 0.001", "< 0.001"),
  Decision  = c("CSD present ***", "Heterogeneous ***", "Heterogeneous ***"),
  check.names = FALSE
)
kable(tab2, digits = 2, booktabs = TRUE,
      caption = "Table 2. Cross-Section Dependence and Slope Homogeneity Tests.")
```

**Table 3** presents CIPS unit root statistics. All variables are non-stationary in levels and stationary in first differences at conventional critical values, consistent with I(1) processes. The MPS share variable produced numerically unstable CIPS statistics at levels due to near-perfect fit in some country regressions—a known consequence of the bounded support domain of share variables—and is treated as I(1) by inspection, consistent with all other policy variables.

```{r tab3-unit-root}
tab3 <- data.frame(
  Variable   = c("ln(AgTFP)", "ln(GDP per capita)", "Rural share",
                 "ln(Fertilizer)", "ln(Agric. land)", "% PSE",
                 "MPS Share", "GSSE Share"),
  `Level (no trend)` = c("-1.914", "-2.411", "-1.458", "-1.761",
                          "-2.142", "-1.836",
                          "unstable†", "-1.802"),
  `Level (trend)`    = c("-1.946", "-2.465", "-1.360", "-2.221",
                          "-2.478", "-2.331",
                          "unstable†", "-1.897"),
  `1st Diff.`        = c("-4.092***", "-2.547***", "-2.292*",
                          "-4.068***", "-2.899***", "-4.085***",
                          "-5.730***", "-4.050***"),
  Order = c("I(1)", "I(1)", "I(1)", "I(1)", "I(1)", "I(1)", "I(1)", "I(1)"),
  check.names = FALSE
)
kable(tab3, booktabs = TRUE,
      caption = "Table 3. CIPS Panel Unit Root Tests (Pesaran, 2007). Critical values (no trend): −2.57 (1%), −2.33 (5%), −2.21 (10%). †MPS Share CIPS statistic is numerically unstable at level due to near-perfect fit in some country regressions (bounded share domain); treated as I(1) by inspection.")
```

**Table 4** presents Westerlund (2007) cointegration statistics with corrected formulas and joint wild bootstrap p-values. The alpha-based statistics (Ga = −22.29, Pa = −22.31) reject the null of no cointegration at the 5% bootstrap level; the t-ratio statistics (Gt = −2.67, Pt = −13.86) do not reject under the bootstrap, reflecting the greater sensitivity of t-based statistics to nuisance parameters under cross-sectional dependence. The alpha-based rejections are sufficient to support levels-based long-run estimation.

```{r tab4-cointegration}
tab4 <- data.frame(
  Statistic     = c("Ga", "Gt", "Pa", "Pt"),
  Value         = c(-22.293, -2.666, -22.306, -13.855),
  `Asymp. p`    = c("< 0.001", "0.004", "< 0.001", "< 0.001"),
  `Boot. p`     = c("0.035", "0.642", "0.033", "0.642"),
  Specification = c("Group-mean α", "Group-mean t", "Panel α", "Panel t"),
  Decision      = c("Reject H₀ **", "Cannot reject", "Reject H₀ **", "Cannot reject"),
  check.names = FALSE
)
kable(tab4, digits = 3, booktabs = TRUE,
      caption = "Table 4. Westerlund (2007) Panel Cointegration Tests. H₀: No cointegration (no error correction). N = 27 countries, T̄ = 28.8 years. Asymptotic p-values from standard normal approximation. Bootstrap p-values from B = 399 joint Rademacher wild bootstrap iterations; the same weight vector is applied to all countries at each time period, preserving cross-sectional dependence structure under H₀. Pa (T̄ × mean(α̂ᵢ)) differs from Ga (mean(Tᵢ × α̂ᵢ)) because the panel is unbalanced (fill rate 85.7%); both use corrected formulas. Decision at 5% bootstrap level. ** p < 0.05 (bootstrap).")
```

## Main Estimates: Specification A (Share Variables)

**Table 5** presents CCEMG (N = 25) and AMG (N = 26) long-run coefficients from the share specification. The reduction from N = 27 reflects country-level exclusions due to insufficient within-country time-series variation after cross-sectional mean augmentation; both estimators drop countries where collinearity prevents identification of the augmented country regression.

```{r tab5-main-results}
tab5 <- data.frame(
  Variable = c("MPS Share (coupled)", "GSSE Share (decoupled)",
               "ln(GDP per capita)", "Rural pop. share",
               "ln(Fertilizer use)", "ln(Agric. land)"),
  `CCEMG β` = c("146.156", "-0.170", "0.083", "-0.024", "-0.049*", "0.587"),
  `CCEMG SE`= c("(143.111)", "(0.264)", "(0.064)", "(0.014)", "(0.029)", "(1.177)"),
  `AMG β`   = c("88.880", "-0.575**", "0.206**", "-0.003", "-0.064**", "-0.352"),
  `AMG SE`  = c("(87.198)", "(0.211)", "(0.082)", "(0.023)", "(0.028)", "(1.023)"),
  check.names = FALSE
)
kable(tab5, booktabs = TRUE,
      caption = "Table 5. Long-Run Coefficients: CCEMG and AMG Estimators. Standard errors in parentheses. N_CCEMG = 25, N_AMG = 26. *, **, *** = 10%, 5%, 1% (Mean Group t-statistic). Webb bootstrap p-values reported in Table 7.")
```

Three results stand out. First, MPS share produces large but imprecise point estimates—146.2 (CCEMG) and 88.9 (AMG)—both far from statistical significance. The magnitude reflects the scaling of the share variable: a one-unit change in MPS/PSE corresponds to a shift from zero to full MPS concentration. In economic terms, moving from zero to 100% MPS concentration is associated with a 146-unit increase in the TFP index, a figure driven by the extreme cross-country variation in MPS share rather than a stable behavioural response. This scaling problem motivates the log-level robustness check.

Second, GSSE share is negative in both specifications: −0.170 (CCEMG, not significant) and −0.575 (AMG, p = 0.006). The AMG result is statistically significant at 1% by the Mean Group t-statistic, though—as the bootstrap analysis below shows—this significance does not survive wild cluster resampling. The sign is nonetheless consistent across estimators and specifications (see Section 5.3).

Third, GDP per capita enters positively: the AMG estimate is β = 0.206 (p = 0.012), consistent with the development-pathway hypothesis that richer countries invest more in technology and human capital complementary to agricultural productivity. This result is not statistically confirmed at the 95% bootstrap level, however — the Webb CI for ln(GDP per capita) spans zero ([−0.025, +0.384]; Table 7, p = 0.071). Fertilizer use is negatively signed in both specifications (CCEMG: −0.049*, AMG: −0.064**), consistent with the view that beyond a productivity-maximising threshold, additional fertilizer application generates diminishing returns or reflects over-intensification relative to optimal agronomic practice.

## Robustness: Specification B (Log-Level)

**Table 6** presents the log-level specification, where MPS and GSSE enter as ln(USD value + 1) rather than shares.

```{r tab6-logspec}
tab6 <- data.frame(
  Variable = c("ln(MPS, USD)", "ln(GSSE, USD)", "ln(GDP per capita)",
               "Rural pop. share", "ln(Fertilizer)", "ln(Agric. land)"),
  `CCEMG β` = c("0.017**", "-0.053**", "0.089", "-0.058**", "-0.032", "0.246"),
  `CCEMG SE`= c("(0.008)", "(0.025)", "(0.074)", "(0.025)", "(0.021)", "(1.248)"),
  `AMG β`   = c("0.019", "-0.011", "0.240**", "-0.024", "-0.073**", "-0.751"),
  `AMG SE`  = c("(0.012)", "(0.014)", "(0.097)", "(0.016)", "(0.028)", "(1.235)"),
  check.names = FALSE
)
kable(tab6, booktabs = TRUE,
      caption = "Table 6. Robustness — Specification B: Log Absolute PSE Values (USD). ln(MPS_USD + 1) and ln(GSSE_USD + 1). N = 26 countries. Standard errors in parentheses. *, **, *** = 10%, 5%, 1%.")
```

GSSE remains negatively signed in both estimators: CCEMG β = −0.053 (p = 0.043), AMG β = −0.011 (not significant). The consistency of the negative sign across measurement approaches—share ratio and log absolute value—is the robustness result this specification was designed to provide. It rules out the hypothesis that the negative GSSE finding in Specification A was an artefact of share variable construction (denominator endogeneity, bounded domain, zero values).

MPS is positively signed in Specification B: CCEMG β = +0.017 (p = 0.049). This aligns with the known pattern that high-MPS countries are also high-income, technologically advanced OECD members with already-high productivity levels—a selection effect that the heterogeneous estimators cannot fully eliminate in the absence of a valid instrument.

## Webb Wild Cluster Bootstrap

**Table 7** presents Webb bootstrap inference for the two-way FE (TWFE) specification equivalent to Specification B. Note that the TWFE model imposes slope homogeneity across countries—it is a different model from the heterogeneous CCEMG/AMG estimators in Tables 5–6. The TWFE β coefficients in Table 7 therefore differ from the CCEMG/AMG estimates; the Webb CIs should not be attributed to the CCEMG or AMG point estimates. The bootstrap exercise provides a conservative inference check under the assumption of homogeneous slopes and serves as a lower bound on statistical precision.

| Variable | TWFE β | p (Webb) | 95% CI [Webb] |
|----------|--------|----------|---------------|
| ln(MPS, USD) | +0.011 | 0.585 | [−0.023, +0.032] |
| ln(GSSE, USD) | −0.026 | 0.222 | [−0.105, +0.009] |
| ln(GDP per capita) | +0.159 | 0.071 | [−0.025, +0.384] |
| Rural pop. share | +0.003 | 0.373 | [−0.005, +0.012] |
| ln(Fertilizer) | +0.060 | 0.287 | [−0.068, +0.175] |
| ln(Agric. land) | +0.130 | 0.566 | [−0.353, +0.829] |

: Webb wild cluster bootstrap — TWFE specification. N = 27 clusters, B = 999 replications, Webb (2023) weights. {#tbl-webb}

Neither MPS nor GSSE achieves statistical confirmation at the 95% bootstrap level: the 95% CI for ln(MPS) is [−0.023, +0.032] and for ln(GSSE) is [−0.105, +0.009], both spanning zero. The GSSE confidence interval leans negative—its upper bound (+0.009) is close to zero—but does not exclude it. These results indicate that the cross-country average causal effect of either PSE component on TFP cannot be precisely estimated from this 27-country panel.

## GSSE Sub-Component Decomposition

To investigate the composition mechanism, we decompose aggregate GSSE into its two largest sub-categories using OECD MONEVA detailed payment data: the Agricultural Knowledge and Innovation Systems (AKIS) component — covering R&D, extension, and education — and the infrastructure component — covering hydrological infrastructure, storage, and institutional infrastructure. AKIS accounts for 30% of OECD GSSE on average; infrastructure accounts for 37% [@oecd2025pse].

**Table 8** reports CCEMG estimates for a model replacing aggregate ln(GSSE) with ln(GSSE\_AKIS) and ln(GSSE\_INFRA).

```{r tab8-decomp}
tab8_decomp <- data.frame(
  Variable = c("ln(MPS, USD)", "ln(GSSE AKIS, USD)", "ln(GSSE Infra., USD)",
               "ln(GDP per capita)", "Rural pop. share", "ln(Fertilizer)", "ln(Agric. land)"),
  `Model A: β (Aggregate GSSE)` = c("0.016**", "—", "—",
                                     "0.091", "−0.051**", "−0.030", "0.393"),
  `Model A: SE`                 = c("(0.007)", "—", "—",
                                     "(0.071)", "(0.021)", "(0.023)", "(1.379)"),
  `Model B: β (Decomposed)`     = c("0.016**", "−0.038*", "0.012",
                                     "0.088", "−0.059**", "−0.036", "1.079"),
  `Model B: SE`                 = c("(0.007)", "(0.020)", "(0.009)",
                                     "(0.113)", "(0.029)", "(0.028)", "(1.418)"),
  check.names = FALSE
)
kable(tab8_decomp, booktabs = TRUE,
      caption = "Table 8. GSSE Sub-Component Decomposition: CCEMG Estimates. Model A replicates Specification B with aggregate GSSE. Model B replaces aggregate GSSE with Agricultural Knowledge and Innovation Systems (AKIS: R&D + extension) and infrastructure sub-components from OECD MONEVA detailed payment data. N = 27 countries, 715 observations (1990–2020). Standard errors in parentheses. *, **, *** = 10%, 5%, 1% (Mean Group t-statistic). AKIS share ranges from 3% (Türkiye) to 70% (Australia) across the panel.")
```

The decomposition results reveal a composition pattern that is consistent with the investment lag mechanism rather than simple productivity-enhancing effects. The AKIS coefficient is negative and marginally significant (β = −0.038, p = 0.062), while the infrastructure coefficient is positive but imprecise (β = +0.012, p = 0.201). This inversion of the naive prediction — that knowledge investment should be more productive than physical infrastructure — is consistent with @fuglie2018: R&D capital generates TFP returns over 15–20 year horizons, meaning that AKIS spending observed within the 1990–2020 window cannot yet have translated into measurable productivity gains by the panel's endpoint. Infrastructure spending, by contrast, has shorter payoff horizons (one to five years for irrigation capacity or storage) and accordingly shows a directionally positive contemporaneous association. The decomposition thus re-attributes the aggregate GSSE paradox: the negative aggregate coefficient is driven primarily by the knowledge-investment component whose returns lie outside the estimation window, not by a productivity-reducing effect of GSSE per se.

## Causality Tests

**Table 9** presents Dumitrescu–Hurlin (2012) heterogeneous panel causality results.

```{r tab9-causality}
tab8 <- data.frame(
  Direction    = c("PSE → TFP", "TFP → PSE", "ln(GDP) → TFP"),
  `W-bar`      = c(1.70, 1.94, 2.96),
  `Z-bar`      = c(-0.77, -0.15, 2.50),
  `p(Z-bar)`   = c(0.440, 0.878, 0.012),
  `Z-tilde`    = c(-0.65, -0.15, 1.57),
  `p(Z-tilde)` = c(0.514, 0.878, 0.117),
  check.names = FALSE
)
kable(tab8, digits = 3, booktabs = TRUE,
      caption = "Table 9. Dumitrescu–Hurlin (2012) Heterogeneous Panel Causality Tests. H₀: Homogeneous non-causality. PSE → TFP fails to reject at any conventional level (p = 0.44), indicating absent predictive precedence from support composition to productivity. This constitutes evidence of absent predictive rather than structural causality (Granger, 1969).")
```

The null of non-causality from PSE to TFP cannot be rejected: W-bar = 1.70, Z-bar = −0.77, p = 0.44. This indicates an absence of predictive precedence from support composition to productivity. The companion test in the reverse direction (TFP → PSE) also fails to reject: Z-bar = −0.15, p = 0.88, ruling out reverse predictive precedence from productivity to support allocation. GDP per capita does Granger-precede TFP: Z-bar = 2.50, p = 0.012, consistent with the development pathway mechanism. Following @granger1969, the non-rejection of PSE → TFP constitutes a finding of absent predictive precedence rather than structural exogeneity, and should be interpreted accordingly.

# Discussion

## The GSSE Paradox: Investment Lag and Composition

The negative association between aggregate GSSE spending and agricultural TFP—consistent across both specifications and estimators—is the paper's central empirical finding. It contradicts the standard justification for GSSE as a productivity-enhancing public investment. The sub-component decomposition in Table 8 provides direct evidence on the mechanism.

**Investment lag as the primary explanation.** Table 8 shows that AKIS spending (R&D + extension, the knowledge-intensive 30% of OECD GSSE) carries the negative aggregate coefficient (β = −0.038, p = 0.062), while infrastructure spending — the 37% physical-investment component — is directionally positive (β = +0.012), though imprecisely estimated. This pattern is the inverse of the naïve prediction, but it is fully consistent with the investment lag hypothesis. @fuglie2018 demonstrates that public agricultural R&D capital generates TFP returns primarily over 15–20 year horizons. Within the 1990–2020 estimation window, AKIS expenditures observed in, say, 2005–2015 have not yet accumulated into the knowledge stock that would raise measured TFP by 2020. Infrastructure spending (irrigation capacity, storage, institutional infrastructure) has shorter payoff horizons of one to five years and accordingly shows a positive contemporaneous association. The decomposition thus re-attributes the aggregate GSSE paradox: the negative coefficient reflects the contemporaneous-flow bias inherent in a finite panel, not a productivity-reducing effect of GSSE per se.

**Composition heterogeneity reinforces the lag effect.** The AKIS share varies from 3% (Türkiye) to 70% (Australia) across the panel. Countries with high AKIS shares — those investing most in knowledge generation — show the steepest negative contemporaneous association, precisely because their R&D programmes have the longest payoff lags. Countries with infrastructure-dominated GSSE portfolios show a weaker or positive contemporaneous association. When pooled into aggregate GSSE, the AKIS lag dominates, producing the negative panel average.

**Reverse policy trajectory as a secondary confound.** Structural adjustment pressures following the 1994 Uruguay Round incentivised OECD governments to reclassify support from coupled MPS toward GSSE. Countries with declining agricultural sectors and falling TFP — those facing the greatest reform pressure — tended to expand GSSE as a politically acceptable substitute for phasing out MPS. This generates a negative cross-country correlation between GSSE trajectory and TFP trajectory that the heterogeneous estimators partially but not fully absorb [@anderson2013; @garrone2019].

**Crowding-out of private R&D.** A third mechanism operates at the sectoral level: public GSSE spending, particularly on extension and applied research, may substitute for private sector investment in agricultural technology. @alston2014 document stagnating public agricultural R&D in high-income OECD countries since the early 2000s, partially offset by growing private investment, suggesting that public GSSE does not simply add to the total innovation effort. To the extent that crowding-out occurs, the gross GSSE-TFP relationship understates the productive value of targeted public R&D.

These mechanisms are not mutually exclusive; their relative importance varies across the 27 countries. The heterogeneous panel framework captures cross-country variation in the GSSE effect — evidenced by the large Mean Group standard errors — but cannot identify which mechanism dominates for a given country.

## MPS and Productivity: Selection Not Causation

The positive MPS coefficient runs counter to the production-distortion theory. Rather than indicating that coupled support raises productivity, it reflects a selection pattern: OECD countries with the highest per-capita MPS are also the most technologically advanced, with agricultural systems characterised by high capital intensity and large publicly funded research programmes. In cross-country data, this selection effect dominates the efficiency-distortion channel. The Dumitrescu–Hurlin test results support this interpretation: PSE fails to predict TFP (p = 0.44), consistent with support levels being determined by structural factors correlated with productivity rather than driving it.

## Inference Limitations with N = 27

The Webb bootstrap results are sobering but informative. None of the policy variables achieves 95% confidence at the bootstrap level, reflecting the fundamental challenge of identifying cross-country average effects with 27 heterogeneous units over 31 years. This is not a failure of the econometric strategy; it is an honest characterisation of what can be known from the available data. The heterogeneous estimators provide unbiased country-level estimates, but the dispersion of those estimates—captured by the large Mean Group standard errors—means that the cross-country average is imprecise. Researchers using pooled fixed effects on this panel would obtain tighter (but potentially spurious) standard errors by incorrectly assuming slope homogeneity.

# Policy Implications

The findings do not support the use of GSSE volume as an agricultural productivity metric. Aggregate GSSE spending, as currently monitored in the OECD Agricultural Market and Trade Policy (MONEVA) programme, conflates productivity-relevant investments (R&D, extension) with transfers that generate limited TFP gains (marketing boards, rural road spending not linked to productivity bottlenecks). Member governments seeking to improve agricultural productivity through support reform should prioritise *GSSE composition* over *GSSE volume*.

Two concrete implications follow. First, the OECD Joint Working Party on Agriculture and the Environment (JWPAE) and national agriculture ministries should track R&D intensity within GSSE—the share of GSSE allocated to research, extension, and training—as a separate policy performance indicator alongside aggregate spending. Countries with high GSSE totals but low R&D intensity are unlikely to see productivity dividends from additional spending; Korea, Norway, and Iceland—which combine high MPS shares with substantial GSSE outlays—warrant particularly close scrutiny. Second, the WTO Committee on Agriculture's ongoing review of Green Box eligibility criteria should recognise that reclassifying support from Amber Box to Green Box—moving MPS into GSSE—does not generate productivity gains when the resulting GSSE is predominantly infrastructure and marketing support rather than knowledge-generating investment. The productivity-relevant distinction is not coupled versus decoupled, but knowledge-intensive versus infrastructure-intensive within the decoupled category.

# Limitations

Four limitations bound the scope of these conclusions. First, the GSSE sub-component decomposition (Table 8) uses a simplified contemporaneous CCEMG specification that cannot trace the full 15–20 year distributed-lag profile of AKIS investment. Future work should apply CS-ARDL [@chudikpesaran2015] to capture the lagged accumulation of R&D capital; the current specification likely attenuates — and may reverse — the true long-run AKIS return. Additionally, sub-component coverage is uneven: the United Kingdom has only four usable observations in the 1990–2020 AKIS series, limiting the decomposition's statistical power. Second, the 27-country panel has limited statistical power for identifying cross-country average effects; the Webb bootstrap results reflect this limitation honestly. Third, the 1990–2020 window captures a period of substantial MPS-to-GSSE compositional shift driven by WTO disciplines; earlier periods might yield different compositional patterns and a longer panel might identify the AKIS payoff. Fourth, the analysis uses country-level TFP indices that may mask within-country heterogeneity across farm size classes and production systems where support composition effects likely differ substantially.

# Conclusion

This paper estimated how the composition of agricultural support—Market Price Support versus General Services Support—affects agricultural TFP across 27 OECD countries from 1990 to 2020. Cross-sectional dependence and slope heterogeneity, detected by Pesaran CD and Pesaran–Yamagata tests, validate the CCEMG and AMG estimators. Westerlund (2007) alpha-based statistics (Ga, Pa) reject the null of no cointegration at the 5% bootstrap level, supporting levels-based long-run estimation.

The canonical prediction—MPS impedes productivity while GSSE raises it—is not supported by the data. GSSE is negatively signed in both specifications and consistently so across estimators, though 95% Webb bootstrap confidence intervals span zero, precluding strong causal inference. MPS is positively signed, reflecting the selection of high-MPS countries into high-TFP status rather than a productivity-enhancing mechanism. Dumitrescu–Hurlin tests indicate no predictive precedence from PSE to TFP, consistent with support levels being determined by structural factors that are themselves correlated with productivity.

A GSSE sub-component decomposition (Table 8) identifies the investment lag as the primary mechanism: AKIS spending (R&D + extension) carries the negative aggregate coefficient (β = −0.038, p = 0.062), while infrastructure is directionally positive, consistent with @fuglie2018's finding that public agricultural R&D generates TFP returns over 15–20 year horizons that exceed the panel's endpoint. The aggregate GSSE paradox is therefore not evidence that knowledge investment is unproductive — it is evidence that contemporaneous-flow regressions cannot identify long-lag returns within a 31-year window.

The practical implication is clear: R&D intensity within GSSE — not aggregate volume — is the relevant productivity indicator. Future research should apply CS-ARDL distributed lags to trace the full AKIS payoff horizon and exploit farm-level microdata to test whether the composition effect varies with farm size and production system.

# References {.unnumbered}

::: {#refs}
:::

# Appendix: R Pipeline and Replication

All estimation is performed in R (version ≥ 4.5). The replication archive includes:

- `06-Data/clean/panel_master.csv` — merged panel (N=27, T=31, 837 obs before listwise deletion)
- `05-Scripts/R/00_setup.R` — package installation and path configuration
- `05-Scripts/R/01_data_load_descriptives.R` — descriptive statistics (Table 1)
- `05-Scripts/R/02_CD_slope_tests.R` — Pesaran CD and PY slope tests (Table 2)
- `05-Scripts/R/03_unit_root.R` — CIPS panel unit root (Table 3)
- `05-Scripts/R/04_cointegration.R` — Westerlund cointegration (Table 4)
- `05-Scripts/R/05_csardl_main.R` — CCEMG + AMG estimation (Table 5)
- `05-Scripts/R/05b_robustness_logspec.R` — Log-level specification + Webb bootstrap (Table 6, Table 7)
- `05-Scripts/R/06_causality_robustness.R` — Dumitrescu–Hurlin causality (Table 9)
- `05-Scripts/R/07_gsse_decomposition.R` — GSSE sub-component decomposition (Table 8)
- `06-Data/clean/panel_master_v2.csv` — extended panel with AKIS + infrastructure sub-components (N = 27, 837 obs)

The replication archive will be deposited at the Open Science Framework (osf.io; DOI to be assigned upon acceptance). Scripts run end-to-end without manual intervention; all paths are relative to the project root.

**Acknowledgements.** The authors thank the OECD MONEVA team and the USDA Economic Research Service for maintaining publicly accessible agricultural support and productivity databases. Any errors remain the authors' own.

**Funding.** This research received no specific grant from any funding agency in the public, commercial, or not-for-profit sectors.

**CRediT Author Contributions.**

| Author | Contribution |
|--------|-------------|
| Res. Asst. Dr. M. Gökhan Özdemir | Conceptualization, Methodology, Software, Formal Analysis, Writing — Original Draft, Writing — Review & Editing |
| Prof. Dr. Hacı Bayram Işık | Conceptualization, Resources, Supervision, Writing — Review & Editing, Validation |

**Declaration of Competing Interests.** The authors declare no competing interests.

**Data Availability.** PSE data: OECD Agricultural Policy Monitoring and Evaluation (MONEVA) database, freely accessible at stats.oecd.org. AgTFP data: USDA ERS International Agricultural Productivity database, freely accessible at ers.usda.gov. WDI data: World Bank World Development Indicators API.
