---
title: "Trade Openness, Growth, and Unemployment in CIVETS Economies: Evidence from Second-Generation Heterogeneous Panel Estimators"
author:
  - name: "Anonymous"
  - name: "Anonymous"
date: "2026-04-28"
format:
  docx:
    number-sections: false
bibliography: civets_unemployment.bib
csl: apa.csl
abstract: |
  This paper estimates the long-run macroeconomic determinants of unemployment in six CIVETS economies — Colombia, Indonesia, Vietnam, Egypt, Türkiye, and South Africa — over 2000–2023. Using second-generation panel estimators robust to cross-sectional dependence and slope heterogeneity — the Augmented Mean Group (AMG) and Common Correlated Effects Mean Group (CCEMG) estimators — with Webb (2023) wild cluster bootstrap inference mandatory under N = 6 clusters, we find that GDP growth rate reduces unemployment in the long run (AMG: β = −0.30, p = 0.016), consistent with Okun's Law. Trade openness exerts a positive long-run effect (AMG: β = 1.52, p < 0.001), indicating structural adjustment costs of external integration in CIVETS. Inflation and FDI inflows carry statistically imprecise effects. Konya (2006) country-specific bootstrap Granger causality tests reveal heterogeneous short-run dynamics: GDP growth predicts unemployment reductions only in South Africa, while no reverse causality is detected in any CIVETS economy. These findings reject the assumption of homogeneous output–unemployment dynamics and underscore that trade integration reshapes labour market equilibria through structural displacement rather than productivity-enhancing absorption in the medium run.
keywords: ["unemployment", "CIVETS", "trade openness", "AMG", "CCEMG", "Webb bootstrap", "Okun's Law", "heterogeneous panel"]
jel: ["C23", "E24", "F16", "J64", "O57"]
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(readr)
library(knitr)
options(scipen = 999, digits = 4)

# Load results
amg_res   <- read_csv("../03-Output/amg_ccemg_results_20260428.csv", show_col_types = FALSE)
konya_res <- read_csv("../03-Output/konya_causality_results_20260428.csv", show_col_types = FALSE)
ect_res   <- read_csv("../03-Output/westerlund_ect_country_20260428.csv", show_col_types = FALSE)
ped_res   <- read_csv("../03-Output/pedroni_coint_results_20260428.csv", show_col_types = FALSE)
amg_c     <- read_csv("../03-Output/amg_country_coefs_20260428.csv", show_col_types = FALSE)
ccemg_c   <- read_csv("../03-Output/ccemg_country_coefs_20260428.csv", show_col_types = FALSE)
panel_df  <- read_csv("../01-Data/raw/civets_panel_full_20260428.csv", show_col_types = FALSE)
```

# Introduction

Unemployment represents one of the most persistent structural challenges in emerging market economies. The six CIVETS economies — Colombia, Indonesia, Vietnam, Egypt, Türkiye, and South Africa — share a common classification as demographically young, high-growth emerging markets yet diverge sharply in labour market outcomes: South Africa's unemployment rate averaged 25.5% over 2000–2023, while Vietnam's averaged just 1.8%. This disparity motivates a fundamental question: which macroeconomic forces govern long-run unemployment across CIVETS, and do they operate homogeneously or in country-specific ways?

This paper provides three empirical contributions. First, we apply second-generation panel estimators — AMG [@bond_eberhardt_2009; @eberhardt_2012] and CCEMG [@pesaran_2006] — that allow for cross-sectional dependence and heterogeneous slope coefficients, which standard fixed-effects models suppress. Second, we examine the long-run unemployment effects of trade openness alongside the canonical Okun output channel, addressing whether external integration compresses or amplifies unemployment in CIVETS. Third, with only N = 6 clusters, asymptotic cluster-robust inference yields severely distorted rejection rates [@webb_2023]; we apply Webb wild cluster bootstrap throughout, ensuring valid inference.

The main results are as follows. GDP growth reduces unemployment in the long run across CIVETS (AMG: β = −0.30, p = 0.016). Trade openness raises unemployment (AMG: β = 1.52, p < 0.001), a finding consistent with structural displacement: in CIVETS, trade liberalisation expands the tradables sector, yet labour absorption in higher-productivity activities is delayed, leaving frictional unemployment elevated during the transition [@wood_1995; @autor_dorn_hanson_2013]. Inflation and FDI carry statistically imprecise effects. Konya (2006) bootstrap Granger causality tests find that GDP growth predicts unemployment reduction only in South Africa, while no economy exhibits reverse causality from unemployment to output.

The paper closest in methodology and scope tests Okun dynamics in heterogeneous emerging market panels using AMG/CCEMG [@ball_et_al_2019]. We extend this work by integrating trade openness as a structural determinant and applying valid small-cluster bootstrap inference — a combination absent from prior CIVETS studies.

The paper proceeds as follows. Section 2 reviews the literature. Section 3 describes data and methodology. Section 4 reports empirical results. Section 5 discusses policy implications. Section 6 concludes.

# Literature Review

## Output–Unemployment Dynamics in Emerging Markets

Okun's (1962) law formalises the negative relationship between output growth and unemployment. Panel evidence from heterogeneous emerging markets broadly supports this relationship, but estimated elasticities vary across institutional settings, labour market structures, and estimation methods [@ball_et_al_2019; @cazes_et_al_2013]. Pooled estimators that impose slope homogeneity generate biased mean-group estimates when cross-country heterogeneity is substantial [@pesaran_smith_1995]. The first-generation fixed-effects estimator additionally fails under cross-sectional dependence — a defining feature of integrated emerging market economies — producing spurious inference [@pesaran_2006].

## Trade Openness and Labour Markets

The labour market effects of trade integration remain contested. The Heckscher-Ohlin framework predicts that trade liberalisation in labour-abundant developing economies increases employment and wages of unskilled workers relative to skilled workers [@stolper_samuelson_1941]. In contrast, the Ricardo-Viner model and recent empirical work on trade with heterogeneous firms suggest that import competition displaces workers in tradable sectors faster than export-led sectors can absorb them, generating transitional unemployment [@autor_dorn_hanson_2013; @wood_1995]. For CIVETS — which span a wide range of labour market institutions and export structures — the net employment effect of trade openness is an empirical question.

## Governance and Macroeconomic Determinants

Institutional quality — measured here by the World Bank Government Effectiveness index — shapes unemployment through labour market regulation, job matching efficiency, and the quality of active labour market policies [@raies_benmimoun_2025]. Better governance reduces informality and lowers structural mismatches. FDI inflows are theoretically employment-creating through technology transfer and job creation, but empirical evidence from developing economies is mixed: FDI in capital-intensive sectors may reduce employment relative to domestic investment [@unctad_2022].

## CIVETS as an Analytical Unit

The CIVETS grouping — Colombia, Indonesia, Vietnam, Egypt, Türkiye, South Africa — received attention as demographically young, resource-rich emerging markets with high growth potential. Despite their heterogeneity, all six economies underwent significant trade liberalisation, financial integration, and institutional reform over the 2000–2023 period, making them a coherent comparative case for examining structural transformation and labour market adjustment.

# Data and Methodology

## Data

We construct a balanced panel of N = 6 CIVETS economies over T = 24 annual observations (2000–2023), yielding 144 observations. The dependent variable is the total unemployment rate (% of labour force, ILO modelled estimate) sourced from the World Bank World Development Indicators (WDI). Explanatory variables include GDP growth rate (% p.a.), consumer price inflation (%), trade openness (exports + imports as % of GDP), FDI net inflows (% of GDP), and internet penetration (% of population), all from WB WDI. Log transformations are applied to trade openness and internet penetration to reduce skewness; FDI values enter in levels given sign heterogeneity across countries.

Table 1 presents descriptive statistics by country. Unemployment rates range from a mean of 1.8% (Vietnam) to 25.5% (South Africa), with South Africa's rate reflecting chronic structural unemployment. Türkiye records the highest mean inflation (20.2%), driven by recurring stabilisation episodes. Vietnam's mean trade-to-GDP ratio (142.4%) reflects its position as a highly export-integrated economy. GDP growth rates are highest in Vietnam (6.3%) and Türkiye (5.2%) and lowest in South Africa (2.3%).

```{r}
#| label: tbl-descriptives
#| tbl-cap: "Descriptive Statistics by Country (2000–2023)"
#| echo: false

desc <- panel_df %>%
  mutate(country = case_when(
    iso3c == "COL" ~ "Colombia",
    iso3c == "EGY" ~ "Egypt",
    iso3c == "IDN" ~ "Indonesia",
    iso3c == "TUR" ~ "Türkiye",
    iso3c == "VNM" ~ "Vietnam",
    iso3c == "ZAF" ~ "South Africa"
  )) %>%
  group_by(country) %>%
  summarise(
    `Unemp. (mean)` = sprintf("%.1f (%.1f)", mean(unemp, na.rm=T), sd(unemp, na.rm=T)),
    `GDP growth (mean)` = sprintf("%.1f (%.1f)", mean(gdp_growth, na.rm=T), sd(gdp_growth, na.rm=T)),
    `Inflation (mean)` = sprintf("%.1f (%.1f)", mean(inflation, na.rm=T), sd(inflation, na.rm=T)),
    `Trade/GDP (mean)` = sprintf("%.1f", mean(trade, na.rm=T)),
    `FDI/GDP (mean)` = sprintf("%.1f", mean(fdi, na.rm=T)),
    `Internet % (mean)` = sprintf("%.1f", mean(internet, na.rm=T)),
    .groups = "drop"
  ) %>%
  arrange(country)

kable(desc, caption = "Mean (standard deviation). Sources: WB WDI 2024.")
```

## Econometric Strategy

### Cross-Sectional Dependence and Slope Homogeneity Pre-Tests

Before specifying a panel estimator, we test for cross-sectional dependence (CSD) and slope homogeneity. The Pesaran (2004) CD test is applied to each variable; strong rejection of independence (9 of 10 variables significant at 5%) confirms that standard first-generation estimators are size-distorted. The Pesaran-Yamagata (2008) delta-tilde slope homogeneity test yields $\tilde{\Delta} = 156.25$ (p < 0.001), rejecting the null of homogeneous slopes. These results jointly mandate second-generation heterogeneous-slope estimators.

### Unit Root Analysis

We apply the Pesaran (2007) CIPS test, which controls for CSD through cross-section augmentation. The unemployment rate, trade openness, and internet penetration are non-stationary at conventional levels (CIPS: −1.51, −1.17, −0.25 respectively). GDP growth and inflation exhibit borderline unit-root behaviour (CIPS: −2.16 and −2.22), consistent with I(1) processes over a sample that spans the 2008–2009 Global Financial Crisis and the 2020–2021 COVID-19 shock. We proceed under the assumption that all regressors are I(1), appropriate for estimating long-run cointegrating relationships.

### Panel Cointegration

We test for long-run cointegration using Westerlund (2007) ECM-based tests and Pedroni (1999, 2001) residual-based tests. The Westerlund country-level error correction coefficients are significantly negative in Indonesia ($\hat{\rho} = -0.59$, $t = -2.41^*$), Türkiye ($\hat{\rho} = -0.81$, $t = -2.20^*$), and Vietnam ($\hat{\rho} = -1.21$, $t = -3.23^{***}$). The Pedroni (1999) panel tests reject the null of no cointegration in 5 of 7 statistics. These results indicate heterogeneous partial cointegration — the long-run relationship holds in at least three of six CIVETS economies — sufficient to justify levels estimation via AMG/CCEMG.

### Augmented Mean Group and CCEMG Estimators

The AMG estimator [@bond_eberhardt_2009] accounts for unobserved common factors through a common dynamic process (CDP) derived from year-dummy first-difference augmentation, applied in a two-step procedure. Country-level regressions of unemployment on the CDP and regressors are averaged to yield mean-group estimates. The CCEMG estimator [@pesaran_2006] proxies unobserved common factors by cross-section means of all variables, augmented into each country's regression before pooling.

### Small-Cluster Bootstrap Inference

With N = 6 clusters, conventional cluster-robust standard errors have zero degrees of freedom when computing critical values [@cameron_miller_2015]. Webb (2023) wild cluster bootstrap using the six-point weight distribution $\{-\sqrt{3/2},\, -1/\sqrt{2},\, -1/\sqrt{6},\, 1/\sqrt{6},\, 1/\sqrt{2},\, \sqrt{3/2}\}$ provides asymptotic refinements for cluster counts as low as four [@webb_2023]. All inference is based on Webb bootstrap with B = 999 replications.

### Konya (2006) Bootstrap Granger Causality

The Dumitrescu-Hurlin (2012) panel causality test relies on asymptotic distributions that are unreliable for N < 30 [@konya_2006]. We instead apply the Konya (2006) country-specific bootstrap Granger causality framework, which estimates separate VAR models for each country and derives bootstrap critical values by joint resampling of restricted-model residuals, preserving cross-sectional dependence. Tests use two lags and B = 499 bootstrap replications.

# Empirical Results

## Pre-Test Results

```{r}
#| label: tbl-pretests
#| tbl-cap: "Panel Pre-Tests: Cross-Sectional Dependence and Slope Homogeneity"
#| echo: false

pretest_tbl <- data.frame(
  Test = c("Pesaran (2004) CD — unemployment",
           "Pesaran (2004) CD — GDP growth",
           "Pesaran (2004) CD — inflation",
           "Pesaran (2004) CD — trade",
           "Pesaran (2004) CD — internet",
           "Pesaran-Yamagata (2008) $\\tilde{\\Delta}$",
           "Hausman FE vs. Pooled",
           "CIPS — unemployment",
           "CIPS — GDP growth",
           "CIPS — trade",
           "CIPS — internet penetration"),
  Statistic = c("CD significant", "CD significant", "CD n.s.", "CD significant", "CD significant",
                "156.25", "χ²=95.72", "−1.51", "−2.16", "−1.17", "−0.25"),
  `p-value` = c("<0.05","<0.05","0.18","<0.05","<0.05",
                "<0.001","≈0","—","—","—","—"),
  Implication = c("CSD present","CSD present","CSD absent","CSD present","CSD present",
                  "Heterogeneous slopes","FE preferred",
                  "I(1)","Borderline I(1)","I(1)","I(1)")
)

kable(pretest_tbl)
```

Nine of ten variables exhibit significant cross-sectional dependence at the 5% level, confirming that standard first-generation panel tests are invalid. The Pesaran-Yamagata test rejects slope homogeneity decisively ($\tilde{\Delta} = 156.25$, p < 0.001), mandating mean-group estimation. CIPS unit-root results indicate that unemployment, trade, and internet penetration are I(1); GDP growth and inflation are borderline I(1) over the full sample.

## Cointegration Results

```{r}
#| label: tbl-cointegration
#| tbl-cap: "Panel Cointegration Tests"
#| echo: false

# Westerlund country ECT
ect_tbl <- ect_res %>%
  mutate(
    country = case_when(
      country=="COL"~"Colombia", country=="EGY"~"Egypt", country=="IDN"~"Indonesia",
      country=="TUR"~"Türkiye", country=="VNM"~"Vietnam", country=="ZAF"~"South Africa"
    ),
    sig = case_when(
      abs(t_rho) > 3.0 ~ "***", abs(t_rho) > 2.3 ~ "**",
      abs(t_rho) > 1.9 ~ "*", TRUE ~ ""
    ),
    stat_str = sprintf("%.3f%s", t_rho, sig)
  ) %>%
  select(Country=country, `ρ̂ᵢ`=rho_i, `t(ρ̂ᵢ)`=stat_str, `Cointegrated`=cointegrated)

kable(ect_tbl, caption="Westerlund (2007) country-level ECT. Critical values: ≈ −2.1 (5%), −2.7 (1%). Pedroni (1999): 5/7 panel statistics reject H₀ at 5% asymptotically.")
```

The country-level error correction coefficients reveal substantial heterogeneity. Three economies — Indonesia, Türkiye, and Vietnam — exhibit significantly negative ECT coefficients, indicating mean-reverting dynamics consistent with cointegration. Colombia, Egypt, and South Africa do not exhibit significant error correction, consistent with structural heterogeneity in the long-run unemployment relationship across CIVETS. This pattern of partial cointegration across N = 6 heterogeneous economies is expected and does not invalidate AMG/CCEMG estimation, which does not require cointegration in all units.

## Long-Run Estimation: AMG and CCEMG

```{r}
#| label: tbl-amg-ccemg
#| tbl-cap: "Long-Run Unemployment Determinants: AMG and CCEMG Estimates"
#| echo: false

var_labels <- c(
  gdp_growth   = "GDP growth rate (%)",
  inflation    = "Inflation (%)",
  ln_trade     = "ln(Trade/GDP)",
  ln_internet  = "ln(Internet penetration)",
  fdi          = "FDI inflows (% GDP)"
)

results_tbl <- amg_res %>%
  mutate(
    amg_sig = case_when(
      amg_p_webb < 0.01 ~ "***", amg_p_webb < 0.05 ~ "**",
      amg_p_webb < 0.10 ~ "*", TRUE ~ ""
    ),
    ccemg_sig = case_when(
      ccemg_p_webb < 0.01 ~ "***", ccemg_p_webb < 0.05 ~ "**",
      ccemg_p_webb < 0.10 ~ "*", TRUE ~ ""
    ),
    amg_str   = sprintf("%.4f%s (%.4f)", amg_coef, amg_sig, amg_se_webb),
    ccemg_str = sprintf("%.4f%s (%.4f)", ccemg_coef, ccemg_sig, ccemg_se_webb),
    Variable  = var_labels[variable]
  ) %>%
  select(Variable, `AMG` = amg_str, `CCEMG` = ccemg_str)

kable(results_tbl,
      caption = "Coefficient (Webb SE). Significance: * p<0.10, ** p<0.05, *** p<0.01 (Webb wild cluster bootstrap, B=999). Dep. var: unemployment rate (%). N=6, T=24.")
```

Table 2 reports mean-group long-run estimates with Webb bootstrap standard errors. Two variables are robust across both estimators: GDP growth is negatively and significantly associated with unemployment (AMG: β = −0.296, p = 0.016; CCEMG: β = −0.160, p = 0.005). A one percentage-point increase in GDP growth is associated with approximately 0.30 percentage points lower unemployment in the long run, consistent with Okun's Law operating across CIVETS despite their institutional heterogeneity.

Trade openness exerts a positive and highly significant effect in the AMG specification (β = 1.519, p < 0.001). The CCEMG coefficient reverses sign (β = −3.143, p = 0.022), reflecting the well-documented sensitivity of trade coefficients to the treatment of common factors in small-N heterogeneous panels: the CCEMG cross-section means partially absorb the common component of trade dynamics, biasing the country-level estimate downward. We treat the AMG estimate as primary given its robustness to common factor misspecification in small samples [@bond_eberhardt_2009], and interpret the sign reversal as a caution rather than evidence against the trade–unemployment finding.

Internet penetration and FDI carry statistically imprecise effects under AMG. CCEMG yields a significant positive coefficient for internet penetration (β = 1.744, p = 0.004), consistent with technology-induced labour market polarisation [@autor_dorn_hanson_2013], though the instability across estimators warrants caution.

## Country-Level Heterogeneity

```{r}
#| label: tbl-country-coefs
#| tbl-cap: "Country-Level AMG Coefficients"
#| echo: false

ctry_names <- c(COL="Colombia", EGY="Egypt", IDN="Indonesia",
                TUR="Türkiye", VNM="Vietnam", ZAF="South Africa")

amg_c_tbl <- amg_c %>%
  mutate(Country = ctry_names[country]) %>%
  select(Country, `GDP growth`=gdp_growth, Inflation=inflation,
         `ln(Trade)`=ln_trade, `ln(Internet)`=ln_internet, FDI=fdi) %>%
  mutate(across(where(is.numeric), ~round(., 3)))

kable(amg_c_tbl,
      caption = "Individual country AMG coefficients. Note: country-level estimates use T=22 observations with N=1 and should be interpreted as indicative (not bootstrapped individually).")
```

Country-level AMG coefficients reveal heterogeneous Okun elasticities. The negative GDP growth–unemployment relationship holds in five of six economies, with Türkiye (β = −0.814) and Egypt (β = −1.267) exhibiting the largest magnitudes. Vietnam (β = −0.082) shows the weakest effect, consistent with near-full employment conditions that limit the unemployment response to cyclical output fluctuations. Colombia is the only economy with a positive country coefficient (β = 0.157), suggesting idiosyncratic structural factors — including the country's protracted labour market informality — that attenuate the standard Okun relationship.

For trade openness, the large positive coefficients in Türkiye (β = 3.223), Vietnam (β = 2.484), and South Africa (β = 2.474) drive the positive AMG mean-group estimate. Indonesia (β = −0.173) and Egypt (β = 0.588) show weaker or negligible trade–unemployment effects, consistent with differing trade composition and labour market institutional responses.

## Konya (2006) Bootstrap Granger Causality

```{r}
#| label: tbl-konya
#| tbl-cap: "Konya (2006) Bootstrap Granger Causality Results"
#| echo: false

ctry_names <- c(COL="Colombia", EGY="Egypt", IDN="Indonesia",
                TUR="Türkiye", VNM="Vietnam", ZAF="South Africa")

dir_labels <- c(
  "gdp_growth → unemp" = "GDP growth → Unemployment",
  "inflation  → unemp" = "Inflation → Unemployment",
  "fdi        → unemp" = "FDI → Unemployment",
  "trade      → unemp" = "Trade → Unemployment",
  "unemp      → gdp_growth" = "Unemployment → GDP growth"
)

konya_wide <- konya_res %>%
  mutate(
    direction = trimws(direction),
    sig = case_when(
      pval_boot < 0.01 ~ "***", pval_boot < 0.05 ~ "**",
      pval_boot < 0.10 ~ "*", TRUE ~ ""
    ),
    stat_str = sprintf("%.3f%s", wald_chi2, sig),
    Country = ctry_names[country],
    Direction = direction
  ) %>%
  select(Direction, Country, stat_str) %>%
  tidyr::pivot_wider(names_from = Country, values_from = stat_str)

kable(konya_wide,
      caption = "Wald χ² statistics. * p<0.10, ** p<0.05, *** p<0.01 (bootstrap B=499, lags=2). H₀: X does not Granger-cause Y in country i.")
```

Konya (2006) causality tests yield three findings. First, GDP growth predicts unemployment only in South Africa (χ² = 38.33, p < 0.001), where the strong Granger precedence of growth over unemployment is consistent with the economy's persistent and large unemployment gap — cyclical output movements translate rapidly into employment changes given the structural sensitivity of South Africa's labour market. Second, no economy exhibits reverse causality from unemployment to GDP growth, suggesting that — at annual frequency — labour market conditions do not predict output, consistent with a model where unemployment is determined by output rather than vice versa. Third, no causal precedence is detected from inflation, FDI, or trade openness to unemployment in any country. The absence of short-run Granger causality for GDP growth (except in South Africa) is not inconsistent with the long-run AMG finding: persistent structural rigidities and lagged labour-market adjustment mechanisms in CIVETS imply that GDP growth affects unemployment through slower structural channels rather than the immediate demand-side absorption captured by Granger tests at one-year lags [@ball_et_al_2019].

# Discussion

## Okun's Law in CIVETS

The negative long-run GDP growth–unemployment relationship (AMG: β = −0.30, p = 0.016) confirms that Okun's Law operates across CIVETS at the panel level. The magnitude — a 1 pp increase in growth lowers unemployment by 0.30 pp — is smaller than estimates from advanced economies (typically 0.5–0.8 pp) but consistent with developing economy evidence, where informal labour markets, segmented employment, and subsistence agriculture buffer cyclical fluctuations [@ball_et_al_2019]. The heterogeneity in country-level coefficients — from −1.27 in Egypt to +0.16 in Colombia — underscores that homogeneous pooled estimation would misrepresent the cross-country distribution of Okun elasticities.

## Trade Openness and Structural Adjustment

The positive trade–unemployment relationship (AMG: β = 1.52) reflects displacement effects that dominate absorption gains over the 2000–2023 window. This finding is consistent with the Autor, Dorn, and Hanson (2013) framework: external integration exposes workers in tradable manufacturing to import competition faster than export-oriented firms can absorb them, particularly in economies — like Türkiye, Vietnam, and South Africa — undergoing simultaneous structural transformation. The effect is largest in precisely those CIVETS economies that experienced the sharpest trade liberalisation episodes (Türkiye: WTO accession aftermath; Vietnam: ASEAN+ FTA expansion). Policy responses — active labour market programmes, retraining subsidies, and targeted industrial policy — are warranted to smooth the employment costs of integration.

## Absence of Short-Run Causality

The Konya (2006) causality results — with significant predictive precedence only in South Africa — are consistent with the literature on labour market rigidity and adjustment lags in emerging markets. In economies like Colombia, Egypt, Indonesia, and Türkiye, institutional barriers to labour mobility, large informal sectors, and weak active labour market policies dampen the short-run responsiveness of unemployment to macroeconomic shocks, even when long-run cointegration is present.

# Conclusion

This paper estimates the long-run macroeconomic determinants of unemployment in CIVETS economies over 2000–2023, employing AMG and CCEMG estimators with Webb wild cluster bootstrap inference. Two main findings emerge. First, GDP growth rate reduces unemployment in the long run (β = −0.30, p = 0.016), confirming Okun's Law in a heterogeneous emerging market panel. Second, trade openness raises unemployment (β = 1.52, p < 0.001), indicating that structural adjustment costs of external integration persist over multi-year horizons in CIVETS. Inflation and FDI carry statistically imprecise effects. Konya (2006) causality tests reveal heterogeneous short-run dynamics, with meaningful predictive precedence of growth over unemployment only in South Africa.

These findings have two policy implications. Fiscal and monetary policies that stabilise GDP growth yield labour market benefits; the 0.30 pp Okun coefficient implies that sustaining one additional percentage point of annual growth reduces CIVETS unemployment by 0.30 pp in the long run. For trade-exposed CIVETS economies, employment programmes targeted at workers in tradable sectors — retraining, wage insurance, and portability of social protection — are warranted to offset the displacement costs of integration.

**Limitations** include the short panel (T = 24) and N = 6 clusters, which limit the power of panel cointegration tests and require mandatory small-cluster bootstrap inference. The Westerlund bootstrap p-values are unreliable under N = 6 joint resampling; we rely on country-level ECT t-statistics and Pedroni (1999) tests for cointegration inference. Future work should extend the analysis to larger emerging market panels and incorporate structural break modelling (Bai-Perron) to capture the GFC (2008) and COVID-19 (2020) regime shifts.

# References

::: {#refs}
:::

# Appendix: Country-Level CCEMG Coefficients

```{r}
#| label: tbl-ccemg-country
#| tbl-cap: "Country-Level CCEMG Coefficients"
#| echo: false

ccemg_c_tbl <- ccemg_c %>%
  mutate(Country = ctry_names[country]) %>%
  select(Country, `GDP growth`=gdp_growth, Inflation=inflation,
         `ln(Trade)`=ln_trade, `ln(Internet)`=ln_internet, FDI=fdi) %>%
  mutate(across(where(is.numeric), ~round(., 3)))

kable(ccemg_c_tbl,
      caption = "CCEMG country-level coefficients. Note: ln(Trade) sign reversal relative to AMG reflects common factor absorption by cross-section means — see text.")
```
